### 2019-05-31 BiometricsPaper/bp_intro.tex
added latest version of the paper

%!TEX root = bioPrediction_main.tex

\todo{NAK: I have heard from either Thomas or CGM a figure that captures how many problems are due to human error (something like 80\%). If we can find that cite, it should figure prominently in the motivation.  UPDATE: Mau added
this book to the bib file, so we can cite it like this\cite{hse99}. The text
says the following: "Accidents can occur through people's involvement with their work. As technical systems have become more reliable, the focus has turned to human causes of accidents. It is estimated that up to 80\% of accidents may be attributed, at least in part, to the actions or omissions of people."}

\section{Introduction}
%Control rooms, while not much different from standard office settings on the surface, differ in the immediacy of their impact. In emergency dispatch offices, factory control rooms, and mining operations centers decision immediately impact safety and profits. A sleepy factory operator that misses early warnings could lose millions of dollars due to a process shutdown; a stressed mining operator that mistakenly dispatches two underground trucks to the same location could cause a deadly crash.  In these demanding settings  the environment must be finely tuned to support the job at hand. For instance, on overnight shifts, if controlling the lighting helps workers shift their circadian sleep cycle, reducing their sleepiness, it should be done, as the consequences of bad decisions are high.

%To ensure the best decisions are made we focus on reducing stress, increasing focus and awakeness. Unfortunately, beyond a few lightly-studied phenomenon such as temperature's affect on awakeness, there are few absolutes that can be applied to workplaces wholesale. Thus, when optimizing a workspace tracking the affect of a given change, if any, is key.

%The challenge with tracking the affect of a given change in a workspace on its workers is that it is most often done via self-reporting, or the process of filling out daily, hourly, or worse surveys on one's current awakeness, stress, and focus. Worse, this process must be repeated every time a change is implemented. Performing this process rapidly becomes impractical since the changes in the environment may be very frequent  (e.g., modifying the room temperature several times a day). Previous studies show that interruptions that undesired constant interruptions that happen at inopportune moments have negative effects, causing more stress and frustration~\cite{adamczyk2004if,czerwinski2000instant,mark2008cost}.

% The importance of the problem
Human decisions in the workplace directly impact worker productivity, product/process quality, and workplace safety. Control rooms are noteworthy for the immediacy of impact. Control room operators make hundreds of decisions during a workday, and the consequences of poor decisions can be severe. A sleepy factory operator could cause millions of dollars in losses by failing to recognize the early warning signs of a process shutdown, an unfocused water-processing plant operator could cause a water line failure by missing an alarm warning of a pressure overload, or a stressed mining operator could cause a deadly crash by mistakenly dispatching two trucks to the same underground location.

% The overall problem
The goal of control room design is to create and maintain an environment that maximizes operator alertness (i.e., that decreases stress and that increases focus and awakeness). Modern control rooms rely on adjustments to ergonomics, light, and temperature to support improved operator alertness. For example, adjusting the lighting in a control room can help operators on overnight shifts to alter their circadian sleep cycles, improving their awakeness while at work. A key challenge in control room design is understanding the effects of a given change to the control room environment on the operators. In particular, reliance on operator self-reporting to capture current levels of stress, focus, and awakeness is cumbersome, distracting, and labor intensive. Indeed, self-reporting can affect the phenomena being measured; e.g., interrupting work to respond to a survey can reduce focus. Moreover, as self-reporting is required not only to create the initial environment, but also to maintain it over time, operator self-reporting is impractical and can be harmful to operator alertness. Previous studies establish that regular interruptions, particularly those that happen at inopportune moments, cause stress and frustration~\cite{adamczyk2004if,czerwinski2000instant,mark2008cost}.

% What others have done
Previous research addresses how various signals such as blood pressure, heartbeat, and temperature can be linked to psychological states and processes~\cite{Kramer90,Rowe98,Eekelen04,valentini10}. However, little research addresses how biometric signals can be used to predict proxies for alertness such as levels of stress, focus, and awakeness. Instead, previous research has focused on using biometric signals to assess task difficulty in software development~\cite{fritz2014using}, code quality online~\cite{Muller16}, or  interruptibility~\cite{zuger18}.

% What we present in this paper
\todo{NAK: Would like the opinion of Thomas here --- should we give more of our vision here or stick to the (current) approach of keeping it specific to this paper. That is, do we want to talk about using biometrics to monitor and improve the workplace, or just stick to talking about predicting based on biometrics?}
The overall goal of our work is to create a machine learning 
model based in anecdotal biometric
data which can infer the level of stress, focus, and 
awakeness of a person by analyzing their biometric data.
This approach will remove the unnecessary constant interruptions by inferring
the stress, focus, and awakeness of an individual from their biometric features.
This will allows to assess if a change performed in the environment 
of an individual is beneficial
or prejudicial to their productivity and alertness and, therefore, be able to 
act upon this information to reverse or increase the change.

% The contributions of this paper
The contributions of this paper are:
\begin{itemize}
\item The results of an exploratory study on the viability of using biometric sensors to predict the stress, focus, and awakeness of an individual
\item A machine learning approach to predict proxies for alertness using biometric signals
\item Guidance regarding the optimal time window and learning curve to generate an effective predictive model
\item A comparison between the proposed approach using biometric signals and an approach using computer interaction data
\end{itemize}

### 2019-05-31 BiometricsPaper/bp_intro_v3.tex
added latest version of the paper

%!TEX root = bioPrediction_main.tex
\section{Introduction}
\textit{`The most valuable asset of a 21st-century institution (whether business or non-business) will be its knowledge workers and their productivity'}~\cite{drucker1999knowledge}. Yet, in today's workplaces, knowledge workers and their time are not necessarily treated as the most valuable asset, and they constantly face challenges, such as a high work fragmentation, continuous disruptions and distractions, highly complex and demanding tasks, and long working hours~\cite{gonzalez2004constant,mark2008cost,czerwinski04diary}. These challenges make it difficult for a knowledge worker to stay focused and can impact both the quality of work performed and productivity. For instance, the continuous disruptions that knowledge workers face in office work environments can lead to a higher error rate and a lower performance~\cite{bailey2001effects,mark2008cost}. In some cases,
the consequences are even more severe; in control room
settings, the consequences can be accidents that cost millions of dollars in losses. 
%when a sleepy operator might, for example, miss early warning signs and cause...
%\todo{Nick: can you find a good example here and citation?}. 
%Similarly, stress at the workplace is a growing concern at the workplace and one of the most common work-related health problem that can lead to fatigue, burnout and various other illnesses, resulting in work absence and a high productivity loss~\cite{hockey1997stress,setz2010stress,wrs2010}. 
Less immediate, but just as insidious, stress at the workplace is a growing concern and one of the most common work-related health problems. It leads to fatigue, burnout and various other illnesses, ultimately resulting in work absences and marked productivity losses~\cite{hockey1997stress,setz2010stress,wrs2010}.

Being able to continuously measure human aspects --- such as focus, awakeness, and stress --- in the workplace bears huge potential for better supporting knowledge workers, combatting workplace challenges, and improving knowledge worker productivity. Such measures could be used to provide better management of disruptions~\cite{iqbal2005effectiveness,zuger2017reducing}, to automatically adjust lighting to reduce sleepiness among control room operators, or to continuously monitor stress levels to trigger pro-active interventions and reduce health problems.

Fortunately, with recent advances in sensing technologies there is increasing hope that we can accurately and automatically measure such human aspects. A number of studies have already discussed and investigated the use of biometrics to measure aspects such as a person's cognitive and emotional states~\cite{beatty82,Kramer90,Rowe98,richter1998psychophysiological,Wilson2002}. Studies have also looked more specifically at using biometrics to measure stress~\cite{dishman2000stress,Setz2010}, or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004emotion}. Yet, very few studies measure productivity-related factors in the workplace over an extended period of time, and little is known about measuring the more abstract concept of focus. Given the importance of focus, described as a combination of engagement and challenge in a work task by Mark et al.~\cite{mark2014bored}, to an office workers' productivity, this factor also deserves further study.

The objective of our research is to develop continuous (and automatic) measures of focus, awakeness, and stress in the workplace. These measures are targeted with an eye towards improving knowledge workers' productivity and well-being in the future. For instance, by automatically protecting the worker from audible interruptions during a high focus period by showering him/her with white noise or by recommending stress-reducing interventions during an extended period of high stress.

Our work builds upon previous work and extends it by performing an eight-week study in the workplace with 14 knowledge workers, collecting biometric and computer interaction data as well as experience samples on focus, awakeness, and stress. Our participants, who perform various job functions for a research and development group within a single large corporation, wore a single biometric armband sensor\footnote{Biovotion Everion sensor~\cite{everion}} with low invasiveness that captured heart- and skin-related measures. This modality was chosen to ease longitudinal deployment in the field. Using machine learning, we created classifiers and analyzed their ability to predict the three productivity-related human aspects. In addition, we compared biometric features with computer interaction features in their predictive power, we analyzed how the combination
of these two training sets compares to the performance of each
set individually.

The results of our analysis show that biometrics of a single minimally invasive sensor can be used to predict productivity-related indicators accurately, with the abstract concept of focus (predictably) being the hardest to detect. The results also show that knowledge workers' self-reported levels of stress, focus and awakeness and their physiological manifestation and prediction can vary a lot between individuals. Our results further show that overall combining biometric features with computer interaction information 
results in an improvement
over each of the training sets individually.
The main contributions of our work are:
\begin{itemize}
	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' awakeness, stress, and focus in the workplace based on an eight-week field study with 14 office workers.
	\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
	\item A discussion on the impact of applying this research to possible target workspaces, besides a reflection on aspects that can be further improved in future studies.
    \end{itemize}

%Ideally businesses believe that \textit{`The most valuable asset of a 21st-century institution (whether business or non-business) [is] its knowledge workers and their productivity'}~\cite{drucker1999knowledge}. In practice these valuable assets are as mercilessly optimized as assembly lines during the industrial age. Knowledge workers' desks occupy increasingly small footprints, optimizing a company's cost per square unit. Their shifts, especially in factories, command centers, or web service organizations, span twenty four hours a day, optimizing uptime. 
%
%
%
%In these settings they face a variety of challenges. Their squeezed workspaces lead to continuous disruptions and distractions, on top of the high work fragmentation and highly complex tasks typical of modern office work~\cite{gonzalez2004constant,zuger2015interruptibility,zuger18,mark2014bored}. %This leads to an up to 80\% of accidents being attributed, at least in part, to actions or omissions of workers\cite{hse99}. 
%Their shift work--and the required overnight shifts that accompany it--naturally causes sleepines. These challenges combine to increase stress and decrease focus, which in turn impacts worker productivity. For instance, the continuous disruptions that knowledge workers face can lead to higher error rates, lower overall performance~\cite{bailey2001effects,mark2008cost}, and, in a control room setting, could easily cause an accident that costs millions of dollars.
%%when a sleepy operator might, for example, miss early warning signs and cause...
%%\todo{Nick: can you find a good example here and citation?}. 
%Less immediate but just as insidious, stress at the workplace is a growing concern and one of the most common work-related health problems. It leads to fatigue, burnout and various other illnesses, ultimately resulting in work absences and marked productivity losses~\cite{hockey1997stress,setz2010stress,wrs2010}. 
%
%The first step in combating these common workplace challenges is to be able to measure their effect on workers. Yet measuring human aspects--focus, stress, and awakeness--has traditionally been done manually, making it difficult to efficiently apply Taylorism's scientific management. And yet the benefits of doing so are clear; measuring human aspects could provide better management of disruptions~\cite{iqbal2005effectiveness,zuger2017reducing}, automatically adjust lighting to circadian cycles to reduce sleepiness among shift workers, or continuously monitor stress levels to trigger pro-active interventions. 
%
%Fortunately, with recent advances in sensing technologies there is increasing hope that we can accurately and automatically measure human aspects. A number of recent studies have discussed and investigated the use of biometrics to measure aspects such as a person's cognitive and emotional states~\cite{beatty82,picard1995affective,Kramer90,Rowe98,richter1998psychophysiological,Wilson2002}. Specifically, studies have looked at using biometrics to measure stress~\cite{dishman2000stress,Setz2010}, or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004}. Yet, very few studies measure productivity-related factors in the workplace over an extended period of time, and little is known about measuring the more abstract concept of focus. Given the importance of focus, described as a combination of engagement and challenge in a work task by Mark et al.~\cite{mark2014bored}, to an office workers' productivity, this factor deserves further study. 
%
%
%Especially with the advances in sensing technologies, there is an increasing number of studies that discussed and investigated the use of biometrics to measure aspects such as a person's cognitive and emotional states~\cite{beatty82,picard1995affective,Kramer90,Rowe98,richter1998psychophysiological,Wilson2002}. Studies have also looked more specifically at using biometrics to measure stress~\cite{dishman2000stress,Setz2010}, or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004}. Yet, only very few studies have looked at measuring such productivity-related factors in the workplace over an extended period of time. Also, little is known about measuring a more abstract concept of focus that is often used in the context of knowledge workers' productivity and was, for example, described as a combination of engagement and challenge in a work task by Mark et al.~\cite{mark2014bored}.

%Thus, the objective of our research is to continuously (and automatically) measure focus, awakeness, and stress in the workplace. These measures are targeted with an eye towards improving knowledge workers' productivity and well-being in the future. For instance, by automatically protecting the worker from audible interruptions during a high focus period by showering him/her with white noise or by recommending stress-reducing interventions such as Tai Chi during an extended period of high stress. 
%
%Our work builds upon previous work and extends it by performing an eight-week study in the workplace with 14 knowledge workers, collecting biometric and computer interaction data as well as experience samples on focus, awakeness, and stress. Our participants, who perform various job functions for a research and development group within a single large corporation, wore a single biometric armband sensor\footnote{Biovotion Everion sensor~\cite{everion}} with low invasiveness that captured heart- and skin-related measures. This modality was chosen to ease longitudinal deployment in the field. Using machine learning, we created classifiers and analyzed their ability to predict the three productivity-related human aspects. In addition, we compared biometric features with computer interaction features in their predictive power, we analyzed .. and ...
%
%The results of our analysis show that biometrics of a single minimally invasive sensor can be used to ... accurately, with the abstract concept of focus (predictably) being the hardest to detect, yet even its detection within reasonable limits was feasible. \tf{Chris: is it possible to state this?} The results also show that knowledge workers' self-reported levels of stress, focus and awakeness and their physiological manifestation and prediction can vary a lot between individuals. Our results further show that ...
%The main contributions of our work are the creation and analysis of measures for the automatic monitoring of knowledge workers' awakeness, stress, and the more abstract concept of focus in the workplace based on an eight-week field study with 14 office workers. a thorough analysis of ... and a discussion on the ...

%
%combine multiple factors, also including a more abstract concept of focus that is highly related to attention, but also engagement and challenge~\cite{mark2014bored}
%compare it with computer interaction sensors that are less invasive
%focus on a single biometric sensor modality to support longitudinal deployment and comfort

% Yet, only few studies looked at a more complete set of productivity-related factors
%focus, awakeness and stress predicting these productivity-related factors in combination, comparing biometrics to 
%I'd say it's the assessment of predicting multiple factors, using two different types of measurements (biometric and computer interaction), and doing all of it in practice over an extended period of time (every few are more than a few hours).

%stressed: high arousal, low valence
%tired: low arousal

%
%
%
%%\todo{NAK: I have heard from either Thomas or CGM a figure that captures how many problems are due to human error (something like 80\%). If we can find that cite, it should figure prominently in the motivation.}
%
%%Control rooms, while not much different from standard office settings on the surface, differ in the immediacy of their impact. In emergency dispatch offices, factory control rooms, and mining operations centers decision immediately impact safety and profits. A sleepy factory operator that misses early warnings could lose millions of dollars due to a process shutdown; a stressed mining operator that mistakenly dispatches two underground trucks to the same location could cause a deadly crash.  In these demanding settings  the environment must be finely tuned to support the job at hand. For instance, on overnight shifts, if controlling the lighting helps workers shift their circadian sleep cycle, reducing their sleepiness, it should be done, as the consequences of bad decisions are high.
%
%%To ensure the best decisions are made we focus on reducing stress, increasing focus and awakeness. Unfortunately, beyond a few lightly-studied phenomenon such as temperature's affect on awakeness, there are few absolutes that can be applied to workplaces wholesale. Thus, when optimizing a workspace tracking the affect of a given change, if any, is key. 
%
%%The challenge with tracking the affect of a given change in a workspace on its workers is that it is most often done via self-reporting, or the process of filling out daily, hourly, or worse surveys on one's current awakeness, stress, and focus. Worse, this process must be repeated every time a change is implemented. Performing this process rapidly becomes impractical since the changes in the environment may be very frequent  (e.g., modifying the room temperature several times a day). Previous studies show that interruptions that undesired constant interruptions that happen at inopportune moments have negative effects, causing more stress and frustration~\cite{adamczyk2004if,czerwinski2000instant,mark2008cost}.
%
%% The importance of the problem
%Human decisions in the workplace directly impact worker productivity, product/process quality, and workplace safety. Control rooms are noteworthy for the immediacy of impact. Control room operators make hundreds of decisions during a workday, and the consequences of poor decisions can be severe. A sleepy factory operator could cause millions of dollars in losses by failing to recognize the early warning signs of a process shutdown, an unfocused water-processing plant operator could cause a water line failure by missing an alarm warning of a pressure overload, or a stressed mining operator could cause a deadly crash by mistakenly dispatching two trucks to the same underground location.
%
%% The overall problem
%The goal of control room design is to create and maintain an environment that maximizes operator alertness (i.e., that decreases stress and that increases focus and awakeness). Modern control rooms rely on adjustments to ergonomics, light, and temperature to support improved operator alertness. For example, adjusting the lighting in a control room can help operators on overnight shifts to alter their circadian sleep cycles, improving their awakeness while at work. A key challenge in control room design is understanding the effects of a given change to the control room environment on the operators. In particular, reliance on operator self-reporting to capture current levels of stress, focus, and awakeness is cumbersome, distracting, and labor intensive. Indeed, self-reporting can affect the phenomena being measured; e.g., interrupting work to respond to a survey can reduce focus. Moreover, as self-reporting is required not only to create the initial environment, but also to maintain it over time, operator self-reporting is impractical and can be harmful to operator alertness. Previous studies establish that regular interruptions, particularly those that happen at inopportune moments, cause stress and frustration~\cite{adamczyk2004if,czerwinski2000instant,mark2008cost}.
%
%% What others have done
%Previous research addresses how various signals such as blood pressure, heartbeat, and temperature can be linked to psychological states and processes~\cite{Kramer90,Rowe98,Eekelen04,valentini10}. However, little research addresses how biometric signals can be used to predict proxies for alertness such as levels of stress, focus, and awakeness. Instead, previous research has focused on using biometric signals to assess task difficulty in software development~\cite{fritz2014using}, code quality online~\cite{Muller16}, or  interruptibility~\cite{zuger18}.
%
%% What we present in this paper
%\todo{NAK: Would like the opinion of Thomas here --- should we give more of our vision here or stick to the (current) approach of keeping it specific to this paper. That is, do we want to talk about using biometrics to monitor and improve the workplace, or just stick to talking about predicting based on biometrics?}
%The overall goal of our work is to create a machine learning 
%model based in anecdotal biometric
%data which can infer the level of stress, focus, and 
%awakeness of a person by analyzing their biometric data.
%This approach will remove the unnecessary constant interruptions by inferring
%the stress, focus, and awakeness of an individual from their biometric features.
%This will allows to assess if a change performed in the environment 
%of an individual is beneficial
%or prejudicial to their productivity and alertness and, therefore, be able to 
%act upon this information to reverse or increase the change. 
%
%% The contributions of this paper
%The contributions of this paper are:
%\begin{itemize}
%\item The results of an exploratory study on the viability of using biometric sensors to predict the stress, focus, and awakeness of an individual
%\item A machine learning approach to predict proxies for alertness using biometric signals
%\item Guidance regarding the optimal time window and learning curve to generate an effective predictive model
%\item A comparison between the proposed approach using biometric signals and an approach using computer interaction data
%\end{itemize}

### 2019-05-31 BiometricsPaper/bp_relatedWork.tex
added latest version of the paper

%!TEX root = bioPrediction_main.tex
\section{Related Work}
We review two categories of related work for this study. First, we review work related to the productivity-related indicators that we considered in our study, including how they are detected and prevented in several risk-inducing scenarios. Second, we review work related to biometrics, including how they have been studied in the office environment or applied to assess cognitive states and processes.

\subsection{Productivity-related Indicators}

Previous studies have sought to measure, detect, and predict productivity indicators.

\subsubsection{Stress}
Much previous work relates to identifying and mitigating stress in an office environment.
Previous studies have measured stress by taking one of two possible approaches.
The first approach is to measure plasma catecholamine and cortisol as stress biomarkers~\cite{piazza10}.
This approach is impractical for use over prolonged periods of time, as in our eight-week study.
Further, this approach is imprecise because of the delay from the stress stimulation to the stress response, which 
may take from minutes to hours~\cite{Chandola10,Hellhammer09}.

The second approach, which we have chosen for our study, is to measure autonomic nervous system (ANS) activity by analyzing biometric signals of the human body, such as blood pressure, heartbeat, and temperature~\cite{kataoka00,Eekelen04,valentini10}.
In particular, changes in heart rate variability are associated with cognitive and emotional stress~\cite{mcduff16,dishman2000stress}.
This second approach has been used successfully by several past studies~\cite{Force96,gal07,montano09}.

Stress in the workplace and its effects on service providers has been analyzed in call centers~\cite{Hernandez11} and in the context of the perceived imbalance between resources and demands~\cite{cherniss80}. These studies considered several factors such as personality traits, career-related goals and attitudes, and life outside of work, and these factors were correlated with stress levels and burnout.

Evans and Johnson~\cite{evans00} investigated the correlation between noise in the workplace and stress levels.
They found that workers exposed to open-office noise showed aftereffects that indicate motivational deficits.
The population for their experiment comprised 40 female clerical workers, who were randomly assigned to a control condition or to three-hour low-intensity noise room designed to simulate typical open-office noise levels.

Hovsepian et al.~\cite{Hovsepian15} have worked to obtain a standard for continuous stress assessment. They used sensors to conduct a seven-day lab study with 26 participants, as well as a field study with 20 participants.

\subsubsection{Awakeness}

Sleepiness (lack of awakeness) and its associated risk of serious injury to passengers has been studied in the context of automobile 
accidents~\cite{connor02,Nordbakke07}.
These studies show a strong association between the level of acute driver sleepiness and the risk of injury crash.
Connor et al.~\cite{connor02} conducted a population-based case study using the Stanford sleepiness scale, which is similar to a seven-point Likert scale and describes seven different levels of sleepiness from ``Could not stay awake, sleep onset was imminent'' (1) to ``Felt active, wide awake'' (7).
Nordbakke and Sagberg~\cite{Nordbakke07} show that drivers are well aware of  various factors influencing the risk of falling asleep while driving. Drivers also have good knowledge of the most effective measures to prevent falling asleep at the wheel.
However, most of drivers continue driving even when recognizing sleepiness signals, due to the desire to arrive at a reasonable time, the length of the drive, or pre-planed commitments.

%Bader~\cite{Bader95} proposed an approach towards monitoring and estimating a person's awakeness by leveraging on a stationary pressure sensor to track the person's body movements and a correlator to produce a change signal which is compared with a preset wakefulness threshold in a threshold detector circuit.

%A wearable system for mood assessment considering smartphone features and data from mobile ECGs. Exler, Schankin, 
%Looked at energetic-arousal (tired-awake)
%This paper focuses on valence, and barely mentions energetic arousal 
%so I will leave it out

\subsubsection{Focus}
Focus refers to the allocation of limited cognitive processing resources~\cite{Anderson04}.
Mark et al.~\cite{mark2014bored} studied engagement in workplace activities by analyzing the digital activity of 32 information workers in situ for 5 days to understand how attentional states change with context. They found that boredom is highest in the early afternoon and focus peaks in the middle of the afternoon. They also found that doing work that requires focus correlates with stress, while rote work correlates with happiness.

Interruptions in the office are a common barrier keeping workers from sustaining focus on their work related activities, particularly when the interruptions occur at inopportune moments.
Such interruptions may include emails, alerts, or interactions with co-workers\cite{gonzalez2004constant,chong2006interruptions,shamsi07}. Interruptions in inopportune times can have negative effects that range from higher error rate and lower overall performance
to an increase in stress and frustration~\cite{bailey2001effects,czerwinski2000instant,mark2008cost}.
External interruptions may cause workers to enter a ``chain of distraction''~\cite{shamsi07}.
This chain is composed by stages of preparation, diversion, resumption and recovery that result in time away from an ongoing task.

Other studied constructs that relate to focus in the workplace include cognitive absorption, cognitive engagement, flow, and mindfulness.
Cognitive absorption describes periods of time in which a person experiences total immersion in an activity.
This state is also accompanied by a sense of deep enjoyment, a feeling of control, curiosity, and not realizing the passing of time.
It has been associated with ease of use and perceived usefulness of information technology~\cite{agarwal00}.
Cognitive engagement is described~\cite{webster97} as a period of strong focus in an activity without the feeling of a sense of control of the situation.
Flow~\cite{Csikszentmihalyi90}, and mindfulness~\cite{Weick06,dane11} are psychological states that describe periods of prolonged attention and total immersion in an activity. Flow occurs when a person is focused on an activity that requires high challenge and high use of the person's skills, whereas mindfulness is characterized by being aware of fine detail, affording the capacity to discover and manage unexpected events.

\subsection{Biometrics}
Our study investigates the prediction of multiple productivity-related indicators (stress, focus, and awakeness) using two different types of measurements, biometric signals and computer interaction, over eight weeks in a real-life office setting.
Existing work~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015} analyzes a broad array of biometric signals and correlates them with individual's cognitive states and processes.
For example, Zuger et al.~\cite{zuger18} used biometrics to sense interruptibility in the office~\cite{zuger18}.
%and found correlations between biometric signals and predicting code quality~\cite{Muller16}, to assess task difficulty in software development~\cite{fritz2014using}, to assess developer's mental load and perceived difficulty while working on small code snippets, and
Biometric signals have also been studied in the context of technology users. For example, Parnin~\cite{Parnin2011} analyzes electromyography to measure sub-vocal utterances, and how these might be correlated with the programmer's perceived difficulty of programming tasks. Similarly, biometrics have been used to measure code difficulty by using biometric sensors~\cite{fritz2014using}
and using Near Infrared Spectroscopy to measure developer's cerebral blood flow~\cite{Nakagawa2014}.

Eye tracking technology~\cite{Bednarik2006,Crosby1990,Rodeghero2014} and brain activity~\cite{Ikutani2014,Siegmund2014} 
have been used in previous studies to analyze different tasks in an office environment.
Eye tracking has been used to analyze memory load and processing load by inspecting task-evoked pupillary response and pupil size~\cite{beatty82}. Similar studies have shown high correlation between pupil size and mental workload of subtasks~\cite{beatty82} and cognitive load~\cite{Klingner10}.
Brain activity has been associated with different mental states~\cite{Berger29} by analyzing specific frequency bands (alpha, beta, gamma, delta, and theta) using electroencephalography (EEG). The increase or decrease of some of these frequencies is correlated with
attentional demand and working memory load~\cite{Smith05,sterman93}.
In contrast to studies that use eye tracking or EEG, we focused on a less invasive technology that can be applied in a real world scenario.

### 2019-05-31 BiometricsPaper/bp_results_1.tex
added latest version of the paper

%!TEX root = bioPrediction_main.tex
%\input{bp_results_1}
%%\input{bp_results_2}

\section{Analysis and Results}

\begin{table*}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{rq1performance.pdf}
  \caption{Prediction results individual models. The baseline row states averaged results of our baseline classifier, the general row states the averaged results of model trained on all participants. Distribution presents the distribution of the binarized outcome variable.}\label{tab:accuracy}%\todo{revise caption}}
%  The distribution columns show a bar chart of the response distribution (negative/positive) for each of the three variables. The baseline row represents the averaged results of our baesline classifier. The general row shows the averaged results of our models trained on all participants.
  \vspace*{-4mm}
\end{table*}

To evaluate the efficacy of continuously predicting a knowledge worker's stress, focus, and awakeness in the workplace, we trained and tested machine learning classifiers using a leave-one-out cross validation. For prediction, we used the features extracted from the collected data as the input data, and binarized the participants' self-reported responses on stress, focus, and awakeness into two classes each (e.g., 'stressed' and 'not stressed') to use as output measures.

\subsection{Predicting Stress, Focus \& Awakeness}
% random forest best
\label{secOverallAccuracy}
As a first step, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11} and performing a grid search analysis to determine the optimal hyperparameters for each classifier. Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, random forest, and a multilayer perceptron neural network. The optimal values for random forest and the three output measures are listed in Table \ref{tab:hyperparams}. For the remainder of this paper, we will be referring to random forest classifiers trained with these hyperparameters.%\\[-0.1cm]
% for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)
\begin{table}[t!]
	\begin{centering}
    \begin{tabular}{llll}
      \hline
      Variable & K & \# Estimators & Minimum Samples Split \\
      \hline
      Stress & 300 & 100 & 4\\
      Focus & 200 & 50 & 4\\
      Awakeness & 800 & 100 & 4\\
      \hline
    \end{tabular}
    \caption{Hyperparameters selected by grid search analysis for tuning our random forest models. K refers to the number of features selected.}\label{tab:hyperparams}
    \vspace*{-1cm}
    \end{centering}
\end{table}

\noindent\textit{Results of Individual Classifiers}\\
Since peoples' experience of stress, focus, and awakeness (as well as their physiological manifestations) can vary a lot (e.g.,~\cite{Hernandez11}), we first trained individual classifiers for each participant (rather than a general one for all participants). The results of our analysis are reported in Table~\ref{tab:accuracy}. For our analysis, we report values of accuracy, one of the most commonly used metric to compare performance, as well as precision and recall of the classes of interest: 'stressed', 'not focused', and 'not awake'. Since the imbalance in the data can lead to high accuracy values if a classifier always just predicts the most likely/frequent class while ignoring the class of higher importance and interest, precision and recall of the class of interest can be more adequate metrics in this case~\cite{yap2014,bhattacharyya_data_2011,Hernandez11}. 
% However, the imbalance in the data can lead to a classifier with high accuracy if it always just predicts the more likely class while ignoring the class of higher importance and interest in our case (i.e. stressed, not focused, not awake). Therefore, we further report the precision and recall for the three classes 'stressed', 'not focused' and 'not awake'.

Overall, we were able to use the extracted physiological features to predict all three aspects with reasonable accuracy, precision, and recall, as well as to improve upon the baseline---a stratified random classifier that randomly chooses one of the two classes with a bias towards the larger class. While the individually trained classifiers improved on average across all participants upon the baseline in all cases excepting recall of 'not focused', the improvement was substantially higher for awakeness (85\% improvement in precision,  62\% in recall, and 7\% in accuracy) than for stress or focus. Also, the performance of the individually trained classifiers varied greatly across participants. While some participants showed a large improvement, for others the baseline performed much better than the individually trained classifier. For instance, for predicting 'stressed', the individual classifiers improved upon the baseline for S4, S6, S8, S11, S12, and S14 with a maximum improvement of 88.4\% in precision and 184.6\% in recall for S12, while they did worse for S1, S3, S5, S7, S9, S10, and S13, and in the worst cases did not correctly predict a single instance of 'stressed'.\\[-0.1cm]
% We attribute these discrepancies to the large differences in response distributions between participants, as well as to the subjectivity of self-reporting.
%
%(improvement: 8\% precision, 1\% recall, 8\% accuracy), the individual performance varied greatly among participants. Some participants showed negligible difference in comparison to the baseline classifier, e.g. subject ..., while others showed a large improvement, e.g. subject ...
%\begin{table}[h!]
%\begin{centering}
%\begin{tabular}{lll}
%\hline
%Participant & Recall Improvement (\%) & Precision Improvement (\%)\\
%\hline
%S12 & 184.6 & 88.4 \\
%S6 & 66.7 & 85.8 \\
%S11 & 50.0 & 44.4 \\
%S8 & 32.4 & 26.7 \\
%S14 & 14.7 & 9.8 \\
%S4 & 5.6 & 5.5 \\
%S9 & -55.4 & -46.8 \\
%S3 & -90.0 & -82.1 \\
%S1 & -100.0 & -100.0 \\
%S5 & -100.0 & -100.0 \\
%S7 & -100.0 & -100.0 \\
%S10 & -100.0 & -100.0 \\
%S13 & -100.0 & -100.0\\
%S2 & - & -\\
%\hline
%\end{tabular}
%\caption{Percentage improvement in recall and precision for stress, using our approach compared to the baseline, on an individual level. For participant 2, the baseline achieved a precision and recall of 0, thus the improvement is undefined}
%\end{centering}
%\label{tab:indImprovement}
%\end{table}

\noindent\textit{Feature Selection and Importance}\\
There are a large variety of features that can be (and have been) calculated in previous research for each of the basic measurements listed in Table~\ref{signals}, such as the mean, standard deviation, maximum, and interquartile range. In addition, each of these metrics can be combined with the various time windows captured of a basic measurement, resulting in a large feature space. To reduce the feature space, we experimented with multiple feature selection methods, including selecting the top k highest correlated features by various metrics such as mutual information, Pearson's correlation coefficient, ANOVA's F-value, as well as wrapper methods such as recursive feature elimination, optimizing mean decrease accuracy by iteratively permuting features, and only selecting features that exceed a certain Gini importance threshold. We found that all methods produced similar results with respect to accuracy, precision, and recall for the individual models. Ultimately, we chose to use the top k features with the highest ANOVA F-value, as it is relatively simple and efficient to calculate. The values of k used were selected by grid search analysis, and are shown in Table \ref{tab:hyperparams}.

%Respiration rate is also highly important for stress (#3 , 14.4%) but I'm not sure what previous works say about correlation with stress
Overall, the features that were selected as the important ones for the individual models based on the random forest algorithm varied greatly across participants. Yet, some feature categories were considered to be important more frequently than others. Table~\ref{tab:featureImportance} shows the averaged Gini importance for the feature categories used for predicting stress. Particularly important for stress were the feature categories heart rate variability (18.3\%) and skin temperature (15.7\%), both of which have been shown in several previous studies to be  indicators for stress~\cite{dishman2000stress,mcduff16,kataoka00}. For predicting focus, the feature categories for blood pulse wave (14.2\%) and heart rate variability (13\%) showed to be the most important categories, while for awakeness the most important ones were heart rate variability (13.6\%) and blood pulse wave (13.1\%).\\[-0.1cm]

\begin{table}[t!]
  \begin{centering}
  \begin{tabular}{llll}
    \hline
    Feature Category & Stress & Focus & Awakeness\\
    \hline
    Heart Rate Variability & 18.3\% & 13\% & 13.6\%\\
    Blood Pulse Wave & 10\% & 14.2\% & 13.1\%\\
    Heart Rate & 8.7\% & 12.6\% & 10.3\%\\
    Skin Temperature & 15.7\% & 9.8\% & 10\%\\
    Galvanic Skin Response & 6.6\% & 8.2\% & 5.1\%\\
    Respiration Rate & 14.8\% & 12.7\% & 10\%\\
    Oxygen Saturation & 5.6\% & 4.3\% & 2\%\\ 
    Energy Expenditure & 6\% & 7.7\% & 4.8\%\\
    Activity & 4.6\% & 7.8\% & 7.6\%\\
    Steps & 1.7\% & 0.8\% & 0.9\%\\
    Time of Day & 0\% & 0.1\% & 0.5\%\\
    \hline
  \end{tabular}
  \caption{Averaged Gini importance of each feature category per response variable.}
  \label{tab:featureImportance}
  \end{centering}
  \vspace*{-7mm}
\end{table}

\noindent\textit{Individual vs.\ General Model}\\
Individual models are trained specifically for each individual and thus require a data collection period before they are capable of making accurate predictions. On the other hand, the idea of general models is to be able to train them on already collected data and then  to be able to apply them even to new and unseen individuals, thus overcoming the cold-start problem. Given the big individual differences in biometrics, training a general model to achieve an adequate accuracy for new individuals is not necessarily possible.

To examine the performance of a general model for our participants, we trained three general models, one for focus, one for awakeness and one for stress. We roughly followed the same procedure as for the individual models. Due to the larger amount of data available in the general case, we used the more common random undersampling, which randomly selects elements in the majority class to exclude from the dataset, instead of random oversampling to balance the distribution of the dataset. The models were trained on the datasets of 13 of the 14 participants, and then evaluated on the dataset of the last, repeating this process for all 14 participants.

The bottom row of Table \ref{tab:accuracy} presents the averaged performance results for this approach in terms of accuracy, precision, and recall. Although the averaged precision and recall are comparable or better than those of the averaged individual results, this was at the cost of a large decrease in overall accuracy. Upon closer investigation into the performance of the general model when testing on each participant, we found that individually trained models for each participant performed much better than a general model trained over all participants. Using stress as an example, for participant S12, for whom we saw the greatest increase compared to the baseline in individual models, the general model was unable to predict a single instance of 'stressed' correctly. This is consistent with our expectations because biometric features are highly specific to individuals.

%\todo{put the content of the following sentence somewhere into the discussion; We attribute these discrepancies to the large differences in response distributions between participants, as well as to the subjectivity of self-reporting.}

### 2019-05-31 BiometricsPaper/bp_results_2.tex
added latest version of the paper

%!TEX root = bioPrediction_main.tex

\subsection{Minimum Number of Training Samples}\label{secLearningCurve}
Collecting experience samples from users is expensive, since participants are being interrupted several times a day and have to answer the questions. To minimize the number of samples to be collected from participants, we examined how the performance of individual classifiers changes over the number of samples used to train the classifier.

Participants in our study had varying levels of responsiveness to the experience sampling ranging from 10 to 76 (see Figure~\ref{responseDistribution}). To maintain a certain generalizability while also being able to examine a range of sample numbers for training the classifier, we decided to include the ten participants with the most samples for this analysis. As a result, we had at least 34 survey responses per participant to examine the learning curve of the classifiers. For our analysis, we thus performed a leave-one-out cross validation with random sample sets of size 1 to 33 and calculated the average through all folds of the validation.

For each of the three productivity-related aspects, we are again more interested in predicting when a worker is stressed, not awake, or not focused, the less common class in all three cases. Since the less common class can be very small, we weighted each participants' classifier performance by the percentage of the samples in this smaller class.

\begin{figure}
  \centering
          \includegraphics[width=0.5\textwidth]{20180912AwakenessLC2.png}
  \caption{Performance of the per participant trained awakeness classifiers, measured in precision, recall (sensitivity), and F-score. The dotted red line represents the F-score trend.}\label{fig:learningCurveInd}
  \vspace*{-3mm}
\end{figure}

\vspace{0.05in}
\noindent\textit{Individual Classifiers for Binary Prediction}\\
The averaged performance of all individually trained random forest classifiers for 'awake' with respect to the training sample size is presented in Figure~\ref{fig:learningCurveInd}. The trend indicates a positive correlation between the number of samples in the training set and the classifiers' F-score performance, with an overall improvement of  114\% (from 0.14 to 0.30) in the F-score between a training set of one sample to one with 33 samples. The trends for
the remaining indicators are 29\% for stress and no overall improvement for focus.\\[-0.1cm]

%\noindent\textit{}
%When contrasting
%models trained and tested on the 
%biometric data of each user individually, against
%a general model trained and tested with the data of all users. 
%We can conclude
%that for up to 33 training samples analyzed in this study,
%the former has better performance when predicting the
%responses of each user than when training 
%a model with the data of several different users. 
%This is partially due to the 
%variability across users and the subjective nature
%of the responses (e.g.,where slightly awake for one person
%may be not awake for another).

\begin{figure}
  \centering
      \includegraphics[width=0.5\textwidth]{20180914Awakeness5PointScaleOnly15Lines.png}
  \caption{Performance of the per participant trained classifiers for predicting 5-point awakeness, measured in root mean squared error.}
   \label{fig:learningCurve5}
\end{figure}

\noindent\textit{Predicting Five Classes}\\
In a second step, we analyzed a more fine-grained prediction using the initial 5-point Likert scale responses rather than the binarized ones as output measure. Figure~\ref{fig:learningCurve5} depicts the performance of the individually trained classifiers in terms of the root mean squared error. The root mean squared error represents the distance of the predicted from the actual value, which provides a more nuanced measure of the performance in the fine-grained prediction case. The figure shows a similar trend as for the binarized prediction, in that the root mean square error averaged over all ten participants decreases with more samples (from 0.49 to 0.41 root mean square error) and thus the performance increases. At the same time, the figure also shows that the performance results for the fine-grained prediction, again, vary substantially across participants.

\subsection{Minimum Time Window}\label{secMinimumTW}
In general, the less biometric data is needed to accurately predict a certain outcome measure, the easier and faster the analysis and data collection. To examine the optimal and minimum time window for the prediction of stress, focus, and awakeness, we used 16 different time windows from 10 seconds to 3 hours.% as depicted in Figure~\ref{timeWindows}. 
For our analysis, we then trained individual classifiers for each of the 16 time windows, using only features that had a time window smaller or equal to the time window rather than using all combinations of \small{$\{Biometric Measures\} X \{Statistical Metrics\} X \{Time Windows\}$}. We again used random forest and a leave-one-out cross validation to train individual classifiers. Since the number of features used for the training changed with each time window, we did not apply our feature selection in this case, but used all features available. Finally, due to the imbalance in the data, we again weighted each participants' classifier performance by the number of instances in the smaller class to calculate the average.

%\begin{figure}
%  \centering
%      \includegraphics[width=0.4\textwidth]{timeWindows.png}
%  \caption{Time windows of biometric data collected prior to each survey response: 10sec, 20sec, 30sec, 45sec, 1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, and 3hour.}
%   \label{timeWindows}
%\end{figure}

\begin{figure}
  \centering
      \includegraphics[width=0.5\textwidth]{20180912AwakenessTWBars.png}
  \caption{Performance (F-score) of individual classifiers trained on the different time windows to predict `awake'.}
  \label{timeWindowsPandR}
  \vspace*{-3mm}
\end{figure}

Figure~\ref{timeWindowsPandR} shows how the F-score changes for predicting `awake' over the 16 different time windows. The figure shows an increasing trend in the F-score, i.e. the higher the number of included time windows included, the higher the F-score. However, there is one exception, the time window of 1200 seconds that achieves a performance close to the one for the time window 10,800 seconds (3 hours), at which point all features are included. Overall, our results thus show that while using all time windows up to 3 hours performs best, and outperforms the feature set that is solely based on a 10 second time window by 28\% (from 0.18 to 0.24), the performance for a time window of 1200 seconds is already very close to optimal (everything up to 3 hours in our case).

%We also compared the performance of a model created per user against
%a model created across users.
%To calculate the performance of the model created per user,
%for each user we create a model using the features 
%related to each time window for the selected user only, 
%while in the model created with data across users,
%we trained the model with the data related to each
%particular time window of all users.
%our results show that the personalized model created
%for each user outperforms in every case
%the model created the with data across all users.

%This study
%shows that in general larger time windows predict stress,
%focus, and awakeness more precisely than 
%shorter time windows. We can also notice
%a very gentle upwards trend
%with a inflexion points at 45 seconds and 180.
%There is a peak of performance in 45 seconds, 
%time window that could be used if time is a scarce resource.
%The performance stabilizes around 180 time window
%which indicates that collecting data for a longer time
%period will not increase significantly the performance of the approach.
%The difference in performance between the 
%lowest performing and the highest performing time window is 
%less than a 10\% improvement.

\subsection{Computer Interaction Data} \label{secCI}
Given our focus on knowledge workers (i.e., workers who generally spend a lot of time interacting with information on their computer at work), we also analyzed the use of computer interaction features to predict focus, awakeness and stress. To collect computer interaction data, we used an open source computer interaction monitor (reference omitted for double-blind.)
%the open source PersonalAnalytics project\footnote{https://github.com/sealuzh/PersonalAnalytics}  \cite{meyer18} 
to track participant's mouse and keyboard activity, as well as details about their active window. The specifics of the features tracked are listed in Table \ref{tracker}. The tracker was installed on the computers of 10 of the 14 participants, with participants S6, S10, S12, and S13 opting out of this part of the study due to privacy concerns.  Therefore, we limited this analysis to the 10 participants for whom we could calculate all features.

\begin{table}
\begin{center}
\begin{tabularx}{\columnwidth}{l l}
\hline
Feature collected by tool & Description \\ 
\hline
Total keystrokes per min& Sum of all types of keystrokes \\ 
Normal keystrokes per min&F[h] Not backspace and navigation \\ 
Backspace keystrokes per min& Backspace keystrokes \\ 
Navigation keystrokes per min& Arrow key keystrokes \\ 
Total clicks per min& Sum of all click types \\ 
Other clicks per min& Not right and left clicks \\ 
Left clicks per min& Left clicks \\ 
Right clicks per min& Right clicks \\ 
Scrolled distance per min& Scrolled distance in pixels \\ 
Moved distance per min& Mouse movements in pixels \\ 
Activity switches per min& Browser window title changes \\ 
Category switches per min& Activity performed category \\ 
\hline
\end{tabularx}
\caption{List of computer interaction features.}%~\cite{meyer18}}
\label{tracker}
\end{center}
\vspace*{-7mm}
\end{table}

For calculating computer interaction features, we again used the aforementioned 16 time windows and scaled the computer interaction values if the time windows did not align. For our comparative analysis of the different sensing techniques---biometrics vs computer interaction---we then created two new feature sets for each participant in addition to the biometric one: one with only computer interaction features, and one with computer interaction features plus biometric features.

Table~\ref{ciPerformance} lists the results of our analysis. The results show that in all cases, the computer interaction based model was able to improve upon the biometric model in terms of precision and recall, but not in accuracy. Further, we found that the combined model was the most effective model in terms of precision and recall for predicting stress and awakeness overall, but performed slightly worse than the model using only computer interaction features for focus. 
%Since we consider precision and recall for predicting the class of higher interest, i.e. stressed, not focused, not awake, to be the most important statistics when interpreting our results, we compared the models against each other by averaging these two statistics.

\begin{table}
\begin{center}
\begin{tabularx}{\columnwidth}{llllll}
\hline
Model/Feature Set & Accuracy & Precision & Recall & F-Score\\
\hline
\textbf{Awakeness}\\
\hspace{3mm}Biometrics only & 0.808 & 0.269 & 0.314 & 0.289\\
\hspace{3mm}CI  & 0.758 & 0.425 & 0.362 & 0.391\\
\hspace{3mm}Biometrics + CI & 0.791 & 0.390 & 0.404 & 0.400\\
\hline
\textbf{Stress}\\
\hspace{3mm}Biometrics only & 0.775 & 0.270 &	0.260 & 0.265\\
\hspace{3mm}CI  & 0.698 & 0.290 & 0.272 & 0.281\\
\hspace{3mm}Biometrics + CI & 0.712 & 0.317 & 0.286 & 0.301\\
\hline
\textbf{Focus}\\
\hspace{3mm}Biometrics only & 0.716 & 0.251 & 0.256 & 0.253\\
\hspace{3mm}CI  & 0.742 & 0.332 & 0.342 & 0.337\\
\hspace{3mm}Biometrics + CI & 0.745 & 0.340 & 0.316 & 0.328\\
\hline
\end{tabularx}
\caption{Comparison of prediction performance for the 3 different features sets for 10 participants. CI stands for computer interaction.  Precision/recall refer to prediction of more important classes, i.e. stressed, not awake, not focused.}
%Comparison of the performance of predicting stress, focus and awakeness using the 3 different feature sets for the 10 participants. The performance is calculated as the average of the performance of the individual classifiers. Computer Interactions is abbreviated as C.I. here for readability. Precision and recall refer to the prediction of the more important classes, i.e. `stressed', `not awake', `not focused'.}
\label{ciPerformance}
\end{center}
\vspace*{-7mm}
\end{table}

As with the biometric models, the individual performance of both the computer interaction only models and the combined models varied quite a bit between participants. Using stress as an example again, in the computer interaction models 5 of the 10 participants saw improvements compared to the baseline, with a maximum improvement of 128\% in precision, and 78\% in recall. In the combined model for stress, 5 of the 10 participants saw improvements compared to the baseline, with a maximum improvement of 117\% in precision and 95\% in recall. Neither model was capable of correctly predicting any instances of 'stressed' for participant S7.

Since the number of features changes  depending on which feature set is used, we adjusted the feature selection parameter for each of the computer interaction and combined computer interaction/biometric models. The values reported in this section were achieved using the optimal feature selection parameters we found, which are shown in Table \ref{ciFeatureSelection}.

\begin{table}
\begin{center}
\begin{tabularx}{\columnwidth}{lc}
\hline
Model/Feature Set & Number of Features Selected\\
\hline
\textbf{Stress}\\
\hspace{3mm}CI & 400\\
\hspace{3mm}Biometrics + CI & 800\\
\hline
\textbf{Focus}\\
\hspace{3mm}CI & 20\\
\hspace{3mm}Biometrics + CI & 300\\
\hline
\textbf{Awakeness}\\
\hspace{3mm}CI & All\\
\hspace{3mm}Biometrics + CI & 50\\
\hline
\end{tabularx}
\caption{Optimal number of features used for each of the three models. CI: computer interaction}
%The optimal number of features we found to select for each of the model/feature set combinations. Computer interactions is abbreviated as C.I.}
\label{ciFeatureSelection}
\end{center}
\vspace*{-7mm}
\end{table}

### 2019-05-31 bp_conclusion.tex
merging the paper folder and the project folder

%!TEX root = bioPrediction_main.tex
\section{Conclusions} 
Stress, awakeness and focus at work are highly relevant aspects when it comes to productivity and well-being at the workplace. In this paper, we presented the results of a study with 14 professional knowledge workers in their workplace over an eight week period to examine the ability of biometrics for predicting these producitivity-related aspects. The longitude and in situ placement of the study---two key components of the study---as well as the breadth of human aspects examined, support and extend previous work. Based on twice collected daily survey responses and continually collected biometrics data we were able to create a model that was able to predict awakeness, stress, and focus, outperforming the baseline by as much as 84.6\% in the case of awakeness. The analysis further shows that there is a continuous increase in performance with the number of training instances, suggesting that there is more potential to improve the model's performance with further data points. Interestingly, while aspects such as stress have previously been predominantly linked to biometrics, our comparison with a model build from computer interaction features shows that even these types of features might bear big potential in predicting some of these productivity-related human aspects in the workplace.

The results also open up new opportunities to help increase knowledge workers' productivity and well-being, ranging from instantaneously taking action to prevent potentially risky situations and prevent accidents due to a lack of focus or awakeness, all the way to recommending interventions to reduce stress if it becomes more chronic.

% \todo{as you can see, the conclusion is way too long and has some cut and paste repetition}

% Stress, awakeness and focus at work are highly relevant aspects when it comes to productivity and well-being at the workplace. In this paper, we presented the results of a study with 14 professional knowledge workers in their workplace over an eight week period to examine the ability of biometrics for predicting these producitivity-related aspects. The longitude and in situ placement of the study as well as the breadth of human aspects examined, support and extend previous work. Based on twice collected daily survey responses and continually collected biometrics data we were able to create a model that was able to predict , outperforming the baseline by as much as 84.6\%, in the case of awakeness. We also show that there is a positive correlation between number of training instances and model performance--a 114\% improvement from training with one sample to training with 33 samples--suggesting that with a larger amount of training data prediction would continue to improve. Due to this, while current user-specific models have higher performance than across-user models, there is hope that with enough data even across-user models could become accurate.

% Our analysis shows that we are able to \todo{fill this}

% Our study with 14 professional knowledge workers in their workplaces over an eight week period has shown that we can use biometrics to predict these producitivity-related aspects of an individual knowledge worker. The longitude and in situ placement of the study as well as the set of human aspects examined support and extend previous work. In addition, our analysis

% as well as computer interaction features to predict these aspects of an individual knowledge worker
% has shown that we can use biometrics as well as computer interaction features to predict these aspects of an individual knowledge worker over a long period of time in situ
% The results also show that the general ... \todo{fill this}

% In this study we analyze biometric signals of office workers to predict the productivity-related human aspects of stress, focus, and awakeness. A key component of this study was its length, an eight week period, longer than previous studies by a factor of four~\cite{zuger18,Muller16}, and its realism , studying professionals in situ instead of students. We collected information from 14 participants from different departments within the company, covering different ages ranges, genders, and work experience. By learning from twice daily survey responses and continually collected biometrics data we were able to create a model that was able to predict human aspect values, outperforming the baseline by as much as 84.6\%, in the case of awakeness. We also show that there is a positive correlation between number of training instances and model performance--a 114\% improvement from training with one sample to training with 33 samples--suggesting that with a larger amount of training data prediction would continue to improve. Due to this, while current user-specific models have higher performance than across-user models, there is hope that with enough data even across-user models could become accurate.

% From these results emerge a series of opportunities to help increase awakeness, focus, and reduce stress in office workers. In the near future, we will therefore be able to detect a decrease in productivity-related aspects following our approach, and be able to take action to prevent potentially risky situations.

%\begin{acks}
%The acknowledgments section will be included for the camera-ready version of the paper.
%\end{acks}

### 2019-05-31 bp_data.tex
merging the paper folder and the project folder

%!TEX root = bioPrediction_main.tex

\subsection{Data Preparation}

In this section we describe how we preprocessed the collected data for use in training and testing machine learning models.

\subsubsection{Data Linking}

We linked the collected biometric data and survey responses for each participant. Linking the data is necessary to construct training and test datasets for use in creating and evaluation machine learning models.

Our linking approach is as follows. From the start time of each survey response, we look back one hour for available biometric data. For each minute in that hour-long time window, we check for biometric data to associate with the survey response. For example, if a participant started a survey response at 11:05am, we look for biometric data in the time frame 10:05am to 11:05am. If biometric data is available in the hour-long time window, we consider the survey response to have associated biometric data. Otherwise, we exclude the survey response from our dataset.

Reasons for a survey response to lack associated biometric data include:
\begin{itemize}
\item The participant not wearing the Everion in the hour before before beginning the survey
\item The Everion not recording data in the hour before the participant began the survey (e.g., due to low battery)
\item Data not being uploaded successfully to the server
%\item Compatibility issues between the sensor and the OS tracking the participant's data
%\item Issues accessing the data uploaded by the participants
\end{itemize}

Figure~\ref{surveyBio} illustrates the number of survey responses with associated biometric data for each study participant. Participant S2 and S12 have particularly low numbers of usable survey responses. In each of these cases, the issue related to biometric data not being uploaded successfully to the server.

\begin{figure}
  \centering
      \includegraphics[width=0.45\textwidth]{DuringTheDay.png}
  \caption{Availability of biometric data per participant. Green: survey responses with biometric
  data; red: responses with no biometric data.}
   \label{surveyBio}
   \vspace*{-6mm}
\end{figure}

\subsubsection{Feature Extraction}
We extracted features from the biometric data to provide as input to machine learning models. Previous studies~ \cite{vorburger05,zuger2015interruptibility} identify time windows as an important factor that impacts the prediction accuracy of a classifier. We considered many time windows from the literature on biometric analysis~\cite{zuger18}, ranging from 10 seconds to 3 hours. Specifically, we considered the following time windows: \textit{10sec, 20sec, 30sec, 45sec, 1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, 3hour}.

From the start time of each survey response, we look back the amount of time that corresponds to each time window, and we create features for all of the biometric data available in that time window. For example, if a participant started a survey response at 11:05am, for the 30min time window, we create features using all of the available biometric data from 10:35am to 11:05am. For each time window, we calculate 10 statistical measurements from the biometric data to create 10 distinct features. Specifically, the 10 statistical measurements are: mean, standard deviation, variance, median, percentile25, percentile75, interquartile range, maximum, minimum, and range. Thus, for each survey response, we generate a large number of corresponding features based on three factors: biometric measurement, time window, and statistical measurement. In addition to these biometric features, we also considered the time of day in which the questions were asked.

\subsubsection{Response Transformations}
Table~\ref{responseDistribution} illustrates the distribution of responses from each participant for each of the three survey questions (which are listed in Section~\ref{sec:Surveys}). The figure shows that there is a notable imbalance in the distribution of the self-reported responses provided by the participants. Most participants did not use all five points of the five-point Likert scale in their responses, and the distributions tend to skew toward one side or the other, depending on the question. Thus, we binarized the survey data into a two-point scale to give the machine learning models the best possible chance to make useful predictions. The two points in the binary scale represent negative or positive responses for each of the three human aspects of interests (e.g., not stressed or stressed).

We binarized the survey responses as follows. For each participant, we calculated the median response value for each question. We classified each response below the median as 0 ('negative') and each response above the median as 1 ('positive'). The distribution for the stress question skewed left, so we included the median values in the 'positive' class, while the distributions for focus and awakeness skewed right, so we included those median values in the 'negative' class.

\subsubsection{Oversampling}
Even after binarizing the responses as described in the previous section, we found the distribution of responses was still quite imbalanced for many of our participants. This can be seen in the distribution columns in Table \ref{tab:accuracy}. To combat this, we applied random oversampling to our training sets, which artificially rebalances the dataset by creating randomly replicated data in the minority class. This has been a commonly used technique in previous studies on unbalanced datasets \cite{chawla2004,yap2014}.

\begin{table}
  \centering
      \includegraphics[width=0.5\textwidth]{distributiontable.pdf}
  \caption{Distribution of responses per participant over the 5-point Likert scale (1: not at all; 5: extremely) for each question. }
   \label{responseDistribution}
   \vspace*{-6mm}
\end{table}

### 2019-05-31 bp_discussion.tex
merging the paper folder and the project folder

%!TEX root = bioPrediction_main.tex
\section{Discussion}

\vspace{0.05in}
\noindent
\textbf{Target Workspaces:}
Currently, the cost of biometric sensors and necessary infrastructure, such as automated light and sound systems for adjusting the environment, makes our approach most appropriate for high-value workspaces, such as control rooms, command centers, or dispatch offices. However, as standard office settings become more personalizable (e.g., via adjustable desks, lighting, and sound showers) and sensor costs decrease, our approach could be applied to any office environment, and thus could impact a large percentage of modern workers. As in modern cars, temperature and lighting could be regulated on a per-person basis, which would allow the environment to react to the person's current state and to maximize each person's preferences and productivity (e.g., preferences of men and women in temperature~\cite{Karjalainen07}). Further research would be needed to identify how to balance needs across a group of office workers and how to handle conflicting levels between different group members.

\vspace{0.05in}
\noindent
\textbf{Imbalanced Data:}
Study participants provided highly imbalanced data in their survey responses, with most participants only taking advantage of a subset of the Likert-scale values and the data points mostly being clustered around the middle of the scale, as can be seen in Table~\ref{responseDistribution}. While some of the imbalance is expected due to certain classes, such as 'not stressed', being more common in the workplace, this imbalance also provides challenges in the training and assessment of a machine learning classifier, as also found by others, e.g.~\cite{Exler16}. We addressed this for the training by oversampling in case of few data samples for the individual models and undersampling in case of a general model where more data was available. Especially in light of this imbalance in the data, the results we achieved with our models are encouraging. For the assessment of the classifiers' performance we addressed the imbalance by not just presenting accuracy, but also by focusing on prediction and recall to examine the classifier's performance in predicting the infrequent (yet more important) cases, such as when a user is struggling to stay awake and an intervention or warning might be needed most. 
% \textbf{Unbalanced Data:}
% As shown in Table~\ref{responseDistribution}, the typical participant provided highly unbalanced survey responses, with most responses exhibiting a central tendency. Many participants provided zero or one very high (or very low) value on the Likert scale. In spite of this, our results show that by using biometric sensors and leveraging a machine learning approach, we are able to predict stress, focus, and awakeness better than the baseline, with improvements up to 84.64\%. Taking into account that the model correctly predicts many cases that happen infrequently, this is especially promising. These infrequent cases, such as when a user is highly stressed or struggling to stay awake, are in fact the most important cases to detect, as they are when intervention is needed most.

\vspace{0.05in}
\noindent
\textbf{Ground Truth and Self-Reporting:} 
One of the key points for developing a good classifier for awakeness, focus and stress, is to collect a valid ground truth to be used as the output measure. When designing the study, we therefore spent an extensive amount of time on determining the exact questions to ask in the experience sampling, consulting experts in the area, and basing the questions and wording on previous research and studies. Yet, the reliability and validity of self-reports have been questioned in the past due to subjective biases, lack of care in reporting, and the highly individual nature of reporting aspects such as stress~\cite{Hernandez11,Hovsepian15}. In addition, in contexts such as the workplace, employees might be afraid to genuinely report levels of aspects, such as sleepiness. Hence, there is a chance that the collected experience samples do not adequately reflect the ground truth of the underlying variable under investigation. It could even be the case that certain biometrics might represent a more accurate ground truth of the studied phenomena than the self-reports. This suggests that a more confirmative study rather than an inquiry study could be a better approach, and we will explore such routes in future work.

%\vspace{0.05in}
%\noindent
%\textbf{Corpus Size:}
%In this study we collected data from 14 professionals over the course of eight weeks. By scholarly standards this constitutes a large dataset --- especially given that the participants were not students. However, in the context of machine learning this corpus is relatively small. Thus, the model's overall performance, while promising, is only an indication of the performance that a larger corpus could provide. Our results show that there is a positive correlation between data added to the training corpus and the performance level of the model. This upwards trend likely continues, but a larger dataset is needed to confirm or reject this hypothesis. 
%
%
%\vspace{0.05in}
%\noindent
%\textbf{Per-User Model vs. Across-Users Model:}
%Our results show that individually trained models outperform on average a more general model. Yet, given the increase in model performance with training data set size, we expect that with enough data, a generic model will perform reasonably close to the individually trained models. This would have significant practical implications, eliminating the need for survey-based feedback (which is only necessary for model training), and thereby completely automating measurement of human aspects. Our follow up work will focus on gathering more data toward this end.
%
%
%% \vspace{0.05in}
%% \noindent
%% \textbf{Per-User Model vs. Across-Users Model:}\
%% While we do see variations across participants, and thus our work shows that personal 
%% models perform better than generic models, we expect that with enough data, a generic 
%% model may perform equally well. This would have significant practical implications, eliminating the need for survey-based feedback (which is only necessary for model training), thereby completely automating measurement of human aspects. Our follow up work will focus on gathering more data toward this end and on evaluating the performance of a generic model.
%
%% \vspace{0.05in}
%% \noindent
%% \textbf{Data Collection:} 
%% One advantage of our study is that it was conducted on professionals in a real office setting (i.e., not on students). However, the disadvantage to working with professionals is that business deadlines and other pressing real world issues lead to missing survey responses. Additionally, busy professionals often fall into the trap of always selecting the same value for a given question that they have answered many times, which reduces the accuracy of the collected data. To minimize the effect of this inertia, we advise collecting data at typical low-productivity times, such as in the early afternoon or on Mondays~\cite{mark2014bored}.

### 2019-05-31 bp_studyMethod.tex
merging the paper folder and the project folder

%!TEX root = bioPrediction_main.tex
\section{Field Study}
We conducted an eight-week field study with 14 participants to investigate the feasibility of predicting stress, focus, and awakeness based on biometric signals. On each workday for eight consecutive weeks, each of the 14 participants wore a biometric sensor while performing their normal work tasks and completed two electronic surveys to report their current self-perceived levels of stress, focus, and awakeness. During the eight week study period, a computer activity tracker was installed and active on 10 of the participants' company-issued laptops (note: a few declined for privacy reasons).

The goal of our study is to understand the efficacy of biometric measurements in predicting stress, focus, and awakeness. To that end we posed the following research questions.

\vspace{0.5em}
\noindent\textbf{RQ1}: Can we use biometric measurements to accurately predict stress,
focus, and awakeness?

\vspace{0.5em}
\noindent\textbf{RQ2}: What is the minimum number of samples needed to accurately predict stress, focus, and awakeness?

\vspace{0.5em}
\noindent\textbf{RQ3}: What is the minimum time window needed to accurately predict stress, focus, and awakeness?

\vspace{0.5em}
\noindent\textbf{RQ4}: How do biometrics compare to computer interaction measurements for predicting stress, focus, and awakeness?

\vspace{0.5em}

%We answer RQ1 in Section~\ref{secOverallAccuracy} by analyzing the overall performance of different
%machine learning models and by comparing the best performing models to baseline approaches.
%To answer RQ2, in Section~\ref{secLearningCurve} we identify the smallest number of training samples necessary to obtain satisfactorily high prediction accuracy and acceptably low mean squared error.
%We answer RQ3 in Section~\ref{secMinimumTW} by considering 16 different time windows for collection of biometric measurements and by identifying the smallest time window on which we can base our training features without sacrificing prediction accuracy.
%Finally, to answer RQ4, in Section~\ref{secCI} we compare the prediction accuracy of machine learning models trained on biometric data versus those trained on computer interaction data.

\subsection{Participants}
We recruited 14 professionals via personal contacts from a large power and automation company. All participants work primarily in an office environment, though half of the participants spend at least 10\% (and up to 50\%) of their time in a laboratory environment. Office workers are a population that generalizes to a variety of contexts, and including part-time laboratory workers guarantees that our participants have varying work patterns that include different levels of computer usage, as well as different levels of activity in both individual and collaborative tasks.

Of the 14 participants, 11 were male, 3 female. The average participant age is ~40, with 5 in the age range 25-34, 7 in the age range 35-44, and 2 in the age range 55+. The participants have an average number of years of professional experience of ~12, with 2 having less than 5 years, 10 having 5-15 years, and 2 having more than 25 years. All participants work for a research organization within the company, but their job functions span line management, laboratory science, scientific research, technology evaluation, and software development.

\begin{comment}
\subsection{Procedure}
\todo{This section does not work here. It is not really the study procedure, but rather the project procedure. Perhaps it can form the basis of an overview section after Field Study but before Analysis and Results.}

%\begin{figure}
%  \centering
%      \includegraphics[width=0.5\textwidth]{Figure1.png}
%  \caption{Process used to gather data and create the machine learning model able to predict the level of stress, focus, and awakeness} 
%  \label{process}
%\end{figure}
%\todo{The figure or text in the boxes could be shortened a lot, maybe even just replaced with pictures of sensor and survey like question with 5point scale. At this point, i'm not sure the figure adds a lot.}

%Figure \ref{process} shows the overall process performed in this study. We started by selecting a population
%of professionals in an office environment.

We use state-of-the-art sensors to capture a stream of biometric data from each individual and a computer interaction tracker to collect computer interaction information during the time
of the study.

The participants were asked to fill a survey 3 times a day: two during the working day, and
one at the end of the day, where they rated their level of stress, focus, and awakeness during
the day; and their level of stress, awakeness, productiveness, and overall feeling at the end of the day.
This was conducted over an eight week period, which is
400\% longer than previous studies~\cite{zuger18,Muller16}.

We then pick a set of time windows (10sec, 20sec, 30sec, 45sec, 1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, 3hour) to analyze. We applied different statistical measurements (mean, standard deviation, variance, median, percentile25, percentile75, interquartile range, maximum, minimum, range) to each of the collected biometric signals and the computer interaction data within the time windows selected. We then 
compared the performance of four different machine learning 
algorithms to be able to predict the survey responses based in the biometric and 
computer interaction data.

Based in the initial corpus of training data, we create a machine learning model that is able to accurately predict the responses for stress, focus, and awakeness for each of the individuals wearing the sensors.
We are therefore able to infer the answer within the spectrum to each of these indicators without the need to ask the participants to provide an answer.
\end{comment}

\subsection{Data Collection}

In this section we describe the two datasets that we collected from each study participant.

\subsubsection{Biometric Sensors}
Figure~\ref{everion} illustrates Biovotion's Everion~\cite{everion}, which we used to track the biometric signals of the study participants. The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements. Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements.

\begin{figure}
  \centering
      \includegraphics[width=0.35\textwidth]{Everion.jpg}
  \caption{Everion sensor used to collect biometric measures.}
   \label{everion}\vspace*{-0.2cm}
\end{figure}

Table~\ref{signals} lists the biometrics measurements that we collected using the Everion. Each measurement is collected once per second, and each recorded observation has an associated timestamp and quality rating. Data collected by the Everion are uploaded to a server, from which we downloaded the data for use in our study.

\begin{table}[t!]
\begin{center}
\begin{tabular}{l l}
\hline
Biometric Measurement & Units of Measure \\ 
\hline
\textbf{Physical Activity} & \cite{fox1999,aldana1996}\\
\hspace{3mm}Intensity of motion & (No unit)  \\
\hspace{3mm}Energy Expenditure & Calories per second (cal/s)  \\
\hspace{3mm}Step counter & Steps  \\
\hline
\textbf{Heart} & \cite{haapalainen2010psycho,healey2005detecting,mulder1992measurement,haag2004emotion} \\
\hspace{3mm}Heart rate & Beats per Minute (bpm) \\
\hspace{3mm}Blood pulse wave & (No Unit)  \\
\hspace{3mm}Heart rate variability & Milliseconds (ms) \\
\hspace{3mm}Blood oxygenation  & Percent (\%)  \\
\hspace{3mm}Blood perfusion & (No unit)  \\
\hline
\textbf{Skin} & \cite{healey2005detecting,haag2004emotion} \\
\hspace{3mm}Galvanic skin response &  kOhm  \\
\hspace{3mm}Skin temperature &  Degrees Celsius ($^{\circ}$C)  \\
\hline
\textbf{Respiration} & \cite{mulder1992measurement,healey2005detecting,haag2004emotion,masaoka1997}\\
\hspace{3mm}Respiratory rate & Breaths per Minute (bpm)  \\
\hline
%\textbf{Environmental}\\
%\hspace{3mm}Barometric pressure & Milibar (mbar)  \\
%\hline
\end{tabular}
\caption{Biometric measurements captured by the Everion, organized by category and with references to previous works using similar data}
\label{signals}
\end{center}
\vspace*{-7mm}
\end{table}

\subsubsection{Surveys}
\label{sec:Surveys}
We sent via text message a survey request to each participant two times per work day. We sent the first request at a random time between 9am and 11am and sent the second request at a random time between 1pm and 3pm. We randomized the request times to avoid either establishing or observing a standard behavioral pattern. That is, we did not want the participants to plan for the arrival of the survey request at a set time, and we did not want the survey request to overlap with a set daily behavior (e.g., coffee break every day at 2:30pm). The same survey was sent each time:
\begin{enumerate}
\item How awake are you right now?
\item How stressed do you feel right now?
\item How focused on work are you right now? 
\end{enumerate}
We used the phrase ``right now'' to capture each aspect in the moment (so as to permit later prediction of each aspect based on biometric data). The wordings of the questions are based on a previous survey of individuals in an organizational context~\cite{Gloor_etal:2010}. The use of awakeness (rather than sleepiness) in Question 1 is inspired by previous work~\cite{Wilhelm_Schoebi:2007} and to some extent also captures the ``arousal'' aspect of the affective space~\cite{Russell:1980}.

%\todo{we are not using the last survey for anything in this paper, so I'd leave it out}
%One last survey was sent at the end of the day, at 4:25pm which asked the 
%four different questions
%detailed below:
%\begin{itemize}
%\item How awake have you been today?
%\item How stressed did you feel today?
%\item How productive have you been today?
%\item How do you feel about your work day?
%\end{itemize}

Following guidelines from similar previous studies~\cite{fogarty05,tanaka11}, we asked the participants to respond to each question using a 5-point Likert scale ranging from 1 (not at all) to 5 (extremely) awake/stressed/focused. Each participant response, as stored by Survey Gizmo, comprised the date, the time at which the response was initiated, the time at which the survey was submitted, the unique identifier for the participant, and the responses submitted by the participant.

### 2019-05-31 bp_threats.tex
merging the paper folder and the project folder

%!TEX root = bioPrediction_main.tex
\section{Threats to Validity}
There are numerous threats to validity to our study.

\textbf{External Validity:}
It is a threat that the results of this study
will not generalize to a broader population of office workers.
To address this concern, we collected participants
from a wide variety of different departments
with different age ranges, genders, work experience, and 
working in different positions.

Another threat is that our results might not generalize
to a different office environment. We have conducted
this study in a typical office environment, common
among technology workers across the world.
These office environments control for a series of
variables to make them standard world wide such
as controlled temperature and lighting.

\textbf{Internal Validity:}
This study tries to find correlations between
biometric features and different productivity-related indicators (stress, focus, and awakeness).
Nonetheless, biometric signals are influenced by far more
variables than the ones this study comprehends.
Therefore, trying to draw a strong causality between the biometric
features and the productivity-related aspects would be inaccurate.
To address this concern we collected the data
in a rote environment and in a regular manner 
to minimize the number of 
external causes that may affect each participant's
biometric signals.

%It is possible that the amount of data collected
%is not enough
%to draw valid conclusions. To address this threat, 
%we collected a data for an eight week period, which is
%400\% longer than previous studies~\cite{zuger18,Muller16}.

\textbf{Construct Validity:}
A threat to the study is that
there are other factors that might either influence the
human aspects of interest or that were considered but
are unrelated biometric signals.
To mitigate this threat we used an state-of-the-art
device that captures a large number of highly accurate biometric
signals. We collected the most commonly analyzed
biometrics that historically have shown correlation with 
productivity aspects from each user.
We also perform a grid analysis in our machine learning model
to pick the number of features that is more predictive
of the productivity-related indicators, 
therefore removing the biometric
features that may be unrelated to the indicators.

### 2019-07-08 bp_conclusion.tex
adapting to the 2020 template

@@
 Based on twice collected daily survey responses and continually collected biometrics data we were able to create a model that was able to predict awakeness, stress, and focus, outperforming the baseline by as much as 84.6\% in the case of awakeness.
-The analysis further shows that there is a continuous increase in performance with the number of training instances, suggesting that there is more potential to improve the model's performance with further data points.
+The analysis further shows that there is a continuous increase in the model performance with the number of training instances, suggesting that there is a lot more potential to improve the model's performance with further data points.
 Interestingly, while aspects such as stress have previously been predominantly linked to biometrics, our comparison with a model build from computer interaction features shows that even these types of features might bear big potential in predicting some of these productivity-related human aspects in the workplace.

### 2019-07-08 bp_data.tex
adapting to the 2020 template

@@
 \item The Everion not recording data in the hour before the participant began the survey (e.g., due to low battery)
-\item Data not being uploaded successfully to the server
+\item Biometric data not being uploaded successfully to the server
 %\item Compatibility issues between the sensor and the OS tracking the participant's data
@@
   \centering
-      \includegraphics[width=0.45\textwidth]{DuringTheDay.png}
-  \caption{Availability of biometric data per participant. Green: survey responses with biometric
-  data; red: responses with no biometric data.}
+      \includegraphics[width=0.5\textwidth]{DuringTheDay.png}
+  \caption{The figure shows the biometric data available per each participant. Green sections represent survey responses with biometric
+  data available, red sections responses with no biometric data available}
    \label{surveyBio}
-   \vspace*{-6mm}
+   \vspace*{-4mm}
 \end{figure}
@@
       \includegraphics[width=0.5\textwidth]{distributiontable.pdf}
-  \caption{Distribution of responses per participant over the 5-point Likert scale (1: not at all; 5: extremely) for each question. }
+  \caption{The distribution of the responses of each participant to the three questions asked during the day are shown.
+Each bar in the histograms represent one of the 5 values on the 5-point Likert scale we asked participants to respond with, where the far left side of the histograms are 1/Not at all, and the far right sides are 5/Extremely}
    \label{responseDistribution}
-   \vspace*{-6mm}
+   \vspace*{-9mm}
 \end{table}

### 2019-07-08 bp_discussion.tex
adapting to the 2020 template

@@
 
-%\vspace{0.05in}
-%\noindent
-%\textbf{Corpus Size:}
-%In this study we collected data from 14 professionals over the course of eight weeks.
+\vspace{0.05in}
+\noindent
+\textbf{Corpus Size:}
+In this study we collected data from 14 professionals over the course of eight weeks.
 By scholarly standards this constitutes a large dataset --- especially given that the participants were not students.
@@
 This upwards trend likely continues, but a larger dataset is needed to confirm or reject this hypothesis.
-%
-%
-%\vspace{0.05in}
-%\noindent
-%\textbf{Per-User Model vs. Across-Users Model:}
-%Our results show that individually trained models outperform on average a more general model.
+
+% \vspace{0.05in}
+% \noindent
+% \textbf{Data Collection:} 
+% One advantage of our study is that it was conducted on professionals in a real office setting (i.e., not on students).
+However, the disadvantage to working with professionals is that business deadlines and other pressing real world issues lead to missing survey responses.
+Additionally, busy professionals often fall into the trap of always selecting the same value for a given question that they have answered many times, which reduces the accuracy of the collected data.
+To minimize the effect of this inertia, we advise collecting data at typical low-productivity times, such as in the early afternoon or on Mondays~\cite{mark2014bored}.
+
+\vspace{0.05in}
+\noindent
+\textbf{Per-User Model vs. Across-Users Model:}
+Our results show that individually trained models outperform on average a more general model.
 Yet, given the increase in model performance with training data set size, we expect that with enough data, a generic model will perform reasonably close to the individually trained models.
@@
 Our follow up work will focus on gathering more data toward this end.
-%
-%
-%% \vspace{0.05in}
-%% \noindent
-%% \textbf{Per-User Model vs. Across-Users Model:}\
-%% While we do see variations across participants, and thus our work shows that personal 
-%% models perform better than generic models, we expect that with enough data, a generic 
-%% model may perform equally well.
+
+% \vspace{0.05in}
+% \noindent
+% \textbf{Per-User Model vs. Across-Users Model:}\
+% While we do see variations across participants, and thus our work shows that personal 
+% models perform better than generic models, we expect that with enough data, a generic 
+% model may perform equally well.
 This would have significant practical implications, eliminating the need for survey-based feedback (which is only necessary for model training), thereby completely automating measurement of human aspects.
 Our follow up work will focus on gathering more data toward this end and on evaluating the performance of a generic model.
-%
-%% \vspace{0.05in}
-%% \noindent
-%% \textbf{Data Collection:} 
-%% One advantage of our study is that it was conducted on professionals in a real office setting (i.e., not on students).
-However, the disadvantage to working with professionals is that business deadlines and other pressing real world issues lead to missing survey responses.
-Additionally, busy professionals often fall into the trap of always selecting the same value for a given question that they have answered many times, which reduces the accuracy of the collected data.
-To minimize the effect of this inertia, we advise collecting data at typical low-productivity times, such as in the early afternoon or on Mondays~\cite{mark2014bored}.

### 2019-07-08 bp_results_1.tex
adapting to the 2020 template

--- 2019-05-31 BiometricsPaper/bp_results_1.tex
@@
   \includegraphics[width=1.0\textwidth]{rq1performance.pdf}
-  \caption{Prediction results individual models.
-The baseline row states averaged results of our baseline classifier, the general row states the averaged results of model trained on all participants.
-Distribution presents the distribution of the binarized outcome variable.}\label{tab:accuracy}%\todo{revise caption}}
-%  The distribution columns show a bar chart of the response distribution (negative/positive) for each of the three variables.
+  \caption{Results of predictions using the individual models.
+The distribution columns show a bar chart of the response distribution (negative/positive) for each of the three variables.
 The baseline row represents the averaged results of our baesline classifier.
-The general row shows the averaged results of our models trained on all participants.
+The general row shows the averaged results of our models trained on all participants.}\label{tab:accuracy}%\todo{revise caption}}
   \vspace*{-4mm}
@@
 % for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)
-\begin{table}[t!]
+\begin{table}[h]
 	\begin{centering}
@@
     \end{tabular}
-    \caption{Hyperparameters selected by grid search analysis for tuning our random forest models. K refers to the number of features selected.}\label{tab:hyperparams}
-    \vspace*{-1cm}
+    \caption{The hyperparameters selected by grid search analysis to tune our random forest models. K refers to the number of features selected.}    \label{tab:hyperparams}
     \end{centering}
@@
 
-\begin{table}[t!]
+\begin{table}[h!]
   \begin{centering}
@@
   \end{tabular}
-  \caption{Averaged Gini importance of each feature category per response variable.}
+  \caption{The averaged Gini importance of each feature category, per response variable.}
   \label{tab:featureImportance}
   \end{centering}
-  \vspace*{-7mm}
+  \vspace*{-5mm}
 \end{table}

### 2019-07-08 bp_results_2.tex
adapting to the 2020 template

--- 2019-05-31 BiometricsPaper/bp_results_2.tex
@@
 In general, the less biometric data is needed to accurately predict a certain outcome measure, the easier and faster the analysis and data collection.
-To examine the optimal and minimum time window for the prediction of stress, focus, and awakeness, we used 16 different time windows from 10 seconds to 3 hours.% as depicted in Figure~\ref{timeWindows}.
-For our analysis, we then trained individual classifiers for each of the 16 time windows, using only features that had a time window smaller or equal to the time window rather than using all combinations of \small{$\{Biometric Measures\} X \{Statistical Metrics\} X \{Time Windows\}$}.
+To examine the optimal and minimum time window for the prediction of stress, focus, and awakeness, we used 16 different time windows from 10 seconds to 3 hours as depicted in Figure~\ref{timeWindows}.
+For our analysis, we then trained individual classifiers for each of the 16 time windows, using only features that had a time window smaller or equal to the time window rather than using all combinations of $\{Biometric Measures\} X \{Statistical Metrics\} X \{Time Windows\}$.
 We again used random forest and a leave-one-out cross validation to train individual classifiers.
@@
 
-%\begin{figure}
-%  \centering
-%      \includegraphics[width=0.4\textwidth]{timeWindows.png}
-%  \caption{Time windows of biometric data collected prior to each survey response: 10sec, 20sec, 30sec, 45sec, 1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, and 3hour.}
-%   \label{timeWindows}
-%\end{figure}
+\begin{figure}
+  \centering
+      \includegraphics[width=0.4\textwidth]{timeWindows.png}
+  \caption{Time windows of biometric data collected prior to each survey response: 10sec, 20sec, 30sec, 45sec, 1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, and 3hour.}
+   \label{timeWindows}
+\end{figure}
 
@@
 \begin{center}
-\begin{tabularx}{\columnwidth}{l l}
+\begin{tabular}{l l}
 \hline
+
 Feature collected by tool & Description \\ 
@@
 \hline
-\end{tabularx}
-\caption{List of computer interaction features.}%~\cite{meyer18}}
+\end{tabular}
+\caption{Features collected per user by the computer interaction tracker}%~\cite{meyer18}}
 \label{tracker}
 \end{center}
-\vspace*{-7mm}
+\vspace*{-4mm}
 \end{table}
@@
 \begin{center}
-\begin{tabularx}{\columnwidth}{llllll}
+\begin{tabular}{llllll}
 \hline
@@
 \hspace{3mm}Biometrics only & 0.808 & 0.269 & 0.314 & 0.289\\
-\hspace{3mm}CI  & 0.758 & 0.425 & 0.362 & 0.391\\
-\hspace{3mm}Biometrics + CI & 0.791 & 0.390 & 0.404 & 0.400\\
+\hspace{3mm}C.I. only & 0.758 & 0.425 & 0.362 & 0.391\\
+\hspace{3mm}Biometrics + C.I. & 0.791 & 0.390 & 0.404 & 0.400\\
 \hline
@@
 \hspace{3mm}Biometrics only & 0.775 & 0.270 &	0.260 & 0.265\\
-\hspace{3mm}CI  & 0.698 & 0.290 & 0.272 & 0.281\\
-\hspace{3mm}Biometrics + CI & 0.712 & 0.317 & 0.286 & 0.301\\
+\hspace{3mm}C.I. only & 0.698 & 0.290 & 0.272 & 0.281\\
+\hspace{3mm}Biometrics + C.I. & 0.712 & 0.317 & 0.286 & 0.301\\
 \hline
@@
 \hspace{3mm}Biometrics only & 0.716 & 0.251 & 0.256 & 0.253\\
-\hspace{3mm}CI  & 0.742 & 0.332 & 0.342 & 0.337\\
-\hspace{3mm}Biometrics + CI & 0.745 & 0.340 & 0.316 & 0.328\\
+\hspace{3mm}C.I. only & 0.742 & 0.332 & 0.342 & 0.337\\
+\hspace{3mm}Biometrics + C.I. & 0.745 & 0.340 & 0.316 & 0.328\\
 \hline
-\end{tabularx}
-\caption{Comparison of prediction performance for the 3 different features sets for 10 participants.
-CI stands for computer interaction.
-Precision/recall refer to prediction of more important classes, i.e.
-stressed, not awake, not focused.}
-%Comparison of the performance of predicting stress, focus and awakeness using the 3 different feature sets for the 10 participants.
+\end{tabular}
+\caption{Comparison of the performance of predicting stress, focus and awakeness using the 3 different feature sets for the 10 participants.
 The performance is calculated as the average of the performance of the individual classifiers.
@@
 \end{center}
-\vspace*{-7mm}
+\vspace*{-6mm}
 \end{table}
@@
 \begin{center}
-\begin{tabularx}{\columnwidth}{lc}
+\begin{tabular}{lc}
 \hline
@@
 \textbf{Stress}\\
-\hspace{3mm}CI & 400\\
-\hspace{3mm}Biometrics + CI & 800\\
+\hspace{3mm}C.I. & 400\\
+\hspace{3mm}Biometrics + C.I. & 800\\
 \hline
 \textbf{Focus}\\
-\hspace{3mm}CI & 20\\
-\hspace{3mm}Biometrics + CI & 300\\
+\hspace{3mm}C.I. & 20\\
+\hspace{3mm}Biometrics + C.I. & 300\\
 \hline
 \textbf{Awakeness}\\
-\hspace{3mm}CI & All\\
-\hspace{3mm}Biometrics + CI & 50\\
+\hspace{3mm}C.I. & All\\
+\hspace{3mm}Biometrics + C.I. & 50\\
 \hline
-\end{tabularx}
-\caption{Optimal number of features used for each of the three models. CI: computer interaction}
-%The optimal number of features we found to select for each of the model/feature set combinations. Computer interactions is abbreviated as C.I.}
+
+\end{tabular}
+\caption{The optimal number of features we found to select for each of the model/feature set combinations. Computer interactions is abbreviated as C.I.}
 \label{ciFeatureSelection}

### 2019-07-08 bp_studyMethod.tex
adapting to the 2020 template

@@
 \vspace{0.5em}
-\noindent\textbf{RQ4}: How do biometrics compare to computer interaction measurements for predicting stress, focus, and awakeness?
+\noindent\textbf{RQ4}: How do biometric measurements compare to computer interaction measurements for predicting stress, focus, and awakeness?
 
@@
 
-%We answer RQ1 in Section~\ref{secOverallAccuracy} by analyzing the overall performance of different
-%machine learning models and by comparing the best performing models to baseline approaches.
-%To answer RQ2, in Section~\ref{secLearningCurve} we identify the smallest number of training samples necessary to obtain satisfactorily high prediction accuracy and acceptably low mean squared error.
-%We answer RQ3 in Section~\ref{secMinimumTW} by considering 16 different time windows for collection of biometric measurements and by identifying the smallest time window on which we can base our training features without sacrificing prediction accuracy.
-%Finally, to answer RQ4, in Section~\ref{secCI} we compare the prediction accuracy of machine learning models trained on biometric data versus those trained on computer interaction data.
+We answer RQ1 in Section~\ref{secOverallAccuracy} by analyzing the overall performance of different
+machine learning models and by comparing the best performing models to baseline approaches.
+To answer RQ2, in Section~\ref{secLearningCurve} we identify the smallest number of training samples necessary to obtain satisfactorily high prediction accuracy and acceptably low mean squared error.
+We answer RQ3 in Section~\ref{secMinimumTW} by considering 16 different time windows for collection of biometric measurements and by identifying the smallest time window on which we can base our training features without sacrificing prediction accuracy.
+Finally, to answer RQ4, in Section~\ref{secCI} we compare the prediction accuracy of machine learning models trained on biometric data versus those trained on computer interaction data.
 
@@
 
-Of the 14 participants, 11 were male, 3 female.
+Of the 14 participants, 11 are male and 3 are female.
 The average participant age is ~40, with 5 in the age range 25-34, 7 in the age range 35-44, and 2 in the age range 55+.
@@
 \subsubsection{Biometric Sensors}
-Figure~\ref{everion} illustrates Biovotion's Everion~\cite{everion}, which we used to track the biometric signals of the study participants.
-The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.
-Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements.
+Figure~\ref{everion} illustrates Biovotion's Everion, which we used to track the biometric signals of the study participants.
+The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/categories/201633909-Everion-Device} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements.
 
@@
   \centering
-      \includegraphics[width=0.35\textwidth]{Everion.jpg}
-  \caption{Everion sensor used to collect biometric measures.}
-   \label{everion}\vspace*{-0.2cm}
+      \includegraphics[width=0.4\textwidth]{Everion.jpg}
+  \caption{We used Biovotion's Everion to collect biometric measurements from the participants}
+   \label{everion}
 \end{figure}
@@
 
-\begin{table}[t!]
+\begin{table}[h!]
 \begin{center}
@@
 \end{center}
-\vspace*{-7mm}
+\vspace*{-4mm}
 \end{table}
@@
 
-Following guidelines from similar previous studies~\cite{fogarty05,tanaka11}, we asked the participants to respond to each question using a 5-point Likert scale ranging from 1 (not at all) to 5 (extremely) awake/stressed/focused.
+Following guidelines from similar previous studies~\cite{fogarty05,tanaka11}, we asked the participants to respond to each question using a 5-point Likert scale ranging from 1 (not at all awake/stressed/focused) to 5 (extremely awake/stressed/focused).
 Each participant response, as stored by Survey Gizmo, comprised the date, the time at which the response was initiated, the time at which the survey was submitted, the unique identifier for the participant, and the responses submitted by the participant.

### 2019-07-08 bp_threats.tex
adapting to the 2020 template

@@
 
-Another threat is that our results might not generalize
+Another threat is that our results won't generalize
 to a different office environment. We have conducted
@@
 
-%It is possible that the amount of data collected
-%is not enough
-%to draw valid conclusions. To address this threat, 
-%we collected a data for an eight week period, which is
-%400\% longer than previous studies~\cite{zuger18,Muller16}.
+It is possible that the amount of data collected
+is not enough
+to draw valid conclusions. To address this threat, 
+we collected a data for an eight week period, which is
+400\% longer than previous studies~\cite{zuger18,Muller16}.
 

### 2019-07-08 bp_studyMethod.tex
addressed concerns from Reviewer 1 using information from their review and our rebuttal

@@
 \hspace{3mm}Blood pulse wave & (No Unit)  \\
-\hspace{3mm}Heart rate variability & Milliseconds (ms) \\
+\hspace{3mm}Heart rate variability (RMSSD) & Milliseconds (ms) \\
 \hspace{3mm}Blood oxygenation  & Percent (\%)  \\
@@
 \label{sec:Surveys}
-We sent via text message a survey request to each participant two times per work day.
-We sent the first request at a random time between 9am and 11am and sent the second request at a random time between 1pm and 3pm.
-We randomized the request times to avoid either establishing or observing a standard behavioral pattern.
-That is, we did not want the participants to plan for the arrival of the survey request at a set time, and we did not want the survey request to overlap with a set daily behavior (e.g., coffee break every day at 2:30pm).
+Following guidelines from previous studies~\cite{Lalle16,Panwar18,Luo18} and 
+following the preferrences of extensive user piloting, we sent via 
+text message a survey request to each participant two times per 
+work day. Pilot participants preferred these over other means, in part, due 
+to them being accessible and noticeable anywhere in the office.
+
+We sent the 
+first request at a random time between 9am and 11am 
+and sent the second request at a random time between 1pm and 3pm. We 
+randomized the request times to avoid either establishing or observing a 
+standard behavioral pattern. That is, we did not want the participants to 
+plan for the arrival of the survey request at a set time, and we did not 
+want the survey request to overlap with a set daily behavior (e.g., coffee 
+break every day at 2:30pm). Similarly, we avoided using tools which allow 
+for too much freedom in response time~\cite{Adams18} since this would 
+discurage participation in stressed timeframes and would bias the 
+corpus. 
+%This sentence was added in response to Reviewer 1 comment:
+%There is already a lot of work on designing a new self-log, a self-report 
+%tool which can blend into the workstation [1] or offers users more freedom 
+%in time to self-report [2], or facilitate the self-log by using wearables.
 The same survey was sent each time:

### 2019-07-09 bp_data.tex
Addressed most of the concerns of reviewer 2

@@
 Figure~\ref{surveyBio} illustrates the number of survey responses with associated biometric data for each study participant. 
-Participant S2 and S12 have particularly low numbers of usable survey responses.
-In each of these cases, the issue related to biometric data not being uploaded successfully to the server.
+%This was added to address REviewer 2's concern: Is it correct that the maximum number of responses should have been 112 on the surveys?
+The total number of responses per participant is affected by their response rate and by the number days out-of-office (e.g., vacations, holidays, etc.). 
+Participant S2 and S12 have particularly low numbers of usable survey responses. In each of these cases, the issue related to biometric data not being uploaded successfully to the server.
 
@@
 \subsubsection{Response Transformations}
-Table~\ref{responseDistribution} illustrates the distribution of responses from each participant for each of the three survey questions (which are listed in Section~\ref{sec:Surveys}).
+Table~\ref{responseDistribution} illustrates the distribution of responses from each participant for each of the three survey questions (listed in Section~\ref{sec:Surveys}).
 The figure shows that there is a notable imbalance in the distribution of the self-reported responses provided by the participants.
@@
    \label{responseDistribution}
-   \vspace*{-9mm}
+   \vspace*{-2mm}
 \end{table}

### 2019-07-09 bp_intro_v3.tex
Addressed most of the concerns of reviewer 2

--- 2019-05-31 BiometricsPaper/bp_intro_v3.tex
@@
 Fortunately, with recent advances in sensing technologies there is increasing hope that we can accurately and automatically measure such human aspects.
-A number of studies have already discussed and investigated the use of biometrics to measure aspects such as a person's cognitive and emotional states~\cite{beatty82,Kramer90,Rowe98,richter1998psychophysiological,Wilson2002}.
-Studies have also looked more specifically at using biometrics to measure stress~\cite{dishman2000stress,Setz2010}, or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004emotion}.
-Yet, very few studies measure productivity-related factors in the workplace over an extended period of time, and little is known about measuring the more abstract concept of focus.
-Given the importance of focus, described as a combination of engagement and challenge in a work task by Mark et al.~\cite{mark2014bored}, to an office workers' productivity, this factor also deserves further study.
+A number of studies have already discussed and investigated the use of biometrics to measure aspects such as a person's cognitive state
+%This change is regarding Reviewer 2's concern: what are the "aspects"?
+(e.g., processing load)~\cite{beatty82,Kramer90,richter1998psychophysiological,Rowe98,Wilson2002} 
+and more specifically to measure stress~\cite{dishman2000stress,Setz2010}, 
+or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004emotion}. Yet, very few 
+studies measure productivity-related factors in the workplace over an 
+extended period of time, and little is known about measuring the more 
+abstract concept of focus (directed attention). 
+% Added regarding concern of Reviewer 2: Define what is meant by focus when it's first introduced. 
+Given the importance of focus, described as a 
+combination of engagement and challenge in a work task by Mark et al.~\cite{mark2014bored}, to an office workers' productivity, this factor also 
+deserves further study.
 

### 2019-07-09 bp_relatedWork.tex
addressed concerns from Reviewer 1 using information from their review and our rebuttal
Addressed most of the concerns of reviewer 2

--- 2019-05-31 BiometricsPaper/bp_relatedWork.tex
@@
 
-Previous studies have sought to measure, detect, and predict productivity indicators.
+Previous approaches~\cite{zuger18,Lalle16,Panwar18} have shown the feasibility of applying machine learning 
+to biometric data to measure variables such as stress, interruptibility, 
+negative emotion, or confusion during computer 
+interaction. These approaches vary in the 
+variables examined, the invasiveness of the sensors, and the length and 
+context of the studies. Many previous studies focused on 
+shorter time frames (mostly hours, but in one case also up to five 
+days~\cite{sano2013stress} 
+or two weeks~\cite{zuger18}). To the best of the authors' knowledge, no 
+other study has examined such an extensive period (eight weeks) in a real-
+world office environment and the inspected variables (stress, awakeness, and focus).
 

### 2019-07-09 bp_results_1.tex
Addressed most of the concerns of reviewer 2

@@
 	\begin{centering}
+	\small\addtolength{\tabcolsep}{-1pt}
     \begin{tabular}{llll}
@@
 While some participants showed a large improvement, for others the baseline performed much better than the individually trained classifier.
-For instance, for predicting 'stressed', the individual classifiers improved upon the baseline for S4, S6, S8, S11, S12, and S14 with a maximum improvement of 88.4\% in precision and 184.6\% in recall for S12, while they did worse for S1, S3, S5, S7, S9, S10, and S13, and in the worst cases did not correctly predict a single instance of 'stressed'.\\[-0.1cm]
+For instance, for predicting 'stressed', the individual classifiers improved upon the baseline for S4, S6, S8, S11, S12, and S14 with a maximum improvement of 88.4\% in precision and 184.6\% in recall for S12, while they did worse for S1, S3, S5, S7, S9, S10, and S13, and in the worst cases did not correctly predict a single instance of 'stressed'.
+Typically users that have the lowest precision and recall values are those where the data is the most unbalanced.
+This translates to a scenario where the classifier will likely label the minority instances as part of the majority class due to the unbalanced ratio of the data.\\[-0.1cm]
 % We attribute these discrepancies to the large differences in response distributions between participants, as well as to the subjectivity of self-reporting.
@@
   \end{centering}
-  \vspace*{-5mm}
+  \vspace*{-2mm}
 \end{table}

### 2019-07-09 bp_results_2.tex
Addressed most of the concerns of reviewer 2

@@
 \subsection{Minimum Number of Training Samples}\label{secLearningCurve}
-Collecting experience samples from users is expensive, since participants are being interrupted several times a day and have to answer the questions.
-To minimize the number of samples to be collected from participants, we examined how the performance of individual classifiers changes over the number of samples used to train the classifier.
+Collecting experience samples from users is expensive, since participants 
+are being interrupted several times a day and have to answer the survey. 
+To minimize the number of samples to be collected from participants, we 
+examined how the performance of individual classifiers changes over the 
+number of samples used to train the classifier.
 
-Participants in our study had varying levels of responsiveness to the experience sampling ranging from 10 to 76 (see Figure~\ref{responseDistribution}).
-To maintain a certain generalizability while also being able to examine a range of sample numbers for training the classifier, we decided to include the ten participants with the most samples for this analysis.
-As a result, we had at least 34 survey responses per participant to examine the learning curve of the classifiers.
-For our analysis, we thus performed a leave-one-out cross validation with random sample sets of size 1 to 33 and calculated the average through all folds of the validation.
+Participants in our study had varying levels of responsiveness to the 
+experience sampling ranging from 10 to 76 (see Figure~
+\ref{responseDistribution}).
+%Answering Reviewer 2's concern: Why only 10 participants? What was the rationale for the cut-off selected? 
+To answer this research question we analyze the maximum number of responses \textbf{all} analyzed participants have to train the model.
+Since some participants have a very small number of maximum answers (ten being the lowest) we excluded the four participants with the least number of responses for this analysis.
+%To maintain a certain generalizability while 
+%also being able to examine a range of sample numbers for training the 
+%classifier, we removed the four participants with the least number of 
+%samples for this analysis. 
+As a result, we obtained a corpus of analysis where all participants have a 
+higher number of responses (34 being the lowest), which allows us to examine 
+the learning curve of the classifiers for a larger number of training data points. For our 
+analysis, we thus performed a leave-one-out cross validation with random 
+sample sets of size 1 to 33 and calculated the average through all folds of 
+the validation.
 
-For each of the three productivity-related aspects, we are again more interested in predicting when a worker is stressed, not awake, or not focused, the less common class in all three cases.
-Since the less common class can be very small, we weighted each participants' classifier performance by the percentage of the samples in this smaller class.
+For each of the three productivity-related aspects, we are again more 
+interested in predicting when a worker is stressed, not awake, or not 
+focused, the less common class in all three cases. Since the less common 
+class can be very small, we weighted each participants' classifier 
+performance by the percentage of the samples in this smaller class.
 
@@
 
-Figure~\ref{timeWindowsPandR} shows how the F-score changes for predicting `awake' over the 16 different time windows.
-The figure shows an increasing trend in the F-score, i.e.
-the higher the number of included time windows included, the higher the F-score.
-However, there is one exception, the time window of 1200 seconds that achieves a performance close to the one for the time window 10,800 seconds (3 hours), at which point all features are included.
-Overall, our results thus show that while using all time windows up to 3 hours performs best, and outperforms the feature set that is solely based on a 10 second time window by 28\% (from 0.18 to 0.24), the performance for a time window of 1200 seconds is already very close to optimal (everything up to 3 hours in our case).
+Figure~\ref{timeWindowsPandR} shows how the F-score changes for predicting 
+`awake' over the 16 different time windows. The figure shows an increasing 
+trend in the F-score, i.e. the higher the number of included time windows, 
+the higher the F-score. However, there is one exception, the time window of 
+1200 seconds that achieves a performance close to the one for the time 
+window 10,800 seconds (3 hours), at which point all features are included. 
+Overall, our results thus show that while using all time windows up to 3 
+hours performs best, and outperforms the feature set that is solely based on 
+a 10 second time window by 28\% (from 0.18 to 0.24), the performance for a 
+time window of 1200 seconds is a good trade-off for selecting a shorter time 
+window while maintaining high performance.
 
@@
 \begin{center}
+\small\addtolength{\tabcolsep}{-1pt}
 \begin{tabular}{l l}
@@
 \end{center}
-\vspace*{-4mm}
+\vspace*{-1mm}
 \end{table}
@@
 Table~\ref{ciPerformance} lists the results of our analysis.
-The results show that in all cases, the computer interaction based model was able to improve upon the biometric model in terms of precision and recall, but not in accuracy.
+The results show that in all cases, the computer interaction based model was able to improve upon the biometric model in terms of precision and recall.
+%MS is removing the accuracy results because it is not what we care about and it is just making the results harder to understand
+%, but not in accuracy. 
 Further, we found that the combined model was the most effective model in terms of precision and recall for predicting stress and awakeness overall, but performed slightly worse than the model using only computer interaction features for focus.
@@
 \begin{center}
+%\small\addtolength{\tabcolsep}{-1pt}
 \begin{tabular}{llllll}
 \hline
-Model/Feature Set & Accuracy & Precision & Recall & F-Score\\
+Model/Feature Set & Precision & Recall & F-Score \\ %& Accuracy\\
 \hline
 \textbf{Awakeness}\\
-\hspace{3mm}Biometrics only & 0.808 & 0.269 & 0.314 & 0.289\\
-\hspace{3mm}C.I. only & 0.758 & 0.425 & 0.362 & 0.391\\
-\hspace{3mm}Biometrics + C.I. & 0.791 & 0.390 & 0.404 & 0.400\\
+\hspace{3mm}Biometrics only  & 0.269 & 0.314 & 0.289 \\ %& 0.808\\
+\hspace{3mm}C.I. only  & 0.425 & 0.362 & 0.391 \\ % & 0.758\\
+\hspace{3mm}Biometrics + C.I. & 0.390 & 0.404 & 0.400 \\ % & 0.791 \\
 \hline
 \textbf{Stress}\\
-\hspace{3mm}Biometrics only & 0.775 & 0.270 &	0.260 & 0.265\\
-\hspace{3mm}C.I. only & 0.698 & 0.290 & 0.272 & 0.281\\
-\hspace{3mm}Biometrics + C.I. & 0.712 & 0.317 & 0.286 & 0.301\\
+\hspace{3mm}Biometrics only  & 0.270 &	0.260 & 0.265 \\ % & 0.775\\
+\hspace{3mm}C.I. only & 0.290 & 0.272 & 0.281 \\ % & 0.698 \\
+\hspace{3mm}Biometrics + C.I. & 0.317 & 0.286 & 0.301 \\ % & 0.712\\
 \hline
 \textbf{Focus}\\
-\hspace{3mm}Biometrics only & 0.716 & 0.251 & 0.256 & 0.253\\
-\hspace{3mm}C.I. only & 0.742 & 0.332 & 0.342 & 0.337\\
-\hspace{3mm}Biometrics + C.I. & 0.745 & 0.340 & 0.316 & 0.328\\
+\hspace{3mm}Biometrics only & 0.251 & 0.256 & 0.253 \\ % & 0.716\\
+\hspace{3mm}C.I. only & 0.332 & 0.342 & 0.337 \\ % & 0.742\\
+\hspace{3mm}Biometrics + C.I. & 0.340 & 0.316 & 0.328 \\ % & 0.745\\
 \hline
@@
 \end{center}
-\vspace*{-6mm}
+\vspace*{-3mm}
 \end{table}
@@
 \hline
-
 \end{tabular}

### 2019-07-09 bp_studyMethod.tex
Addressed most of the concerns of reviewer 2

@@
 On each workday for eight consecutive weeks, each of the 14 participants wore a biometric sensor while performing their normal work tasks and completed two electronic surveys to report their current self-perceived levels of stress, focus, and awakeness.
-During the eight week study period, a computer activity tracker was installed and active on 10 of the participants' company-issued laptops (note: a few declined for privacy reasons).
+During the eight week study period, a computer activity tracker was installed and active on 10 of the participants' company-issued laptops (note: four participants from the original sample declined for privacy reasons).
 
@@
 Figure~\ref{everion} illustrates Biovotion's Everion, which we used to track the biometric signals of the study participants.
-The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/categories/201633909-Everion-Device} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements.
+The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/categories/201633909-Everion-Device} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements for shorter periods of time or capturing a smaller number of measurements.
 
@@
 \begin{center}
+%\small\addtolength{\tabcolsep}{-2pt}
 \begin{tabular}{l l}

### 2019-07-10 bp_results_1.tex
addressed concerns of Reviewer 3

@@
 As a first step, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11} and performing a grid search analysis to determine the optimal hyperparameters for each classifier.
-Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, random forest, and a multilayer perceptron neural network.
+Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, and a multilayer perceptron neural network.
 The optimal values for random forest and the three output measures are listed in Table \ref{tab:hyperparams}.

### 2019-07-10 bp_studyMethod.tex
addressed concerns of Reviewer 3

@@
 We conducted an eight-week field study with 14 participants to investigate the feasibility of predicting stress, focus, and awakeness based on biometric signals.
-On each workday for eight consecutive weeks, each of the 14 participants wore a biometric sensor while performing their normal work tasks and completed two electronic surveys to report their current self-perceived levels of stress, focus, and awakeness.
+On each workday for eight consecutive weeks, each of the 14 participants wore a biometric sensor while performing their normal work tasks and completed two electronic surveys to report their current self-perceived levels of stress, focus, and awakeness to establish a ground truth.
 During the eight week study period, a computer activity tracker was installed and active on 10 of the participants' company-issued laptops (note: four participants from the original sample declined for privacy reasons).
@@
 \begin{center}
-%\small\addtolength{\tabcolsep}{-2pt}
+\small\addtolength{\tabcolsep}{-1pt}
 \begin{tabular}{l l}

### 2019-07-10 bp_discussion.tex
addressed concerns of Reviewer 3
modified intro and discussion to frame it more towards HCI

@@
 \section{Discussion}
+%\vspace{0.05in}
+\noindent
+\textbf{Implications for HCI Design and Workspaces:}
+The results discussed in this paper establish great potential
+to increase the well-being of knowledge workers and the quality of their work
+through monitoring their biometric signals and computer interactions with the
+goal of recognizing their stress, awakeness, and focus levels in an uninvasive 
+and automatic manner. Being able to accurately recognize periods of high-
+stress in workers enables companies to prevent or de-escalate potentially 
+dangerous situations in the workplace, e.g. confrontation between co-workers; 
+or it can help to design more team-based approaches such as an 
+awareness dashboard of a teams’ stress level.
+Similarly, awakeness and focus
+can be enhanced given such a technology
+e.g. through individual approaches such as adapting the lighting in a control 
+room, or scheduling breaks to prevent focus loss.
 
-\vspace{0.05in}
-\noindent
-\textbf{Target Workspaces:}
 Currently, the cost of biometric sensors and necessary infrastructure, such as automated light and sound systems for adjusting the environment, makes our approach most appropriate for high-value workspaces, such as control rooms, command centers, or dispatch offices.
@@
 
-\vspace{0.05in}
+%\vspace{0.05in}
 \noindent

### 2019-07-10 bp_intro_v3.tex
modified intro and discussion to frame it more towards HCI

@@
 \section{Introduction}
-\textit{`The most valuable asset of a 21st-century institution (whether business or non-business) will be its knowledge workers and their productivity'}~\cite{drucker1999knowledge}.
-Yet, in today's workplaces, knowledge workers and their time are not necessarily treated as the most valuable asset, and they constantly face challenges, such as a high work fragmentation, continuous disruptions and distractions, highly complex and demanding tasks, and long working hours~\cite{gonzalez2004constant,mark2008cost,czerwinski04diary}.
-These challenges make it difficult for a knowledge worker to stay focused and can impact both the quality of work performed and productivity.
-For instance, the continuous disruptions that knowledge workers face in office work environments can lead to a higher error rate and a lower performance~\cite{bailey2001effects,mark2008cost}.
-In some cases,
-the consequences are even more severe; in control room
-settings, the consequences can be accidents that cost millions of dollars in losses. 
+\textit{`The most valuable asset of a 21st-century institution (whether 
+business or non-business) will be its knowledge workers and their 
+productivity'}~\cite{drucker1999knowledge}. Knowledge workers constantly face 
+challenges, such as a high work fragmentation, continuous disruptions and 
+distractions, highly complex and demanding tasks, and long working 
+hours~\cite{gonzalez2004constant,mark2008cost,czerwinski04diary}. These 
+challenges make it difficult for a knowledge worker to stay focused and
+impact interaction with their digital environment, the quality of work performed and productivity. For instance, the 
+continuous disruptions that knowledge workers face in office work environments 
+can lead to a higher error rate and a lower 
+performance~\cite{bailey2001effects,mark2008cost}. In some cases,
+the consequences are even more severe; e.g., in control room
+settings, the consequences of an erroneous intraction can lead to accidents valued in millions of dollars in losses. 
 %when a sleepy operator might, for example, miss early warning signs and cause...
@@
 %Similarly, stress at the workplace is a growing concern at the workplace and one of the most common work-related health problem that can lead to fatigue, burnout and various other illnesses, resulting in work absence and a high productivity loss~\cite{hockey1997stress,setz2010stress,wrs2010}.
-Less immediate, but just as insidious, stress at the workplace is a growing concern and one of the most common work-related health problems.
-It leads to fatigue, burnout and various other illnesses, ultimately resulting in work absences and marked productivity losses~\cite{hockey1997stress,setz2010stress,wrs2010}.
+Less immediate, but just as insidious, stress at the workplace is a growing 
+concern and one of the most common work-related health problems. It leads to 
+fatigue, burnout and various other illnesses, ultimately resulting in work 
+absences and marked productivity 
+losses~\cite{hockey1997stress,setz2010stress,wrs2010}.
 
-Being able to continuously measure human aspects --- such as focus, awakeness, and stress --- in the workplace bears huge potential for better supporting knowledge workers, combatting workplace challenges, and improving knowledge worker productivity.
-Such measures could be used to provide better management of disruptions~\cite{iqbal2005effectiveness,zuger2017reducing}, to automatically adjust lighting to reduce sleepiness among control room operators, or to continuously monitor stress levels to trigger pro-active interventions and reduce health problems.
+Being able to continuously measure human aspects --- such as focus, awakeness, 
+and stress --- in the workplace bears huge potential for better supporting 
+knowledge workers combatting workplace challenges, and improving productivity 
+and well-being. Such measures could be used to provide better interaction with 
+their environment, such as management 
+of disruptions~\cite{iqbal2005effectiveness,zuger2017reducing}, to 
+automatically adjust lighting and temperature to reduce sleepiness, or to continuously monitor stress levels to trigger pro-active 
+interventions and reduce health problems.
 
-Fortunately, with recent advances in sensing technologies there is increasing hope that we can accurately and automatically measure such human aspects.
-A number of studies have already discussed and investigated the use of biometrics to measure aspects such as a person's cognitive state
+Fortunately, with recent advances in sensing technologies there is increasing 
+hope that we can accurately and automatically measure such human aspects. A 
+number of HCI studies have discussed and investigated the use of biometrics to 
+measure aspects such as a person's cognitive state 
 %This change is regarding Reviewer 2's concern: what are the "aspects"?
-(e.g., processing load)~\cite{beatty82,Kramer90,richter1998psychophysiological,Rowe98,Wilson2002} 
+(e.g., processing load)~\cite{beatty82,Kramer90,richter1998psychophysiological,Rowe98,Wilson2002}, 
 and more specifically to measure stress~\cite{dishman2000stress,Setz2010}, 
-or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004emotion}. Yet, very few 
-studies measure productivity-related factors in the workplace over an 
-extended period of time, and little is known about measuring the more 
-abstract concept of focus (directed attention). 
+or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004emotion}. Similarly,
+studies have examined the occurrence, cause and effect of similar 
+aspects in various contexts~\cite{mark2014bored,mark2008cost}, how these can be measured~\cite{Hernandez14}, and 
+ how they might be modulated/regulated~\cite{Snyder15}. 
+While previous work mostly focused on smaller or controlled lab studies, our study extends this work by performing a field study to examine whether and how these three variables ---stress, focus and awakeness--- can be measured over an extended period of time in a knowledge worker's context,
+%Yet, very few 
+%studies measure productivity-related factors in the workplace over an 
+%extended period of time
+and little is known about measuring the more 
+abstract concept of focus (i.e., directed attention). 
 % Added regarding concern of Reviewer 2: Define what is meant by focus when it's first introduced. 
@@
 
-The objective of our research is to develop continuous (and automatic) measures of focus, awakeness, and stress in the workplace.
-These measures are targeted with an eye towards improving knowledge workers' productivity and well-being in the future.
-For instance, by automatically protecting the worker from audible interruptions during a high focus period by showering him/her with white noise or by recommending stress-reducing interventions during an extended period of high stress.
+The objective of our research is to develop continuous (and automatic) 
+measures of focus, awakeness, and stress in the workplace. These measures are 
+targeted towards improving knowledge workers' productivity and 
+well-being in the future by customizing interaction with their environment; e.g., by automatically protecting the workers
+from audible interruptions during a high focus period by showering them 
+with white noise or by recommending stress-reducing interventions during an 
+extended period of high stress.
 
-Our work builds upon previous work and extends it by performing an eight-week study in the workplace with 14 knowledge workers, collecting biometric and computer interaction data as well as experience samples on focus, awakeness, and stress.
-Our participants, who perform various job functions for a research and development group within a single large corporation, wore a single biometric armband sensor\footnote{Biovotion Everion sensor~\cite{everion}} with low invasiveness that captured heart- and skin-related measures.
-This modality was chosen to ease longitudinal deployment in the field.
-Using machine learning, we created classifiers and analyzed their ability to predict the three productivity-related human aspects.
-In addition, we compared biometric features with computer interaction features in their predictive power, we analyzed how the combination
-of these two training sets compares to the performance of each
+Our work builds upon previous work and extends it by performing an eight-week 
+study in the workplace with 14 knowledge workers, collecting biometric and 
+computer interaction data as well as experience samples on focus, awakeness, 
+and stress. Our participants, who perform various job functions for a research 
+and development group within a single large corporation, wore a single 
+biometric armband sensor\footnote{Biovotion Everion sensor~\cite{everion}} 
+with low invasiveness that captured heart-, respiration- and skin-related measures. This 
+modality was chosen to ease longitudinal deployment in the field. Using 
+machine learning, we created classifiers and analyzed their ability to predict 
+the three productivity-related human aspects. Similarly, we captured computer 
+interaction features from participants' everyday computer usage and
+analyzed how a model trained using the combination
+of these two training sets compares to the performance of models using each
 set individually.
@@
 	\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
-	\item A discussion on the impact of applying this research to possible target workspaces, besides a reflection on aspects that can be further improved in future studies.
+	\item A discussion on the impact of applying this research to improve
+the interaction of knowledge workers with their digital environment leading to an improvement in their productivity and well-being, besides a reflection on aspects that can be further improved in future studies.
     \end{itemize}

### 2019-07-11 bp_discussion.tex
small change in discussion section

@@
 stress in workers enables companies to prevent or de-escalate potentially 
-dangerous situations in the workplace, e.g. confrontation between co-workers; 
-or it can help to design more team-based approaches such as an 
-awareness dashboard of a teams’ stress level.
-Similarly, awakeness and focus
+dangerous situations in the workplace, e.g. confrontation between co-workers. Similarly, this can help to design more team-based approaches such as an 
+awareness dashboard of a team's stress level, and avoid digital interruptions in high-stress or high-focus periods similar to previous studies~\cite{zuger2017reducing}.
+Likewise, awakeness and focus
 can be enhanced given such a technology
-e.g. through individual approaches such as adapting the lighting in a control 
+through individual approaches such as adapting the lighting in a control 
 room, or scheduling breaks to prevent focus loss.

### 2019-09-19 bp_results_1.tex
update paper title & add section on observation results

@@
 For prediction, we used the features extracted from the collected data as the input data, and binarized the participants' self-reported responses on stress, focus, and awakeness into two classes each (e.g., 'stressed' and 'not stressed') to use as output measures.
+
+\subsection{Observed Trends in Knowledge Worker Stress Levels}
+We examined the end of day survey responses collected from each participant to see if any identifiable trends emerged.
+We used data from 13 of the 14 participants - one participant was excluded from this analysis as they experienced atypical stress levels in the latter half of the study due to factors outside of our control.
+
+\subsubsection{Baseline stress levels}
+Common along all participants was a trend to select one stress rating far more frequently than any other.
+We will refer to this value as the participant's baseline stress level.
+All but one participant reported their perceived stress level for the day as their baseline stress level more than 50\% of the time.
+In total, the baseline values made up 65\% of the reported values collected from participants.
+Interestingly, while participants sometimes saw periods of sustained increases in stress, lasting as many as 6 consecutive workdays in the most extreme case, participants would always return to their baseline stress level at some point.
+
+The baseline stress level varied significantly between participants.
+7 participants reported feeling average stress levels most frequently (rating 3 on our scale), while 5 reported feeling little stress (rating 2) and 1 reported feeling no stress at all (1).
+
+\subsubsection{Stressful days tend to come in clusters}
+Accounting for the variance between participants perceived stress baselines, we consider a stressful day to be one that represents a deviation of 1 or more stress levels above the participants baseline.
+Of the 93 stressful days we observed in total, we found that 39 (41\%) of these days occurred in groupings of two or more consecutive stressful workdays.
+The most common size of these groups was two workdays, while the largest group we observed was six workdays.
+The day after a stressful day is much more likely to be a stressful day as compared to any other day. (0.55 average increase over baseline, compared to 0.02 average increase over baseline)
+
+\subsubsection{Extreme changes in stress levels are rare}
+After accounting for each participant's perceived stress baseline, we examined the frequency of deviations from the baseline.
+We found that participants were far more likely to report a stress level that was within 1 point of their baseline, than to report a stress level 2 or more points away.
+These extreme deviations represented only 15\% of all reported values which differed from the participants baseline.
+As well, the majority (78\%) of these deviations came from just two participants.
+This suggests that some people may be less resilient to the stress of the workplace than others - while for most people extremely stressful days are few and far between.
+
+\subsubsection{Stress fluctuations are difficult to explain without invasive measures}
+We created a linear mixed model to attempt to explain some of the stress that our participants were experiencing.
+We experimented with multiple independent variables, including those derived from computer interaction data such as total time spent actively on the computer, percentage of time idle, and amount of time spent on non work-related web browsing.
+We also looked at the day of the week and proximity to beginning or end of month as possible explanatory variables.
+However, ultimately we were unable to find any variable which showed a significant correlation with our participants perceived stress levels.
+This points to the need for additional instrumentation if we are to succesfully understand and predict stress in the workplace.
+
+\subsubsection{Tai Chi Intervention}
+We linked participants tai chi session attendance data with their self-reported percieved stress levels.
+Building a linear mixed model with tai chi attendance as a random effect, we found that tai chi attendance contributed a small amount to decreased stress in the work week immediately following the tai chi session (slope of -0.188, p < 0.05).
+However, we did not find a connection between attendance and stress on the day of the session.
 

### 2019-09-19 bp_studyMethod.tex
modified intro and discussion to frame it more towards HCI
update paper title & add section on observation results

@@
 \noindent\textbf{RQ4}: How do biometric measurements compare to computer interaction measurements for predicting stress, focus, and awakeness?
+
+\vspace{0.5em}
+\noindent\textbf{RQ5}: How do knowledge worker stress levels change over long periods of time?
+
+\vspace{0.5em}
+\noindent\textbf{RQ6}: Can practicing Tai Chi help to reduce stress levels in the office?
 
@@
 \end{center}
-\vspace*{-4mm}
+%\vspace*{-4mm}
 \end{table}
@@
 %\todo{we are not using the last survey for anything in this paper, so I'd leave it out}
-%One last survey was sent at the end of the day, at 4:25pm which asked the 
-%four different questions
-%detailed below:
-%\begin{itemize}
-%\item How awake have you been today?
-%\item How stressed did you feel today?
-%\item How productive have you been today?
-%\item How do you feel about your work day?
-%\end{itemize}
+One last survey was sent at the end of the day, at 4:25pm which asked the 
+four different questions
+detailed below:
+\begin{itemize}
+\item How awake have you been today?
+\item How stressed did you feel today?
+\item How productive have you been today?
+\item How do you feel about your work day?
+\end{itemize}
 
@@
 Each participant response, as stored by Survey Gizmo, comprised the date, the time at which the response was initiated, the time at which the survey was submitted, the unique identifier for the participant, and the responses submitted by the participant.
+
+\subsection{Intervention}
+At the midpoint of our study, we introduced an intervention in the form of Tai Chi classes, with the goal of examining the efficacy of Tai Chi or other mindfulness excercises for reduction of office stress.
+Classes were offered two times a week in the middle of the work day.
+Participants were asked to attend one class a week.
+We recorded each participants attendance, including which of the two sessions they attended.

### 2019-09-19 bp_relatedWork.tex
additions to stress section of related work

@@
 
-Stress in the workplace and its effects on service providers has been analyzed in call centers~\cite{Hernandez11} and in the context of the perceived imbalance between resources and demands~\cite{cherniss80}.
-These studies considered several factors such as personality traits, career-related goals and attitudes, and life outside of work, and these factors were correlated with stress levels and burnout.
+Hovsepian et al.~\cite{Hovsepian15} have worked to obtain a standard for continuous stress assessment.
+They used sensors to conduct a seven-day lab study with 26 participants, as well as a field study with 20 participants.
+They found that their model showed significant improvement over simple heart rate variability measurements.
+However, this model requires the use of a suite of invasive sensors that would be impractical for a study the length of ours.
+
+Hernandez et al.~\cite{Hernandez14} investigated the use of pressure-sensitive keyboard and a capacitative mouse as non-intrusive means for measuring computer users stress levels.
+They found participants exhibited significantly increased typing pressure and mouse contact when in stressful conditions.
+McDuff et al.~\cite{mcduff16} experimented with a camera to measure photoplethysmographic signals indicative of cognitive stress.
+Vizer et al.~\cite{vizer_automated_2009} used keystroke and linguistic features to automatically measure stress levels in response to cognitive and physical stress conditions.
+All of these studies were performed in a controlled lab setting over a short duration and the results have yet to be replicated in the field.
 
@@
 
-Hovsepian et al.~\cite{Hovsepian15} have worked to obtain a standard for continuous stress assessment.
-They used sensors to conduct a seven-day lab study with 26 participants, as well as a field study with 20 participants.
+Stress in the workplace and its effects on service providers has been analyzed in call centers~\cite{Hernandez11} and in the context of the perceived imbalance between resources and demands~\cite{cherniss80}.
+These studies considered several factors such as personality traits, career-related goals and attitudes, and life outside of work, and these factors were correlated with stress levels and burnout.
+
+Kocielnik et al.~\cite{kocielnik_smart_2013} developed a framework for unobtrusive and continuous measurement of stress in real life conditions.
+They equipped university staff members with a wristband sensor and combined this data with information from the participants calendars over the course of four weeks.
+They observed that the data they collected reflected well the participants perceptions of their own stress levels.
+However this work focused mostly on providing retrospective information to users so they can work to improve their own stress balance.
+They did not attempt to make observations about the big picture of participants stress profiles or make predictions in real time.
+
+Our study stands out from these works by nature of its length and focus on knowledge worker stress in everyday office life, using unobtrusive measures.
+To the best of our knowledge there is no longitudinal study which attempts to explain and predict knowledge worker stress that comes close to the length of our own study.
+The eight week duration gives us authority to speak on the nature of day-to-day fluctuations in knowledge worker stress levels.
 

### 2019-09-20 bp_results_1.tex
Restructured section levels and removed detail on classifiers.

--- 2019-07-10 bp_results_1.tex
@@
 
-\section{Analysis and Results}
+%\section{Analysis and Results}
 
@@
 
-To evaluate the efficacy of continuously predicting a knowledge worker's stress, focus, and awakeness in the workplace, we trained and tested machine learning classifiers using a leave-one-out cross validation.
+%% To evaluate the efficacy of continuously predicting a knowledge worker's stress, focus, and awakeness in the workplace, we trained and tested machine learning classifiers using a leave-one-out cross validation.
 For prediction, we used the features extracted from the collected data as the input data, and binarized the participants' self-reported responses on stress, focus, and awakeness into two classes each (e.g., 'stressed' and 'not stressed') to use as output measures.
 
-\subsection{Predicting Stress, Focus \& Awakeness}
+\section{Observed Trends Over Time in Stress Levels}
+
+To gain insights into how knowledge workers experience stress over an
+extended period of work, we examined the end of day survey responses
+collected from each participant to see if any identifiable trends
+emerged. We used data from 13 of the 14 participants - we excluded one participant
+from this analysis as they experienced atypical stress
+levels in the latter half of the study due to factors outside of our
+control.
+
+\subsection{Baseline Stress Levels}
+Common amongst all participants was a trend to select one stress rating
+far more frequently than any other. We will refer to this value as the
+participant's baseline stress level. All but one participant reported
+their perceived stress level for the day as their baseline stress
+level more than 50\% of the time. In total, the baseline values made
+up 65\% of the reported values collected from
+participants. Interestingly, while participants sometimes saw periods
+of sustained increases in stress, lasting as many as 6 consecutive
+workdays in the most extreme case, participants would always return to
+their baseline stress level at some point.
+
+The baseline stress level varied significantly between participants. 7
+participants (54\%) reported feeling average stress levels most frequently
+(rating 3 on our scale), while 5 (38\%) reported feeling little stress
+(rating 2) and 1 (8\%) reported feeling no stress at all (rating 1).
+
+\subsection{Stressful Days Tend to Cluster}
+Accounting for the variance between participants perceived stress
+baselines, we consider a stressful day to be one that represents a
+deviation of 1 or more stress levels above the participant's
+baseline. Of the 93 stressful days we observed in total, we found that
+39 (41\%) of these days occurred in groupings of two or more
+consecutive stressful workdays. The most common size of these groups
+was two workdays, while the largest group we observed was six
+workdays.  The day after a stressful day is much more likely to be a
+stressful day as compared to any other day with a 0.55 average increase
+over baseline, compared to 0.02 average increase over baseline.
+
+\subsection{Extreme Changes in Stress Levels are Rare}
+After accounting for each participant's perceived stress baseline, we
+examined the frequency of deviations from the baseline. We found that
+participants were far more likely to report a stress level that was
+within 1 point of their baseline, than to report a stress level 2 or
+more points away. These extreme deviations represented only 15\% of
+all reported values which differed from the participants baseline. As
+well, the majority (78\%) of these deviations came from just two
+participants. This suggests that some people may be less resilient to
+the stress of the workplace than others. For most participants,
+extremely stressful days were few and far between.
+
+\subsection{Explaining Stress Fluctuations}
+We created a linear mixed model to attempt to explain some of the
+stress that our participants were experiencing. We experimented with
+multiple independent variables, including those derived from computer
+interaction data such as total time spent actively on the computer,
+percentage of time idle, and amount of time spent on non work-related
+web browsing. We also looked at the day of the week and proximity to
+beginning or end of month as possible explanatory variables. However,
+ultimately we were unable to find any variable which showed a
+significant correlation with our participants perceived stress
+levels. This points to the need for additional instrumentation if we
+are to succesfully understand and predict stress in the workplace.
+
+%% \subsubsection{Tai Chi Intervention}
+%% We linked participants tai chi session attendance data with their self-reported percieved stress levels.
+Building a linear mixed model with tai chi attendance as a random effect, we found that tai chi attendance contributed a small amount to decreased stress in the work week immediately following the tai chi session (slope of -0.188, p < 0.05).
+However, we did not find a connection between attendance and stress on the day of the session.
+
+\section{Predicting Stress in the Moment}
 % random forest best
 \label{secOverallAccuracy}
-As a first step, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11} and performing a grid search analysis to determine the optimal hyperparameters for each classifier.
+
+To investigate whether stress, focus and awakeness can be predicted in
+the moment based on biometric measures.
+we investigated classifiers trained
+for each individual and across all participants. We report on the effectiveness of
+these classifiers and the features that are important in prediction of stress, focus
+and awakeness.
+
+\subsection{Selecting a Classifier Algorithm}
+
+There are many different algorithms that can be used to build a classifier. To select an
+algorithm, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11} and performed a grid search analysis to determine the optimal hyperparameters for each classifier.
 Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, and a multilayer perceptron neural network.
 The optimal values for random forest and the three output measures are listed in Table \ref{tab:hyperparams}.
-For the remainder of this paper, we will be referring to random forest classifiers trained with these hyperparameters.%\\[-0.1cm]
+For the remainder of this paper, we refer to random forest classifiers trained with these hyperparameters.%\\[-0.1cm]
 % for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)
@@
 
-\noindent\textit{Results of Individual Classifiers}\\
-Since peoples' experience of stress, focus, and awakeness (as well as their physiological manifestations) can vary a lot (e.g.,~\cite{Hernandez11}), we first trained individual classifiers for each participant (rather than a general one for all participants).
-The results of our analysis are reported in Table~\ref{tab:accuracy}.
-For our analysis, we report values of accuracy, one of the most commonly used metric to compare performance, as well as precision and recall of the classes of interest: 'stressed', 'not focused', and 'not awake'.
-Since the imbalance in the data can lead to high accuracy values if a classifier always just predicts the most likely/frequent class while ignoring the class of higher importance and interest, precision and recall of the class of interest can be more adequate metrics in this case~\cite{yap2014,bhattacharyya_data_2011,Hernandez11}.
+\subsection{Individual Classifiers}
+%\noindent\textit{Results of Individual Classifiers}\\
+Since peoples' experience of stress, focus, and awakeness (as well as
+their physiological manifestations) can vary substantially
+(e.g.,~\cite{Hernandez11}), we first trained individual classifiers
+for each participant (rather than a general one for all
+participants). The results of our analysis are reported in
+Table~\ref{tab:accuracy}. For our analysis, we report values of
+accuracy, one of the most commonly used metric to compare performance,
+as well as precision and recall of the classes of interest:
+'stressed', 'not focused', and 'not awake'. Since the imbalance in the
+data can lead to high accuracy values if a classifier always just
+predicts the most likely/frequent class while ignoring the class of
+higher importance and interest, precision and recall of the class of
+interest are also important to
+consider~\cite{yap2014,bhattacharyya_data_2011,Hernandez11}.
+
 % However, the imbalance in the data can lead to a classifier with high accuracy if it always just predicts the more likely class while ignoring the class of higher importance and interest in our case (i.e.
@@
 
-Overall, we were able to use the extracted physiological features to predict all three aspects with reasonable accuracy, precision, and recall, as well as to improve upon the baseline---a stratified random classifier that randomly chooses one of the two classes with a bias towards the larger class.
-While the individually trained classifiers improved on average across all participants upon the baseline in all cases excepting recall of 'not focused', the improvement was substantially higher for awakeness (85\% improvement in precision,  62\% in recall, and 7\% in accuracy) than for stress or focus.
-Also, the performance of the individually trained classifiers varied greatly across participants.
-While some participants showed a large improvement, for others the baseline performed much better than the individually trained classifier.
-For instance, for predicting 'stressed', the individual classifiers improved upon the baseline for S4, S6, S8, S11, S12, and S14 with a maximum improvement of 88.4\% in precision and 184.6\% in recall for S12, while they did worse for S1, S3, S5, S7, S9, S10, and S13, and in the worst cases did not correctly predict a single instance of 'stressed'.
-Typically users that have the lowest precision and recall values are those where the data is the most unbalanced.
-This translates to a scenario where the classifier will likely label the minority instances as part of the majority class due to the unbalanced ratio of the data.\\[-0.1cm]
+Overall, we were able to use extracted physiological features to
+predict all three aspects with reasonable accuracy, precision, and
+recall, as well as to improve upon the baseline---a stratified random
+classifier that randomly chooses one of the two classes with a bias
+towards the larger class. While the individually trained classifiers
+improved on average across all participants upon the baseline in all
+cases excepting recall of 'not focused', the improvement was
+substantially higher for awakeness (85\% improvement in precision,
+62\% in recall, and 7\% in accuracy) than for stress or focus. Also,
+the performance of the individually trained classifiers varied greatly
+across participants. While some participants showed a large
+improvement, for others the baseline performed much better than the
+individually trained classifier. For instance, for predicting
+'stressed', the individual classifiers improved upon the baseline for
+S4, S6, S8, S11, S12, and S14 with a maximum improvement of 88.4\% in
+precision and 184.6\% in recall for S12, while they did worse for S1,
+S3, S5, S7, S9, S10, and S13, and in the worst cases did not correctly
+predict a single instance of 'stressed'. Typically users that have the
+lowest precision and recall values are those where the data is the
+most unbalanced. This translates to a scenario where the classifier
+will likely label the minority instances as part of the majority class
+due to the unbalanced ratio of the data.\\[-0.1cm]
 % We attribute these discrepancies to the large differences in response distributions between participants, as well as to the subjectivity of self-reporting.
@@
 
-\noindent\textit{Feature Selection and Importance}\\
+\subsection{Feature Selection and Importance}
+%\noindent\textit{Feature Selection and Importance}\\
 There are a large variety of features that can be (and have been) calculated in previous research for each of the basic measurements listed in Table~\ref{signals}, such as the mean, standard deviation, maximum, and interquartile range.
@@
 
-\noindent\textit{Individual vs.\ General Model}\\
+\subsection{Individual vs.\ General Model}
+%\noindent\textit{Individual vs.\ General Model}\\
 Individual models are trained specifically for each individual and thus require a data collection period before they are capable of making accurate predictions.
 On the other hand, the idea of general models is to be able to train them on already collected data and then  to be able to apply them even to new and unseen individuals, thus overcoming the cold-start problem.
-Given the big individual differences in biometrics, training a general model to achieve an adequate accuracy for new individuals is not necessarily possible.
+Given the large individual differences in biometrics, training a general model to achieve an adequate accuracy for new individuals is not necessarily possible.
 

### 2019-09-20 bp_intro_v3.tex
refocused introduction on stress

@@
 distractions, highly complex and demanding tasks, and long working 
-hours~\cite{gonzalez2004constant,mark2008cost,czerwinski04diary}. These 
-challenges make it difficult for a knowledge worker to stay focused and
-impact interaction with their digital environment, the quality of work performed and productivity. For instance, the 
-continuous disruptions that knowledge workers face in office work environments 
-can lead to a higher error rate and a lower 
-performance~\cite{bailey2001effects,mark2008cost}. In some cases,
-the consequences are even more severe; e.g., in control room
-settings, the consequences of an erroneous intraction can lead to accidents valued in millions of dollars in losses. 
-%when a sleepy operator might, for example, miss early warning signs and cause...
-%\todo{Nick: can you find a good example here and citation?}. 
-%Similarly, stress at the workplace is a growing concern at the workplace and one of the most common work-related health problem that can lead to fatigue, burnout and various other illnesses, resulting in work absence and a high productivity loss~\cite{hockey1997stress,setz2010stress,wrs2010}.
-Less immediate, but just as insidious, stress at the workplace is a growing 
-concern and one of the most common work-related health problems. It leads to 
+hours~\cite{gonzalez2004constant,mark2008cost,czerwinski04diary}. 
+Consequently, stress in the workplace is an ever-growing concern. It leads to 
 fatigue, burnout and various other illnesses, ultimately resulting in work 
@@
 losses~\cite{hockey1997stress,setz2010stress,wrs2010}.
+But while the effects of stress are obvious the causes can be more subtle. Stress can be caused by
+a multitude of factors within the office including but not limited to workload, management issues, workplace harassment and bullying,
+constant interruptions, and noisy environments. It could also be the result of external stressors seeping into the workplace.
+Regardless of the cause, garnering an understanding of what stress looks like in the workplace could
+prove invaluable to helping us develop effective interventions.
+Likewise, being able to continuously measure stress in the workplace bears huge potential for better supporting 
+knowledge workers in combatting workplace challenges, and improving productivity 
+and well-being.
 
-Being able to continuously measure human aspects --- such as focus, awakeness, 
-and stress --- in the workplace bears huge potential for better supporting 
-knowledge workers combatting workplace challenges, and improving productivity 
-and well-being. Such measures could be used to provide better interaction with 
-their environment, such as management 
-of disruptions~\cite{iqbal2005effectiveness,zuger2017reducing}, to 
-automatically adjust lighting and temperature to reduce sleepiness, or to continuously monitor stress levels to trigger pro-active 
-interventions and reduce health problems.
+At the same time, these 
+challenges that contribute to stress levels can make it hard for knowledge workers to stay focused, impact interaction with their digital environment,degrade the quality of work performed and hinder productivity.
+For instance, the continuous disruptions that knowledge workers face in office work environments can lead to a higher error rate and lower 
+performance~\cite{bailey2001effects,mark2008cost}.
 
 Fortunately, with recent advances in sensing technologies there is increasing 
-hope that we can accurately and automatically measure such human aspects. A 
+hope that we can accurately and automatically measure stress and it's related aspects. A 
 number of HCI studies have discussed and investigated the use of biometrics to 
@@
 
-The objective of our research is to develop continuous (and automatic) 
-measures of focus, awakeness, and stress in the workplace. These measures are 
-targeted towards improving knowledge workers' productivity and 
-well-being in the future by customizing interaction with their environment; e.g., by automatically protecting the workers
-from audible interruptions during a high focus period by showering them 
-with white noise or by recommending stress-reducing interventions during an 
-extended period of high stress.
+In this research we examine what office stress actually looks like over long periods of time, and develop continuous (and automatic) 
+measures of stress, focus, and awakeness in the workplace. These measures are targeted towards improving knowledge workers' productivity and 
+well-being in the future by customizing interaction with their environment; e.g., by continuously monitoring stress levels and recommending stress-reducing interventions during 
+extended periods of high stress, automatically adjusting lighting and temperature to heighten awakeness, or improving management 
+of disruptions~\cite{iqbal2005effectiveness,zuger2017reducing} to aid with focus and productivity.
 
@@
 
-The results of our analysis show that biometrics of a single minimally invasive sensor can be used to predict productivity-related indicators accurately, with the abstract concept of focus (predictably) being the hardest to detect.
+The results of our analysis show that biometric signals collected from a single minimally invasive sensor can be used to predict stress and its related aspects accurately, with the abstract concept of focus (predictably) being the hardest to detect.
 The results also show that knowledge workers' self-reported levels of stress, focus and awakeness and their physiological manifestation and prediction can vary a lot between individuals.
@@
 \begin{itemize}
+	\item An qualitative examination of how knowledge worker stress presents and fluctuates in the wild.
 	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' awakeness, stress, and focus in the workplace based on an eight-week field study with 14 office workers.
-	\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
+	%\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
 	\item A discussion on the impact of applying this research to improve

### 2019-09-20 bp_intro_v3.tex
removed references to CI data

@@
 At the same time, these 
-challenges that contribute to stress levels can make it hard for knowledge workers to stay focused, impact interaction with their digital environment,degrade the quality of work performed and hinder productivity.
+challenges that contribute to stress levels can make it hard for knowledge workers to stay focused, impact interaction with their digital environment, degrade the quality of work performed and hinder productivity.
 For instance, the continuous disruptions that knowledge workers face in office work environments can lead to a higher error rate and lower 
@@
 modality was chosen to ease longitudinal deployment in the field. Using 
-machine learning, we created classifiers and analyzed their ability to predict 
-the three productivity-related human aspects. Similarly, we captured computer 
-interaction features from participants' everyday computer usage and
-analyzed how a model trained using the combination
-of these two training sets compares to the performance of models using each
-set individually.
+machine learning, we created classifiers and analyzed their ability to predict stress, focus, and awakeness levels.
+%Similarly, we captured computer 
+%interaction features from participants' everyday computer usage and
+%analyzed how a model trained using the combination
+%of these two training sets compares to the performance of models using each
+%set individually.
 
-The results of our analysis show that biometric signals collected from a single minimally invasive sensor can be used to predict stress and its related aspects accurately, with the abstract concept of focus (predictably) being the hardest to detect.
-The results also show that knowledge workers' self-reported levels of stress, focus and awakeness and their physiological manifestation and prediction can vary a lot between individuals.
-Our results further show that overall combining biometric features with computer interaction information
-results in an improvement
-over each of the training sets individually.
+In our analysis we identify several trends in day-to-day stress levels that emerged over the course of our eight-week study.
+The results of our analysis also show that biometric signals collected from a single minimally invasive sensor can be used to predict stress and its related aspects accurately, with the abstract concept of focus (predictably) being the hardest to detect.
+Our results further show that knowledge workers' self-reported levels of stress, focus and awakeness and their physiological manifestation and prediction can vary a lot between individuals.
+%Our results further show that overall combining biometric features with computer interaction information 
+%results in an improvement
+%over each of the training sets individually.
 The main contributions of our work are:

### 2019-09-20 bp_discussion.tex
Reformatted discussion

--- 2019-07-08 bp_discussion.tex
@@
 
-\vspace{0.05in}
-\noindent
-\textbf{Target Workspaces:}
-Currently, the cost of biometric sensors and necessary infrastructure, such as automated light and sound systems for adjusting the environment, makes our approach most appropriate for high-value workspaces, such as control rooms, command centers, or dispatch offices.
-However, as standard office settings become more personalizable (e.g., via adjustable desks, lighting, and sound showers) and sensor costs decrease, our approach could be applied to any office environment, and thus could impact a large percentage of modern workers.
-As in modern cars, temperature and lighting could be regulated on a per-person basis, which would allow the environment to react to the person's current state and to maximize each person's preferences and productivity (e.g., preferences of men and women in temperature~\cite{Karjalainen07}).
-Further research would be needed to identify how to balance needs across a group of office workers and how to handle conflicting levels between different group members.
+Humans experience stress, focus and awakeness in different ways.  In
+this paper, we have attempted to study these mental states in the
+workplace using both participant self-reports and biometric data. We
+discuss implications from our study for the workplace, including ways
+in which the information might inform digitally-controlled or
+digitally-informed parts of the workplace. We also discuss 
+challenges imposed by the data and possible future paths of
+research.
 
-\vspace{0.05in}
-\noindent
-\textbf{Imbalanced Data:}
-Study participants provided highly imbalanced data in their survey responses, with most participants only taking advantage of a subset of the Likert-scale values and the data points mostly being clustered around the middle of the scale, as can be seen in Table~\ref{responseDistribution}.
-While some of the imbalance is expected due to certain classes, such as 'not stressed', being more common in the workplace, this imbalance also provides challenges in the training and assessment of a machine learning classifier, as also found by others, e.g.~\cite{Exler16}.
-We addressed this for the training by oversampling in case of few data samples for the individual models and undersampling in case of a general model where more data was available.
-Especially in light of this imbalance in the data, the results we achieved with our models are encouraging.
-For the assessment of the classifiers' performance we addressed the imbalance by not just presenting accuracy, but also by focusing on prediction and recall to examine the classifier's performance in predicting the infrequent (yet more important) cases, such as when a user is struggling to stay awake and an intervention or warning might be needed most.
+\subsection{Impliciations for Workplaces}
+Being able to accurately recognize periods of high-stress in knowledge
+workers could enable more respectful workplaces. For example, an
+ability to sense and predict stress in the moment could help companies
+to prevent or de-escalate potentially dangerous situations in the
+workplace, e.g. confrontation between co-workers. Building an 
+understanding of stress and focus over time and in the moment
+could also help create workplaces that are more conducive
+to enabling knowledge workers to be more productive. For example,
+this information could be used to feed an  
+ an awareness dashboard of a team's stress level,
+and avoid digital interruptions in high-stress or high-focus periods
+similar to previous studies~\cite{zuger2017reducing}.  Building
+an understanding of awakeness and focus could also enable the creation
+of workplaces that are conducive to workers producing higher-quality
+work. For example, if awakeness or focus wans, they might 
+be enhanced by adapting lighting in the workplace 
+or scheduling breaks to prevent focus loss.
+
+Currently, the cost of biometric sensors and necessary infrastructure,
+such as automated light and sound systems for adjusting the
+environment, makes our approach most appropriate for high-value
+workspaces, such as control rooms, command centers, or dispatch
+offices. However, as standard office settings become more
+personalizable (e.g., via adjustable desks, lighting, and sound
+showers) and sensor costs decrease, our approach could be applied to
+any office environment, and thus could impact a large percentage of
+modern workers. As in modern cars, temperature and lighting could be
+regulated on a per-person basis, which would allow the environment to
+react to the person's current state and to maximize each person's
+preferences and productivity (e.g., preferences of men and women in
+temperature~\cite{Karjalainen07}). Further research would be needed to
+identify how to balance needs across a group of office workers and how
+to handle conflicting levels between different group members.
+
+%%\vspace{0.05in}
+%\noindent
+%\textbf{Implications for HCI Design and Workspaces}
+%% The results discussed in this paper establish great potential
+%% to increase the well-being of knowledge workers and the quality of their work
+%% through monitoring their biometric signals and computer interactions with the
+%% goal of recognizing their stress, awakeness, and focus levels in an uninvasive 
+%% and automatic manner. e
+
+\subsection{Tai Chi Intervention}
+As we described early, the second half of the field study data
+collection involved an intervention of Tai Chi. This intervention was
+used to keep participants motivated to continue over the long study
+period.  To investigate whether the intervention had any effect, we
+linked participants tai chi session attendance data with their
+self-reported percieved stress levels. Building a linear mixed model
+with tai chi attendance as a random effect, we found that tai chi
+attendance contributed a small amount to decreased stress in the work
+week immediately following the tai chi session (slope of -0.188, p <
+0.05). However, we did not find a connection between attendance and
+stress on the day of the session suggesting that tai chi might have
+long-term effects, but is not conducive at relieving stress close
+in time to the intervention.
+
+%% %\vspace{0.05in}
+%% \noindent
+%% \textbf{Ground Truth and Self-Reporting:} 
+\subsection{Ground Truth and Self-Reporting}
+Studying mental states, such as stress, awakeness and focus,
+requires collecting a valid ground truth from each participant.
+We spent considerable time when designing the study
+ determining the exact questions to ask of participants,
+consulting experts in the area, and basing
+the questions and wording on previous research and studies. 
+Despite the care taken, it is possible that the 
+gathered data lacks reliability and validity. Some have
+questioned the
+reliability and validity of self-reports as we used in our study
+due to subjective biases, lack of care in reporting, and the
+highly individual nature of reporting aspects such as
+stress~\cite{Hernandez11,Hovsepian15}. In addition, in contexts such
+as the workplace, as in our study,
+participants  might be afraid to genuinely report levels
+of aspects, such as sleepiness. Hence, there is a chance that the
+self-reports we gatheered  do not adequately reflect the ground
+truth of the underlying variables under investigation. It could even be
+the case that certain biometrics might represent a more accurate
+ground truth of the studied phenomena than the self-reports. This
+suggests that a more confirmative study rather than an inquiry study
+could be a better approach, and we will explore such routes in future
+work.
+
+%% %\vspace{0.05in}
+%% \noindent
+%% \textbf{Imbalanced Data:} 
+\subsection{Imbalanced Data}
+Study participants provided highly
+imbalanced data in their survey responses, with most participants only
+taking advantage of a subset of the Likert-scale values and the data
+points mostly being clustered around the middle of the scale, as can
+be seen in Table~\ref{responseDistribution}. While some of the
+imbalance is expected due to certain classes, such as 'not stressed',
+being more common in the workplace, this imbalance also provides
+challenges in the training and assessment of a machine learning
+classifier, as also found by others, e.g.~\cite{Exler16}. We addressed
+this for the training by oversampling in case of few data samples for
+the individual models and undersampling in case of a general model
+where more data was available. Especially in light of this imbalance
+in the data, the results we achieved with our models are
+encouraging. For the assessment of the classifiers' performance we
+addressed the imbalance by not just presenting accuracy, but also by
+focusing on prediction and recall to examine the classifier's
+performance in predicting the infrequent (yet more important) cases,
+such as when a user is struggling to stay awake and an intervention or
+warning might be needed most.
 % \textbf{Unbalanced Data:}
@@
 
-\vspace{0.05in}
-\noindent
-\textbf{Ground Truth and Self-Reporting:} 
-One of the key points for developing a good classifier for awakeness, focus and stress, is to collect a valid ground truth to be used as the output measure.
-When designing the study, we therefore spent an extensive amount of time on determining the exact questions to ask in the experience sampling, consulting experts in the area, and basing the questions and wording on previous research and studies.
-Yet, the reliability and validity of self-reports have been questioned in the past due to subjective biases, lack of care in reporting, and the highly individual nature of reporting aspects such as stress~\cite{Hernandez11,Hovsepian15}.
-In addition, in contexts such as the workplace, employees might be afraid to genuinely report levels of aspects, such as sleepiness.
-Hence, there is a chance that the collected experience samples do not adequately reflect the ground truth of the underlying variable under investigation.
-It could even be the case that certain biometrics might represent a more accurate ground truth of the studied phenomena than the self-reports.
-This suggests that a more confirmative study rather than an inquiry study could be a better approach, and we will explore such routes in future work.
+\subsection{Predicting Stress with Computer Interaction Data}
+Knowledge workers, a focus of our work and study, often spend
+a large amount of time each day interacting with information on 
+their computer at work. An interesting direction for future study
+is to consider whether this computer interaction data, which can
+be gathered non-invasively as work occurs, could serve to sense
+and predict stress, focus and awakeness. Features that could be
+investigated include keystrokes per minute, mouse clicks per minute
+and changes in the active title window.
 
-\vspace{0.05in}
-\noindent
-\textbf{Corpus Size:}
-In this study we collected data from 14 professionals over the course of eight weeks.
+%% %\vspace{0.05in}
+%% \noindent
+%% \textbf{Corpus Size:}
+%% In this study we collected data from 14 professionals over the course of eight weeks.
 By scholarly standards this constitutes a large dataset --- especially given that the participants were not students.
@@
 
-\vspace{0.05in}
-\noindent
-\textbf{Per-User Model vs. Across-Users Model:}
-Our results show that individually trained models outperform on average a more general model.
+%% %\vspace{0.05in}
+%% \noindent
+%% \textbf{Per-User Model vs. Across-Users Model:}
+%% Our results show that individually trained models outperform on average a more general model.
 Yet, given the increase in model performance with training data set size, we expect that with enough data, a generic model will perform reasonably close to the individually trained models.

### 2019-09-20 bp_results_1.tex
removed some RQs, moved data prep section to be with rest of ML stuff

@@
 \section{Observed Trends Over Time in Stress Levels}
+\label{stressTrends}
 
@@
 and awakeness.
+
+\subsection{Data Preparation}
+
+In this section we describe how we preprocessed the collected data for use in training and testing machine learning models.
+
+\subsubsection{Data Linking}
+
+We linked the collected biometric data and survey responses for each participant.
+Linking the data is necessary to construct training and test datasets for use in creating and evaluation machine learning models.
+
+Our linking approach is as follows.
+From the start time of each survey response, we look back one hour for available biometric data.
+For each minute in that hour-long time window, we check for biometric data to associate with the survey response.
+For example, if a participant started a survey response at 11:05am, we look for biometric data in the time frame 10:05am to 11:05am.
+If biometric data is available in the hour-long time window, we consider the survey response to have associated biometric data.
+Otherwise, we exclude the survey response from our dataset.
+
+Reasons for a survey response to lack associated biometric data include:
+\begin{itemize}
+\item The participant not wearing the Everion in the hour before before beginning the survey
+\item The Everion not recording data in the hour before the participant began the survey (e.g., due to low battery)
+\item Biometric data not being uploaded successfully to the server
+%\item Compatibility issues between the sensor and the OS tracking the participant's data
+%\item Issues accessing the data uploaded by the participants
+\end{itemize}
+
+Figure~\ref{surveyBio} illustrates the number of survey responses with associated biometric data for each study participant. 
+%This was added to address REviewer 2's concern: Is it correct that the maximum number of responses should have been 112 on the surveys?
+The total number of responses per participant is affected by their response rate and by the number days out-of-office (e.g., vacations, holidays, etc.). 
+Participant S2 and S12 have particularly low numbers of usable survey responses. In each of these cases, the issue related to biometric data not being uploaded successfully to the server.
+
+\begin{figure}
+  \centering
+      \includegraphics[width=0.5\textwidth]{DuringTheDay.png}
+  \caption{The figure shows the biometric data available per each participant. Green sections represent survey responses with biometric
+  data available, red sections responses with no biometric data available}
+   \label{surveyBio}
+   \vspace*{-4mm}
+\end{figure}
+
+\subsubsection{Feature Extraction}
+We extracted features from the biometric data to provide as input to machine 
+learning models. Previous 
+studies~ \cite{vorburger05,zuger2015interruptibility} identify time windows as an 
+important factor that impacts the prediction accuracy of a classifier. We 
+considered many time windows from the literature on biometric 
+analysis~\cite{zuger18}, ranging from 10 seconds to 3 hours. Specifically, we 
+considered the following time windows: \textit{10sec, 20sec, 30sec, 45sec, 
+1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, 
+3hour}.
+
+From the start time of each survey response, we look back the amount of time 
+that corresponds to each time window, and we create features for all of the 
+biometric data available in that time window. For example, if a participant 
+started a survey response at 11:05am, for the 30min time window, we create 
+features using all of the available biometric data from 10:35am to 11:05am. 
+For each time window, we calculate 10 statistical measurements from the 
+biometric data to create 10 distinct features. Specifically, the 10 
+statistical measurements are: mean, standard deviation, variance, median, 
+percentile25, percentile75, interquartile range, maximum, minimum, and 
+range. Thus, for each survey response, we generate a large number of 
+corresponding features based on three factors: biometric measurement, time 
+window, and statistical measurement. In addition to these biometric 
+features, we also considered the time of day in which the questions were 
+asked. These features are created to predict the responses described by the 
+ground truth.
+
+\subsubsection{Response Transformations}
+Table~\ref{responseDistribution} illustrates the distribution of responses from each participant for each of the three survey questions (listed in Section~\ref{sec:Surveys}).
+The figure shows that there is a notable imbalance in the distribution of the self-reported responses provided by the participants.
+Most participants did not use all five points of the five-point Likert scale in their responses, and the distributions tend to skew toward one side or the other, depending on the question.
+Thus, we binarized the survey data into a two-point scale to give the machine learning models the best possible chance to make useful predictions.
+The two points in the binary scale represent negative or positive responses for each of the three human aspects of interests (e.g., not stressed or stressed).
+
+We binarized the survey responses as follows.
+For each participant, we calculated the median response value for each question.
+We classified each response below the median as 0 ('negative') and each response above the median as 1 ('positive').
+The distribution for the stress question skewed left, so we included the median values in the 'positive' class, while the distributions for focus and awakeness skewed right, so we included those median values in the 'negative' class.
+
+\subsubsection{Oversampling}
+Even after binarizing the responses as described in the previous section, we found the distribution of responses was still quite imbalanced for many of our participants.
+This can be seen in the distribution columns in Table \ref{tab:accuracy}.
+To combat this, we applied random oversampling to our training sets, which artificially rebalances the dataset by creating randomly replicated data in the minority class.
+This has been a commonly used technique in previous studies on unbalanced datasets \cite{chawla2004,yap2014}.
+
+\begin{table}
+  \centering
+      \includegraphics[width=0.5\textwidth]{distributiontable.pdf}
+  \caption{The distribution of the responses of each participant to the three questions asked during the day are shown.
+Each bar in the histograms represent one of the 5 values on the 5-point Likert scale we asked participants to respond with, where the far left side of the histograms are 1/Not at all, and the far right sides are 5/Extremely}
+   \label{responseDistribution}
+   \vspace*{-2mm}
+\end{table}
 

### 2019-09-20 bp_studyMethod.tex
removed some RQs, moved data prep section to be with rest of ML stuff

@@
 
-The goal of our study is to understand the efficacy of biometric measurements in predicting stress, focus, and awakeness. To that end we posed the following research questions.
+The goal of our study is to better understand understand stress in the workplace, as well as investigating the efficacy of biometric measurements in predicting stress, focus, and awakeness.
+To that end we posed the following research questions.
 
 \vspace{0.5em}
-\noindent\textbf{RQ1}: Can we use biometric measurements to accurately predict stress,
+\noindent\textbf{RQ1}: How do knowledge worker stress levels change over long periods of time?
+
+\vspace{0.5em}
+\noindent\textbf{RQ2}: Can we use biometric measurements to accurately predict stress,
 focus, and awakeness?
@@
 \vspace{0.5em}
-\noindent\textbf{RQ2}: What is the minimum number of samples needed to accurately predict stress, focus, and awakeness?
 
-\vspace{0.5em}
-\noindent\textbf{RQ3}: What is the minimum time window needed to accurately predict stress, focus, and awakeness?
-
-\vspace{0.5em}
-\noindent\textbf{RQ4}: How do biometric measurements compare to computer interaction measurements for predicting stress, focus, and awakeness?
-
-\vspace{0.5em}
-\noindent\textbf{RQ5}: How do knowledge worker stress levels change over long periods of time?
-
-\vspace{0.5em}
-\noindent\textbf{RQ6}: Can practicing Tai Chi help to reduce stress levels in the office?
-
-\vspace{0.5em}
-
-We answer RQ1 in Section~\ref{secOverallAccuracy} by analyzing the overall performance of different
+To answer RQ1, in Section~\ref{stressTrends} we perform a qualitative evaluation of the day-to-day stress levels experienced by our participants, as gathered from daily surveys.
+We answer RQ2 in Section~\ref{secOverallAccuracy} by analyzing the overall performance of different
 machine learning models and by comparing the best performing models to baseline approaches.
-To answer RQ2, in Section~\ref{secLearningCurve} we identify the smallest number of training samples necessary to obtain satisfactorily high prediction accuracy and acceptably low mean squared error.
-We answer RQ3 in Section~\ref{secMinimumTW} by considering 16 different time windows for collection of biometric measurements and by identifying the smallest time window on which we can base our training features without sacrificing prediction accuracy.
-Finally, to answer RQ4, in Section~\ref{secCI} we compare the prediction accuracy of machine learning models trained on biometric data versus those trained on computer interaction data.
 
@@
 detailed below:
-\begin{itemize}
+\begin{enumerate}
 \item How awake have you been today?
@@
 \item How do you feel about your work day?
-\end{itemize}
+\end{enumerate}
 

### 2019-09-20 bp_conclusion.tex
Updated

--- 2019-05-31 bp_conclusion.tex
@@
 \section{Conclusions} 
-Stress, awakeness and focus at work are highly relevant aspects when it comes to productivity and well-being at the workplace.
-In this paper, we presented the results of a study with 14 professional knowledge workers in their workplace over an eight week period to examine the ability of biometrics for predicting these producitivity-related aspects.
-The longitude and in situ placement of the study---two key components of the study---as well as the breadth of human aspects examined, support and extend previous work.
-Based on twice collected daily survey responses and continually collected biometrics data we were able to create a model that was able to predict awakeness, stress, and focus, outperforming the baseline by as much as 84.6\% in the case of awakeness.
-The analysis further shows that there is a continuous increase in performance with the number of training instances, suggesting that there is more potential to improve the model's performance with further data points.
-Interestingly, while aspects such as stress have previously been predominantly linked to biometrics, our comparison with a model build from computer interaction features shows that even these types of features might bear big potential in predicting some of these productivity-related human aspects in the workplace.
+Stress, awakeness and focus at work are highly relevant aspects when
+it comes to productivity and well-being at the workplace. In this
+paper, we presented the results of a study with 14 professional
+knowledge workers in their workplace over an eight week period to
+better understand how workres experience stress over time and to
+examine the ability of biometrics to predict these
+productivity-related aspects. The longitudinal and in situ placement of
+the study support and extend previous work. Based on
+twice collected daily survey responses, we observed that
+although participants sometimees saw preiods of sustained strses,
+they would always return to a baseline stress level for them
+at some point. We also observed that stress levels seldom spiked,
+but when they did rise, the rise in stress tended to last more
+than a day. In addition to the survey responses, we 
+continually collected
+biometrics data with which we were able to create a model that was able to
+predict awakeness, stress, and focus.
 
-The results also open up new opportunities to help increase knowledge workers' productivity and well-being, ranging from instantaneously taking action to prevent potentially risky situations and prevent accidents due to a lack of focus or awakeness, all the way to recommending interventions to reduce stress if it becomes more chronic.
+These results open up new opportunities to help increase knowledge
+workers' productivity and well-being, ranging from instantaneously
+taking action to prevent potentially risky situations and prevent
+accidents due to a lack of focus or awakeness, all the way to
+recommending interventions to reduce stress if it becomes more
+chronic.
 

### 2019-09-20 bp_threats.tex
addressed concerns from Reviewer 1 using information from their review and our rebuttal
Updated

@@
 \textbf{External Validity:}
-It is a threat that the results of this study
-will not generalize to a broader population of office workers.
-To address this concern, we collected participants
+The results of our study may 
+ not generalize to a broader population of office workers.
+To mitigate this risk, we collected participants
 from a wide variety of different departments
@@
 
-Another threat is that our results won't generalize
-to a different office environment. We have conducted
-this study in a typical office environment, common
+Another threat is that our results may notgeneralize
+to a different office environment. We  conducted
+this study in a typical office environment,  similar to many
 among technology workers across the world.
@@
 features and the productivity-related aspects would be inaccurate.
-To address this concern we collected the data
+To mitigate this risk,  we collected the data
 in a rote environment and in a regular manner 
@@
 It is possible that the amount of data collected
-is not enough
+is not sufficienet
 to draw valid conclusions. To address this threat, 
@@
 are unrelated biometric signals.
-To mitigate this threat we used an state-of-the-art
+To mitigate this risk, we used an state-of-the-art
 device that captures a large number of highly accurate biometric
@@
 productivity aspects from each user.
+In an effort to maintain this research applicable to real-world environments, we picked this already existing device, even when, as a trade-off, we could not capture more descriptive and more intrusive signals such as SDNN, SCL, SCR, eye tracking, or brain activity.
+
 We also perform a grid analysis in our machine learning model

### 2019-09-20 bp_discussion.tex
updated

@@
 As we described early, the second half of the field study data
-collection involved an intervention of Tai Chi. This intervention was
+collection involved an intervention of tai chi. This intervention was
 used to keep participants motivated to continue over the long study
@@
 linked participants tai chi session attendance data with their
-self-reported percieved stress levels. Building a linear mixed model
-with tai chi attendance as a random effect, we found that tai chi
+self-reported percieved stress levels. Building a linear mixed model, we found that tai chi
 attendance contributed a small amount to decreased stress in the work

### 2019-09-20 bp_results_1.tex
Not sure what changed
updated

@@
 
+\begin{figure}
+  \centering
+      \includegraphics[width=0.5\textwidth]{dailystress_s1.pdf}
+      \includegraphics[width=0.5\textwidth]{dailystress_s3.pdf}
+  \caption{The day-to-day stress levels reported by two participants (S1 and S3) are shown.
+The y-axis represents values on the 5-point Likert scale we asked participants to respond with ranging from 1/Not at all stressed to 5/Extremely stressed.
+The x-axis represents the day of the study on which the value was recorded (from 0-45).
+The x-axes are slightly different for the 2 participants as they did not begin and end the study on exactly the same days.}
+   \vspace*{-2mm}
+\end{figure}
+
 The baseline stress level varied significantly between participants. 7
@@
 are to succesfully understand and predict stress in the workplace.
-
-%% \subsubsection{Tai Chi Intervention}
-%% We linked participants tai chi session attendance data with their self-reported percieved stress levels.
-Building a linear mixed model with tai chi attendance as a random effect, we found that tai chi attendance contributed a small amount to decreased stress in the work week immediately following the tai chi session (slope of -0.188, p < 0.05).
-However, we did not find a connection between attendance and stress on the day of the session.
 

### 2019-09-20 bp_studyMethod.tex
updated

@@
 During the eight week study period, a computer activity tracker was installed and active on 10 of the participants' company-issued laptops (note: four participants from the original sample declined for privacy reasons).
-
-The goal of our study is to better understand understand stress in the workplace, as well as investigating the efficacy of biometric measurements in predicting stress, focus, and awakeness.
-To that end we posed the following research questions.
-
-\vspace{0.5em}
-\noindent\textbf{RQ1}: How do knowledge worker stress levels change over long periods of time?
-
-\vspace{0.5em}
-\noindent\textbf{RQ2}: Can we use biometric measurements to accurately predict stress,
-focus, and awakeness?
-
-\vspace{0.5em}
-
-To answer RQ1, in Section~\ref{stressTrends} we perform a qualitative evaluation of the day-to-day stress levels experienced by our participants, as gathered from daily surveys.
-We answer RQ2 in Section~\ref{secOverallAccuracy} by analyzing the overall performance of different
-machine learning models and by comparing the best performing models to baseline approaches.
 

### 2019-09-20 bp_results_1.tex
updating intro with gail's comments
fixing typos
updated results

@@
 %\section{Analysis and Results}
-
-\begin{table*}[h]
-  \centering
-  \includegraphics[width=1.0\textwidth]{rq1performance.pdf}
-  \caption{Results of predictions using the individual models.
-The distribution columns show a bar chart of the response distribution (negative/positive) for each of the three variables.
-The baseline row represents the averaged results of our baesline classifier.
-The general row shows the averaged results of our models trained on all participants.}\label{tab:accuracy}%\todo{revise caption}}
-  \vspace*{-4mm}
-\end{table*}
 
@@
    \vspace*{-2mm}
+   \label{fig:dailyStress}
 \end{figure}
@@
 (rating 2) and 1 (8\%) reported feeling no stress at all (rating 1).
+
+Figure \ref{fig:dailyStress} illustrates these points, showing both participants tendency to report and return to baseline stress levels
 
@@
 
-\subsection{Selecting a Classifier Algorithm}
-
-There are many different algorithms that can be used to build a classifier. To select an
-algorithm, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11} and performed a grid search analysis to determine the optimal hyperparameters for each classifier.
-Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, and a multilayer perceptron neural network.
-The optimal values for random forest and the three output measures are listed in Table \ref{tab:hyperparams}.
-For the remainder of this paper, we refer to random forest classifiers trained with these hyperparameters.%\\[-0.1cm]
-% for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)
 \begin{table}[h]
@@
 \end{table}
+
+\subsection{Selecting a Classifier Algorithm}
+
+There are many different algorithms that can be used to build a classifier. To select an
+algorithm, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11} and performed a grid search analysis to determine the optimal hyperparameters for each classifier.
+Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, and a multilayer perceptron neural network.
+The optimal values for random forest and the three output measures are listed in Table \ref{tab:hyperparams}.
+For the remainder of this paper, we refer to random forest classifiers trained with these hyperparameters.%\\[-0.1cm]
+% for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)
+
+\begin{table*}[h]
+  \centering
+  \includegraphics[width=1.0\textwidth]{rq1performance.pdf}
+  \caption{Results of predictions using the individual models.
+The distribution columns show a bar chart of the response distribution (negative/positive) for each of the three variables.
+The baseline row represents the averaged results of our baesline classifier.
+The general row shows the averaged results of our models trained on all participants.}\label{tab:accuracy}%\todo{revise caption}}
+  \vspace*{-4mm}
+\end{table*}
 
@@
 consider~\cite{yap2014,bhattacharyya_data_2011,Hernandez11}.
+For some users (i.e., S1.
+as seen In Table \ref{responseDistribution}), the imbalance in their data was so extreme that even after adjusting by oversampling it was impossible to create a reasonable classifier.
+These scenarios are difficult to predict as any classifier will not have an appropriate balance of stress and not stressed situations to learn from.
 
@@
 lowest precision and recall values are those where the data is the
-most unbalanced. This translates to a scenario where the classifier
-will likely label the minority instances as part of the majority class
-due to the unbalanced ratio of the data.\\[-0.1cm]
+most unbalanced.\\[-0.1cm]
 % We attribute these discrepancies to the large differences in response distributions between participants, as well as to the subjectivity of self-reporting.

### 2019-09-20 bp_results_1.tex
update results

@@
 as seen In Table \ref{responseDistribution}), the imbalance in their data was so extreme that even after adjusting by oversampling it was impossible to create a reasonable classifier.
-These scenarios are difficult to predict as any classifier will not have an appropriate balance of stress and not stressed situations to learn from.
+These scenarios are difficult to predict as any classifier will not have enough variance in its training data for the 'stressed' situation to adequately distinguish it from the non-stressed case.
 

### 2019-09-20 bp_results_1.tex
added ref to figure 2

@@
 
-Figure \ref{fig:dailyStress} illustrates these points, showing both participants tendency to report and return to baseline stress levels
+Figure \ref{fig:dailyStress} illustrates these points, showing both participants tendency to report and return to baseline stress levels, as well
+as a distinct difference in baseline stress level (rating 2 for S1 vs rating 3 for S3).
 

### 2019-09-20 bp_studyMethod.tex

@@
 \section{Field Study}
-We conducted an eight-week field study with 14 participants to investigate the feasibility of predicting stress, focus, and awakeness based on biometric signals.
-On each workday for eight consecutive weeks, each of the 14 participants wore a biometric sensor while performing their normal work tasks and completed two electronic surveys to report their current self-perceived levels of stress, focus, and awakeness to establish a ground truth.
-During the eight week study period, a computer activity tracker was installed and active on 10 of the participants' company-issued laptops (note: four participants from the original sample declined for privacy reasons).
+We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress over time and the feasibility of predicting stress, focus, and awakeness based on biometric signals.
 
@@
 All participants work for a research organization within the company, but their job functions span line management, laboratory science, scientific research, technology evaluation, and software development.
+
+\subsection{Procedure}
+We performed an in-situ study over the course of eight weeks.
+For this, we first informed participants about the study purpose and procedure, handed out biometric sensors and introduced and explained the self-reporting to the participants.
+For 10 of the participants, we further installed a computer activity tracker on their company-issued laptops (note: four participants from the original sample declined for privacy reasons).
+After the initial setup, participants were asked to fill out 3 surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}).
+At the end of the eight weeks, we collected the biometric sensors and performed a short follow-up interview on the study and the participants' experiences.
+
+In the second month of the study, we offered participants two Tai Chi classes per week (each in the middle of the day) in addition to the daily experience sampling and asked them to attend one class a week.
+We chose to offer Tai Chi class for two reasons: first, to incentivise participants for their continued participation and for keeping them engaged; and second, for their said benefits of reducing stress based on our consultation with researchers in psychiatry, psychology and more specifically on mindfulness practices.
 
@@
 
-\subsection{Intervention}
-At the midpoint of our study, we introduced an intervention in the form of Tai Chi classes, with the goal of examining the efficacy of Tai Chi or other mindfulness excercises for reduction of office stress.
+%\subsection{Intervention}
+%At the midpoint of our study, we introduced an intervention in the form of Tai Chi classes, with the goal of examining the efficacy of Tai Chi or other mindfulness excercises for reduction of office stress.
 Classes were offered two times a week in the middle of the work day.

### 2019-09-20 bp_results_1.tex
updates

@@
 \subsection{Explaining Stress Fluctuations}
-We created a linear mixed model to attempt to explain some of the
-stress that our participants were experiencing. We experimented with
-multiple independent variables, including those derived from computer
-interaction data such as total time spent actively on the computer,
+We created a linear mixed model with participants as random effects
+to attempt to explain some of the
+stress that our participants were experiencing. We experimented with day of the week and proximity to
+beginning or end of month as possible explanatory variables. We also calculated
+features from the gathered computer interaction data where available - namely total time spent actively on the computer,
 percentage of time idle, and amount of time spent on non work-related
-web browsing. We also looked at the day of the week and proximity to
-beginning or end of month as possible explanatory variables. However,
-ultimately we were unable to find any variable which showed a
-significant correlation with our participants perceived stress
+web browsing. However,
+ultimately we were unable to find any variable which had
+significant (p < 0.05) explanatory power with respect to our participants perceived stress
 levels. This points to the need for additional instrumentation if we

### 2019-09-20 bp_studyMethod.tex
updates

@@
 
-In the second month of the study, we offered participants two Tai Chi classes per week (each in the middle of the day) in addition to the daily experience sampling and asked them to attend one class a week.
+In the second month of the study, we offered participants two one-hour Tai Chi classes per week (each in the middle of the day) in addition to the daily experience sampling and asked them to attend one class a week.
 We chose to offer Tai Chi class for two reasons: first, to incentivise participants for their continued participation and for keeping them engaged; and second, for their said benefits of reducing stress based on our consultation with researchers in psychiatry, psychology and more specifically on mindfulness practices.
@@
 
+\subsubsection{Computer Interaction Data}
+To gain a better understanding of our participants day-to-day work activities, we installed an open source computer interaction monitor (reference omitted for doubleblind.) to track the participant's active window, as well as keyboard and mouse activity.
+4 of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13) so for this reason we did not include this data in our predictive model.
+However we did look at the computer interaction data as a potential part of an explanatory model.
+
 \subsubsection{Surveys}

### 2019-09-20 bp_studyMethod.tex
no message

--- 2019-09-20 bp_studyMethod.tex
@@
 
+\subsubsection{Computer Interaction Data}
+To gain a better understanding of our participants day-to-day work activities, we installed an open source computer interaction monitor (reference omitted for doubleblin.).
+The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity.
+Four of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13).
+Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.
+
 \subsubsection{Surveys}

### 2019-09-20 bp_discussion.tex

@@
 
-\subsection{Impliciations for Workplaces}
+\subsection{Implications for Workplaces}
 Being able to accurately recognize periods of high-stress in knowledge
@@
 to enabling knowledge workers to be more productive. For example,
-this information could be used to feed an  
- an awareness dashboard of a team's stress level,
+this information could be used to feed an awareness dashboard of a team's stress level,
 and avoid digital interruptions in high-stress or high-focus periods
@@
 of workplaces that are conducive to workers producing higher-quality
-work. For example, if awakeness or focus wans, they might 
+work. For example, if awakeness or focus decreases, they might 
 be enhanced by adapting lighting in the workplace 
@@
 
-\subsection{Tai Chi Intervention}
-As we described early, the second half of the field study data
-collection involved an intervention of tai chi. This intervention was
-used to keep participants motivated to continue over the long study
-period.  To investigate whether the intervention had any effect, we
-linked participants tai chi session attendance data with their
-self-reported percieved stress levels. Building a linear mixed model, we found that tai chi
-attendance contributed a small amount to decreased stress in the work
-week immediately following the tai chi session (slope of -0.188, p <
-0.05). However, we did not find a connection between attendance and
-stress on the day of the session suggesting that tai chi might have
-long-term effects, but is not conducive at relieving stress close
-in time to the intervention.
+\subsection{The Effect of Tai Chi on Stress}
+As described in our study procedure, we offered and asked participants to attend a one-hour Tai Chi class per week for the last four weeks of the study.
+We recorded each participant's attendance, including which day they attended the offered classes.
+The two primary reasons for this intervention was to keep participants motivated to continue the self-reports over the long study period, and to offer a technique that might help reduce stress.
+
+While this was not the focus of our study, we performed a secondary analysis to examine whether the Tai Chi classes had any effect on the participants' stress levels in the work days directly following the Tai Chi session.
+For this, we build a linear mixed model with the self-reported daily stress level as dependent variables and the participants as random effects.
+We found that Tai Chi attendance contributed a small amount to decreased stress in the work week immediately following the Tai Chi session (slope of -0.188, p <0.05).
+However, we did not find a connection between attendance and stress on the day of the session suggesting that Tai Chi might have long-term effects, but is not conducive at relieving stress close in time to the intervention.
+Overall, the Tai Chi thus had a small impact on the collected data of the second month of the study, which poses a threat to the validity to our observed trends for this period of time.
+However, since knowledge workers might attend these kinds of classes on their own doing, we believe that this is negligible, and the analysis rather provides a weak indication that this kind of stress intervention might in fact help to reduce stress in the wild.
 
@@
 of aspects, such as sleepiness. Hence, there is a chance that the
-self-reports we gatheered  do not adequately reflect the ground
+self-reports we gathered  do not adequately reflect the ground
 truth of the underlying variables under investigation. It could even be
@@
 investigated include keystrokes per minute, mouse clicks per minute
-and changes in the active title window.
+and changes in the active window title.
 

### 2020-02-03 ChiSubmission/RebuttalCHI2019.txt
Adding a new folder UbiCompSubmission with ACM Large format for Ubicomp

Rebuttal
We would like to thank the reviewers for valuable and constructive feedback and appreciate the opportunity to address the main comments 
expressed in them and how we plan to address them in the revised version.

Relevance to CHI:
In regard to the relevance of our work to the CHI community, we agree with the reviewers and will provide a better framing of the work’s 
relevance and potential impact to the CHI community. In particular, we will state that knowledge workers’ stress, and lack of focus or 
awakeness can have a significant impact on their interaction with computers and their work in general. For control room operators, similar to 
car drivers, a lack of focus and awakeness can even lead to severe incidents. We will further state, that various HCI researchers have 
therefore previously examined these and similar variables, in particular, (a) their occurrence, cause and effect in various context, including 
knowledge work, e.g. [50,51], (b) how they can be measured, e.g. [Reb5], and (c) how they might be modulated/regulated, e.g. [Reb7]. While 
previous work mostly focused on smaller or controlled lab studies, our study extends this work by performing a field study to examine whether 
and how these three variables--stress, focus and awakeness--can be measured over an extended period of time in a knowledge workers’ work 
context. We will further state that our results show that a combination of sensors can be used in the field to automatically measure these 
variables, in particular awakeness and stress with a certain accuracy, and that even less invasive computer interaction sensors can be used for 
measuring these variables in certain contexts. In addition, we will extend the discussion to include the usage of such a measurement in the 
workplace, in particular in control rooms as well as regular knowledge worker’s contexts, to modulate stress and awakeness with the goal to 
increase the wellbeing of the worker and the quality of their work, e.g. through individual approaches such as adapting the lighting in a 
control room, or scheduling breaks, and more team based approaches such as an awareness dashboard of a teams’ stress level.

Related work
Regarding the related work, we agree with the reviewers and will extend the comparison of our work to previous work. As stated above, we will 
provide a better framing of our work and situate it better in related work in terms of the study of these variables, the measuring thereof and 
the approaches to regulate/modulate them. We further thank the reviewers for their pointers to relevant related work and will add them. In 
particular, we will also state that previous approaches have already shown the feasibility of applying machine learning to biometric data to 
measure variables, such as stress, interruptibility, negative emotion or confusion during computer interaction [Reb1, Reb3, Reb4] and that 
these approaches vary in the variables examined, the invasiveness of the sensors and the length and context of the studies. We will state that 
many previous studies focused on shorter time frames (mostly hours, but in one case also up to 5 days [Reb2] or 2 weeks [Reb1]), but that none 
examined an 8 week period in a real-world office environment and these variables.

Data collection and preparation:
We thank the reviewers for their comments on the data collection and will clarify this aspect in our paper. We will state that we decided upon 
our data collection method based on prior work [Reb3,Reb4] and through extensive piloting with users. In the end we decided on the text 
messaging approach since pilot participants preferred these over other means also due to them being accessible and noticeable anywhere in the 
office. Biometric data collection was performed with the state-of-the-art Everion device which captures a broad set of signals (Table 1). We 
are aware that there are biometric signals not being captured by this device, or captured at a higher granularity that desired. However, in an 
effort to maintain this research applicable to real-world environments, we picked this already existing device, even when, as a trade-off, we 
could not capture more descriptive and more intrusive signals as SDNN, SCL, SCR, eye tracking, or brain activity.

Technical implementations:
Low precision and recall values in some of the users in Table 3 are due to the unbalanced nature of the responses of these particular users. It 
is worth noticing how users that have the lowest precision and recall values are those where the data is the most unbalanced (column 
“Distribution”, Table 3). This translates to a scenario where the classifier will likely label the minority instances as part of the majority 
class due to the unbalanced ratio of the data.
As to clarify how the data is labelled, the participants do not label each time window analyzed by the features (e.g., the participant does not 
label the last 3 hours as “awake”). The participants respond to how they are currently feeling, and based in the response, we create a set of 
features using different time windows (Figure 5), finally we analyze which features (time windows and biometric signals) can predict best the 
“awake” value.

Reb1. M. Zuger et al. Analyzing Interruptibility in the Office: A Field Study on the Use of Biometric and Computer Interaction Sensors. CHI18
Reb2. A. Sano et al. Stress Recognition using Wearable Sensors and Mobile Phones. ACII13
Reb3. S. Lallé et al. Predicting Confusion in Information Visualization from Eye Tracking and Interaction Data. IJCAI16
Reb4. P. Panwar et al. Detecting Negative Emotion for Mixed Initiative Visual Analytics. CHI18
Reb5. Hernandez et al. Under pressure: Sensing Stress of Computer Users. CHI14
Reb6. G. Mark et al. Email Duration, Batching and Self-interruption: Patterns of Email Use on Productivity and Stress. CHI16
Reb7. J. Snyder et al. MoodLight: Exploring Personal and Social Implications of Ambient Display of Biosensor Data. CSCW15

### 2020-02-03 UbiCompSubmission/bp_conclusion.tex
Adding a new folder UbiCompSubmission with ACM Large format for Ubicomp

--- 2019-09-20 bp_conclusion.tex
@@
 knowledge workers in their workplace over an eight week period to
-better understand how workres experience stress over time and to
+better understand how workers experience stress over time and to
 examine the ability of biometrics to predict these
@@
 twice collected daily survey responses, we observed that
-although participants sometimees saw preiods of sustained strses,
+although participants sometimes saw periods of sustained stress,
 they would always return to a baseline stress level for them

### 2020-02-03 UbiCompSubmission/bp_relatedWork.tex
Adding a new folder UbiCompSubmission with ACM Large format for Ubicomp

--- 2019-09-19 bp_relatedWork.tex
@@
 \section{Related Work}
-We review two categories of related work for this study.
-First, we review work related to the productivity-related indicators that we considered in our study, including how they are detected and prevented in several risk-inducing scenarios.
-Second, we review work related to biometrics, including how they have been studied in the office environment or applied to assess cognitive states and processes.
+
+The study and prediction approach on which we report in this paper is related
+to previous studies of productivity-related indicators and studies
+using biometerics to predict these indicators. We consider
+related work in each of these categories in turn.
 
@@
 
-Previous approaches~\cite{zuger18,Lalle16,Panwar18} have shown the feasibility of applying machine learning 
-to biometric data to measure variables such as stress, interruptibility, 
-negative emotion, or confusion during computer 
-interaction. These approaches vary in the 
-variables examined, the invasiveness of the sensors, and the length and 
-context of the studies. Many previous studies focused on 
-shorter time frames (mostly hours, but in one case also up to five 
-days~\cite{sano2013stress} 
-or two weeks~\cite{zuger18}). To the best of the authors' knowledge, no 
-other study has examined such an extensive period (eight weeks) in a real-
-world office environment and the inspected variables (stress, awakeness, and focus).
+Our study focuses on the productivity-related indicators of stress, focus and awakeness.
+
+%% Previous approaches~\cite{zuger18,Lalle16,Panwar18} have shown the feasibility of applying machine learning 
+%% to biometric data to measure variables such as stress, interruptibility, 
+%% negative emotion, or confusion during computer 
+%% interaction. These approaches vary in the 
+%% variables examined, the invasiveness of the sensors, and the length and 
+%% context of the studies. Many previous studies focused on 
+%% shorter time frames (mostly hours, but in one case also up to five 
+%% days~\cite{sano2013stress} 
+%% or two weeks~\cite{zuger18}). To the best of the authors' knowledge, no 
+%% other study has examined such an extensive period (eight weeks) in a real-
+%% world office environment and the inspected variables (stress, awakeness, and focus).
 
@@
 
-\subsubsection{Awakeness}
-
-Sleepiness (lack of awakeness) and its associated risk of serious injury to passengers has been studied in the context of automobile 
-accidents~\cite{connor02,Nordbakke07}.
-These studies show a strong association between the level of acute driver sleepiness and the risk of injury crash.
-Connor et al.~\cite{connor02} conducted a population-based case study using the Stanford sleepiness scale, which is similar to a seven-point Likert scale and describes seven different levels of sleepiness from ``Could not stay awake, sleep onset was imminent'' (1) to ``Felt active, wide awake'' (7).
-Nordbakke and Sagberg~\cite{Nordbakke07} show that drivers are well aware of  various factors influencing the risk of falling asleep while driving.
-Drivers also have good knowledge of the most effective measures to prevent falling asleep at the wheel.
-However, most of drivers continue driving even when recognizing sleepiness signals, due to the desire to arrive at a reasonable time, the length of the drive, or pre-planed commitments.
-
 %Bader~\cite{Bader95} proposed an approach towards monitoring and estimating a person's awakeness by leveraging on a stationary pressure sensor to track the person's body movements and a correlator to produce a change signal which is compared with a preset wakefulness threshold in a threshold detector circuit.
@@
 
+\subsubsection{Awakeness}
+
+Sleepiness (lack of awakeness) and its associated risk of serious injury to passengers has been studied in the context of automobile 
+accidents~\cite{connor02,Nordbakke07}.
+These studies show a strong association between the level of acute driver sleepiness and the risk of injury crash.
+Connor et al.~\cite{connor02} conducted a population-based case study using the Stanford sleepiness scale, which is similar to a seven-point Likert scale and describes seven different levels of sleepiness from ``Could not stay awake, sleep onset was imminent'' (1) to ``Felt active, wide awake'' (7).
+Nordbakke and Sagberg~\cite{Nordbakke07} show that drivers are well aware of  various factors influencing the risk of falling asleep while driving.
+Drivers also have good knowledge of the most effective measures to prevent falling asleep at the wheel.
+However, most of drivers continue driving even when recognizing sleepiness signals, due to the desire to arrive at a reasonable time, the length of the drive, or pre-planed commitments.
+
 \subsection{Biometrics}

### 2020-02-03 UbiCompSubmission/bp_results_1.tex
Adding a new folder UbiCompSubmission with ACM Large format for Ubicomp

--- 2019-09-20 bp_results_1.tex
@@
 \subsection{Explaining Stress Fluctuations}
-We created a linear mixed model with participants as random effects
-to attempt to explain some of the
-stress that our participants were experiencing. We experimented with day of the week and proximity to
-beginning or end of month as possible explanatory variables. We also calculated
-features from the gathered computer interaction data where available - namely total time spent actively on the computer,
-percentage of time idle, and amount of time spent on non work-related
-web browsing. However,
-ultimately we were unable to find any variable which had
-significant (p < 0.05) explanatory power with respect to our participants perceived stress
-levels. This points to the need for additional instrumentation if we
-are to succesfully understand and predict stress in the workplace.
+In an attempt to explain some of the stress that our participants were experiencing, we created a linear mixed model with the self-reported daily stress levels as dependent variable and the participants as random effects.
+We experimented with day of the week and proximity to beginning or end of month as possible explanatory variables.
+We also calculated daily features from the gathered computer interaction data where available---namely total time spent actively on the computer, percentage of time idle, and amount of time spent on non work-related web browsing per day.
+Ultimately, the analysis showed that none of the variables that we examine had a significant explanatory power with respect to our participants perceived stress levels.
+This indicates that aspects such as the time spent on a computer or the non work-related web browsing do not have a significant impact on the stress level.
+Furthermore, this points to the need for additional instrumentation and data collection, if we are to succesfully understand and predict stress in the workplace.
 

### 2020-02-03 UbiCompSubmission/bp_studyMethod.tex
Adding a new folder UbiCompSubmission with ACM Large format for Ubicomp

--- 2019-09-20 bp_studyMethod.tex
@@
 \subsubsection{Computer Interaction Data}
-To gain a better understanding of our participants day-to-day work activities, we installed an open source computer interaction monitor (reference omitted for doubleblin.).
+To gain a better understanding of our participants day-to-day work activities, we installed an open source computer interaction monitor (reference omitted for doubleblind).
 The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity.

### 2020-02-04 UbiCompSubmission/bp_discussion.tex
addressing rebuttal bullets

--- 2019-09-20 bp_discussion.tex
@@
 preferences and productivity (e.g., preferences of men and women in
-temperature~\cite{Karjalainen07}). Further research would be needed to
+temperature~\cite{Karjalainen07}).
+We believe that our results present a good step to more in situ usage of biometric sensing.
+Future studies can build on the evidence, and for example, reduce the effort required for the experience sampling from continuously rating stress levels to using biometrics as a ground truth with occasional validation of the predicted stress levels, or
 identify how to balance needs across a group of office workers and how
@@
 \subsection{The Effect of Tai Chi on Stress}
-As described in our study procedure, we offered and asked participants to attend a one-hour Tai Chi class per week for the last four weeks of the study.
-We recorded each participant's attendance, including which day they attended the offered classes.
+As described in our study procedure,
+due to our focus on stress, we offered participants an opportunity to be part of Thai Chi classes.
 The two primary reasons for this intervention was to keep participants motivated to continue the self-reports over the long study period, and to offer a technique that might help reduce stress.
+After consulting with other researchers, we decided to offer Tai Chi in the last month of the study, leaving the first month of the study unaltered.
+While participants were not required to take a Tai Chi class, they all took one per week except for two participants that opted out of the last two weeks of classes.
+Four participants explicitly stated that they liked the Tai Chi sessions or that they were “excellent”.
+These classes were performed once per week with a duration of one hour.
 

### 2020-02-04 UbiCompSubmission/bp_intro_v4.tex
addressing rebuttal bullets

--- 2019-09-20 bp_intro_v3.tex
@@
 hours~\cite{gonzalez2004constant,mark2008cost,czerwinski04diary}. 
-Consequently, stress in the workplace is an ever-growing concern. It leads to 
+These challenges, amongst others, can lead to stress in the workplace.
+Stress is an ever-growing concern as it can lead to
 fatigue, burnout and various other illnesses, ultimately resulting in work 
@@
 losses~\cite{hockey1997stress,setz2010stress,wrs2010}.
-But while the effects of stress are obvious the causes can be more subtle. Stress can be caused by
-a multitude of factors within the office including but not limited to workload, management issues, workplace harassment and bullying,
-constant interruptions, and noisy environments. It could also be the result of external stressors seeping into the workplace.
-Regardless of the cause, garnering an understanding of what stress looks like in the workplace could
-prove invaluable to helping us develop effective interventions.
-Likewise, being able to continuously measure stress in the workplace bears huge potential for better supporting 
-knowledge workers in combatting workplace challenges, and improving productivity 
-and well-being.
 
-At the same time, these 
-challenges that contribute to stress levels can make it hard for knowledge workers to stay focused, impact interaction with their digital environment, degrade the quality of work performed and hinder productivity.
-For instance, the continuous disruptions that knowledge workers face in office work environments can lead to a higher error rate and lower 
-performance~\cite{bailey2001effects,mark2008cost}.
+Given the importance of understanding stress and its relationship and
+effect on work, a number of studies have been conducted using a
+variety of methods. Some have studied stress using invasive
+techniques, such as measuring cortisol as a stress biomarker~\cite{piazza10} which are best suited
+for laboratory environments.  Further research has shown the benefit of using
+autonomic nervous system (ANS) activity by analyzing biometric signals
+of the human body, such as blood pressure~\cite{kataoka00}.
+These less invasive approaches can be used in longer-term field
+studies, opening up the question of what stress
+looks like in the wild.
 
-Fortunately, with recent advances in sensing technologies there is increasing 
-hope that we can accurately and automatically measure stress and it's related aspects. A 
-number of HCI studies have discussed and investigated the use of biometrics to 
-measure aspects such as a person's cognitive state 
-%This change is regarding Reviewer 2's concern: what are the "aspects"?
-(e.g., processing load)~\cite{beatty82,Kramer90,richter1998psychophysiological,Rowe98,Wilson2002}, 
-and more specifically to measure stress~\cite{dishman2000stress,Setz2010}, 
-or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004emotion}. Similarly,
-studies have examined the occurrence, cause and effect of similar 
-aspects in various contexts~\cite{mark2014bored,mark2008cost}, how these can be measured~\cite{Hernandez14}, and 
- how they might be modulated/regulated~\cite{Snyder15}. 
-While previous work mostly focused on smaller or controlled lab studies, our study extends this work by performing a field study to examine whether and how these three variables ---stress, focus and awakeness--- can be measured over an extended period of time in a knowledge worker's context,
-%Yet, very few 
-%studies measure productivity-related factors in the workplace over an 
-%extended period of time
-and little is known about measuring the more 
-abstract concept of focus (i.e., directed attention). 
-% Added regarding concern of Reviewer 2: Define what is meant by focus when it's first introduced. 
-Given the importance of focus, described as a 
-combination of engagement and challenge in a work task by Mark et al.~\cite{mark2014bored}, to an office workers' productivity, this factor also 
-deserves further study.
+In this paper, we build on this ability to non-invasively monitor stress
+to investigate two questions: RQ1) how do knowledge
+workers at work experience stress over time, and RQ2) can we predict
+whether a knowledge worker is experiencing stress in the moment based
+on biometric data. A better understanding of how stress is experienced (RQ1)
+can help inform the design of workplaces to manage
+and alleviate stress. An ability to predict stress in the moment (RQ2)
+can enable the development of digital tools to avoid stress.
 
-In this research we examine what office stress actually looks like over long periods of time, and develop continuous (and automatic) 
-measures of stress, focus, and awakeness in the workplace. These measures are targeted towards improving knowledge workers' productivity and 
-well-being in the future by customizing interaction with their environment; e.g., by continuously monitoring stress levels and recommending stress-reducing interventions during 
-extended periods of high stress, automatically adjusting lighting and temperature to heighten awakeness, or improving management 
-of disruptions~\cite{iqbal2005effectiveness,zuger2017reducing} to aid with focus and productivity.
+Our work builds upon previous work and extends it by performing an
+eight-week study in the workplace with 14 knowledge 
+workers. To the best of our knowledge, this is the 
+longest in-situ study performed in a real-world 
+environment of knowledge workers to analyze an extensive collection of biometric signals with experience samples examining real-time prediction of stress in the workplace~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}.
+Previous studies are predominantly controlled lab studies which used only a subset of the relevant biometric signals we captured, have been much more limited in duration, or are not focused on the workplace.
 
-Our work builds upon previous work and extends it by performing an eight-week 
-study in the workplace with 14 knowledge workers, collecting biometric and 
-computer interaction data as well as experience samples on focus, awakeness, 
-and stress. Our participants, who perform various job functions for a research 
-and development group within a single large corporation, wore a single 
-biometric armband sensor\footnote{Biovotion Everion sensor~\cite{everion}} 
-with low invasiveness that captured heart-, respiration- and skin-related measures. This 
-modality was chosen to ease longitudinal deployment in the field. Using 
-machine learning, we created classifiers and analyzed their ability to predict stress, focus, and awakeness levels.
-%Similarly, we captured computer 
-%interaction features from participants' everyday computer usage and
-%analyzed how a model trained using the combination
-%of these two training sets compares to the performance of models using each
-%set individually.
+We collected biometric data as well as
+experience samples on stress, and two related human aspects: focus and
+awakeness. We collected participant reports on focus because the
+challenges that contribute to stress levels can also make it hard for
+knowledge workers to stay focused and the level of focus
+can affect
+productivity~\cite{mark2014bored}.  We collected participant reports
+on awakeness because a lack of awakeness (sleepiness) can result in
+undesired consequences on work~\cite{connor02}.  The participants in
+our study, who perform various job functions for a research and
+development group within a single large corporation, wore a novel, state-of-the-art
+biometric armband sensor\footnote{Biovotion Everion
+  sensor~\cite{everion}} with low invasiveness that captured heart-,
+respiration- and skin-related measures. This modality was chosen to
+ease longitudinal deployment in the field. Using machine learning, we
+created classifiers and analyzed their ability to predict stress, focus, and awakeness levels.
+Considering all of these three human aspects helps us to consider whether one aspect might be easier to detect than another.
+If one aspect is easier to detect, it might serve
+as a proxy or an indicator of the presence (or absence) of other aspects.
 
-In our analysis we identify several trends in day-to-day stress levels that emerged over the course of our eight-week study.
-The results of our analysis also show that biometric signals collected from a single minimally invasive sensor can be used to predict stress and its related aspects accurately, with the abstract concept of focus (predictably) being the hardest to detect.
-Our results further show that knowledge workers' self-reported levels of stress, focus and awakeness and their physiological manifestation and prediction can vary a lot between individuals.
+In our analysis, we identify several trends in day-to-day stress levels
+that emerged over the course of our eight-week study.  The results of
+our analysis also show that biometric signals collected from a single
+minimally invasive sensor can be used to predict stress and its
+related aspects accurately, with the abstract concept of focus
+(predictably) being the hardest to detect. Our results further show
+that knowledge workers' self-reported levels of stress, focus and
+awakeness and their physiological manifestation and prediction can
+vary substantially between individuals.
 %Our results further show that overall combining biometric features with computer interaction information 
@@
 \begin{itemize}
-	\item An qualitative examination of how knowledge worker stress presents and fluctuates in the wild.
-	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' awakeness, stress, and focus in the workplace based on an eight-week field study with 14 office workers.
+	\item A qualitative examination of how knowledge worker stress presents and fluctuates in the wild.
+	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' stress, focus and awakeneess in the workplace based on an eight-week field study with 14 office workers.
 	%\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.

### 2020-02-04 UbiCompSubmission/bp_relatedWork.tex
addressing rebuttal bullets

@@
 In contrast to studies that use eye tracking or EEG, we focused on a less invasive technology that can be applied in a real world scenario.
+Similarly, we considered adding additional metrics such as Depression, Anxiety and Stress Scale (DASS) and Perceived Stress Scale (PSS)~\cite{Ferdous15} but opted for the ones presented in the paper after several discussions with experts in the area, and after piloting surveys to maxime participant compliance.

### 2020-02-04 UbiCompSubmission/bp_results_1.tex
addressing rebuttal bullets

@@
 Due to the larger amount of data available in the general case, we used the more common random undersampling, which randomly selects elements in the majority class to exclude from the dataset, instead of random oversampling to balance the distribution of the dataset.
-The models were trained on the datasets of 13 of the 14 participants, and then evaluated on the dataset of the last, repeating this process for all 14 participants.
+The models were trained on the datasets of 13 of the 14 participants, and then evaluated on the dataset of the last (leave-one-group-out cross-validation), repeating this process for all 14 participants.
 

### 2020-02-04 UbiCompSubmission/bp_studyMethod.tex
addressing rebuttal bullets

@@
 \subsection{Participants}
-We recruited 14 professionals via personal contacts from a large power and automation company.
+We recruited group of 14 professionals via personal contacts from a large power and automation company.
+The group is very diverse in terms of age, work experience, gender, and work responsibilities.
 All participants work primarily in an office environment, though half of the participants spend at least 10\% (and up to 50\%) of their time in a laboratory environment.

### 2020-02-06 UbiCompSubmission/bp_conclusion.tex
Extra revissions

@@
 examine the ability of biometrics to predict these
-productivity-related aspects. The longitudinal and in situ placement of
+productivity-related aspects. The longitudinal and in-situ placement of
 the study support and extend previous work. Based on
@@
 continually collected
-biometrics data with which we were able to create a model that was able to
-predict awakeness, stress, and focus.
+biometrics data with which we were able to create a model that is able to
+predict user stress, sleepyness and lack of focus with an up to 84.65\% improvement over the baseline.
 

### 2020-02-06 UbiCompSubmission/bp_intro_v4.tex
Extra revissions

@@
 effect on work, a number of studies have been conducted using a
-variety of methods. Some have studied stress using invasive
-techniques, such as measuring cortisol as a stress biomarker~\cite{piazza10} which are best suited
-for laboratory environments.  Further research has shown the benefit of using
+variety of methods. Minimally invasive studies have shown the benefit of using
 autonomic nervous system (ANS) activity by analyzing biometric signals
 of the human body, such as blood pressure~\cite{kataoka00}.
-These less invasive approaches can be used in longer-term field
+These approaches can be used in long-term field
 studies, opening up the question of what stress
 looks like in the wild. 
+In contrast, other studies have studied stress using invasive
+techniques, such as measuring cortisol as a stress biomarker~\cite{piazza10}. These studies are best suited
+for laboratory environments.
 
@@
 environment of knowledge workers to analyze an extensive collection of biometric signals with experience samples examining real-time prediction of stress in the workplace~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}.
-Previous studies are predominantly controlled lab studies which used only a subset of the relevant biometric signals we captured, have been much more limited in duration, or are not focused on the workplace.
+Previous studies are predominantly controlled lab studies which used only a subset of the relevant biometric signals we captured, have been much more limited in duration, or are not focused on a real-world environment0.
 
@@
 undesired consequences on work~\cite{connor02}.  The participants in
-our study, who perform various job functions for a research and
-development group within a single large corporation, wore a novel, state-of-the-art
+our study perform various job functions for a research and
+development group within a single large corporation. For this study they wore a novel, state-of-the-art
 biometric armband sensor\footnote{Biovotion Everion
@@
 \begin{itemize}
-	\item A qualitative examination of how knowledge worker stress presents and fluctuates in the wild.
-	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' stress, focus and awakeneess in the workplace based on an eight-week field study with 14 office workers.
+	\item A qualitative examination of how knowledge worker stress presents and fluctuates in the wild based on an eight-week field study with 14 office workers.
+	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' stress, focus and awakeneess in the workplace based on a machine learning model trained on biometric data and experience samples.
 	%\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.

### 2020-02-06 UbiCompSubmission/bp_results_1.tex
Extra revissions

@@
 To investigate whether stress, focus and awakeness can be predicted in
-the moment based on biometric measures.
+the moment based on biometric measures,
 we investigated classifiers trained
@@
 statistical measurements are: mean, standard deviation, variance, median, 
-percentile25, percentile75, interquartile range, maximum, minimum, and 
+25\textsuperscript{th} percentile, 75\textsuperscript{th} percentile, interquartile range, maximum, minimum, and 
 range. Thus, for each survey response, we generate a large number of 
@@
 (e.g.,~\cite{Hernandez11}), we first trained individual classifiers
-for each participant (rather than a general one for all
+for each participant (as opposed to a general one for all
 participants). The results of our analysis are reported in
@@
 For some users (i.e., S1.
-as seen In Table \ref{responseDistribution}), the imbalance in their data was so extreme that even after adjusting by oversampling it was impossible to create a reasonable classifier.
+as seen in Table \ref{responseDistribution}), the imbalance in their data was so extreme that even after adjusting by oversampling it was impossible to create a reasonable classifier.
 These scenarios are difficult to predict as any classifier will not have enough variance in its training data for the 'stressed' situation to adequately distinguish it from the non-stressed case.

### 2020-02-06 UbiCompSubmission/bp_studyMethod.tex
Extra revissions

@@
 We recruited group of 14 professionals via personal contacts from a large power and automation company.
-The group is very diverse in terms of age, work experience, gender, and work responsibilities.
+The group is diverse in terms of age, work experience, gender, and work responsibilities.
 All participants work primarily in an office environment, though half of the participants spend at least 10\% (and up to 50\%) of their time in a laboratory environment.
@@
 
-Of the 14 participants, 11 are male and 3 are female.
+From the 14 participants, 11 are male and 3 are female.
 The average participant age is ~40, with 5 in the age range 25-34, 7 in the age range 35-44, and 2 in the age range 55+.
-The participants have an average number of years of professional experience of ~12, with 2 having less than 5 years, 10 having 5-15 years, and 2 having more than 25 years.
+The average number of years of professional experience of the participants is ~12, with 2 having less than 5 years, 10 having 5-15 years, and 2 having more than 25 years.
 All participants work for a research organization within the company, but their job functions span line management, laboratory science, scientific research, technology evaluation, and software development.
@@
 \subsection{Procedure}
-We performed an in-situ study over the course of eight weeks.
-For this, we first informed participants about the study purpose and procedure, handed out biometric sensors and introduced and explained the self-reporting to the participants.
+We performed a study over the course of eight weeks.
+This study includes the collection of the participant's biometric data, computer interaction data, and self-reporting surveys.
+We informed participants about the study purpose and procedure, handed out biometric sensors, and introduced and explained the self-reporting to the participants.
 For 10 of the participants, we further installed a computer activity tracker on their company-issued laptops (note: four participants from the original sample declined for privacy reasons).
+
 After the initial setup, participants were asked to fill out 3 surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}).
@@
 
-In the second month of the study, we offered participants two one-hour Tai Chi classes per week (each in the middle of the day) in addition to the daily experience sampling and asked them to attend one class a week.
+In the second month of the study, we also offered participants two one-hour Tai Chi classes per week (each in the middle of the day) and asked them to attend one class a week.
 We chose to offer Tai Chi class for two reasons: first, to incentivise participants for their continued participation and for keeping them engaged; and second, for their said benefits of reducing stress based on our consultation with researchers in psychiatry, psychology and more specifically on mindfulness practices.
@@
 \begin{comment}
-\subsection{Procedure}
-\todo{This section does not work here.
+%\subsection{Procedure}
+%\todo{This section does not work here.
 It is not really the study procedure, but rather the project procedure.
@@
 
-We use state-of-the-art sensors to capture a stream of biometric data from each individual and a computer interaction tracker to collect computer interaction information during the time
-of the study.
+%We use state-of-the-art sensors to capture a stream of biometric data from each individual and a computer interaction tracker to collect computer interaction information during the time
+%of the study.
 
-The participants were asked to fill a survey 3 times a day: two during the working day, and
-one at the end of the day, where they rated their level of stress, focus, and awakeness during
-the day; and their level of stress, awakeness, productiveness, and overall feeling at the end of the day.
-This was conducted over an eight week period, which is
-400\% longer than previous studies~\cite{zuger18,Muller16}.
+%The participants were asked to fill a survey 3 times a day: two during the working day, and
+%one at the end of the day, where they rated their level of stress, focus, and awakeness during
+%the day; and their level of stress, awakeness, productiveness, and overall feeling at the end of the day.
+%This was conducted over an eight week period, which is
+%400\% longer than previous studies~\cite{zuger18,Muller16}.
 
-We then pick a set of time windows (10sec, 20sec, 30sec, 45sec, 1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, 3hour) to analyze.
+%We then pick a set of time windows (10sec, 20sec, 30sec, 45sec, 1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, 3hour) to analyze.
 We applied different statistical measurements (mean, standard deviation, variance, median, percentile25, percentile75, interquartile range, maximum, minimum, range) to each of the collected biometric signals and the computer interaction data within the time windows selected.
 We then
-compared the performance of four different machine learning 
-algorithms to be able to predict the survey responses based in the biometric and 
-computer interaction data.
+%compared the performance of four different machine learning 
+%algorithms to be able to predict the survey responses based in the biometric and 
+%computer interaction data.
 
-Based in the initial corpus of training data, we create a machine learning model that is able to accurately predict the responses for stress, focus, and awakeness for each of the individuals wearing the sensors.
-We are therefore able to infer the answer within the spectrum to each of these indicators without the need to ask the participants to provide an answer.
+%Based in the initial corpus of training data, we create a machine learning model that is able to accurately predict the responses for stress, focus, and awakeness for each of the individuals wearing the sensors.
+%We are therefore able to infer the answer within the spectrum to each of these indicators without the need to ask the participants to provide an answer.
 \end{comment}
@@
 \subsubsection{Computer Interaction Data}
-To gain a better understanding of our participants day-to-day work activities, we installed an open source computer interaction monitor (reference omitted for doubleblind).
+To gain a better understanding of our participants day-to-day work activities, we installed an open source computer interaction monitor~\footnote{reference omitted for double-blind review} \todo{One of the conditions to submit to Ubicomp is to NOT omit references due to anonymization purposes.
+So either we add the reference (which will be fine as long as we don't say we made it, or completely remove this reference)}.
 The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity.
 Four of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13).
-Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.
+%Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.
 
@@
 text message a survey request to each participant two times per 
-work day. Pilot participants preferred these over other means, in part, due 
+work day. Pilot participants preferred the usage of text message, in part, due 
 to them being accessible and noticeable anywhere in the office.

### 2020-02-14 UbiCompSubmission/bp_studyMethod.tex
update with gail's changes

@@
 
-In the second month of the study, we also offered participants two one-hour Tai Chi classes per week (each in the middle of the day) and asked them to attend one class a week.
-We chose to offer Tai Chi class for two reasons: first, to incentivise participants for their continued participation and for keeping them engaged; and second, for their said benefits of reducing stress based on our consultation with researchers in psychiatry, psychology and more specifically on mindfulness practices.
+Keeping participants engaged in an eight-week study can be
+difficult. In the second month of the study, largely to incent participants to
+continue, we offered participants two one-hour Tai Chi classes per
+week (each in the middle of the day) and asked them to attend one
+class a week. The choice of Tai Chi provided a link to mindfulness,
+a topic of growing interest in communities. Consultations we held
+with researchers in psychiatry and psychology suggested Tai Chi as
+a good link to mindfulness and an intervention that might alleviate
+stress. By recording attendence of participants at the Tai Chi
+sessions, we are able to analyze the impact of the sessions
+on the results. As a brief summary, the impact of the
+sessions was minimal; a a result, we present an analysis
+of the effects in the Discussion section of the paper.
 

### 2020-02-14 UbiCompSubmission/bp_conclusion.tex
Submission to IMWUT/UbiComp 2020

@@
 \section{Conclusions} 
-Stress, awakeness and focus at work are highly relevant aspects when
+Stress, awakeness, and focus at work are highly relevant aspects when
 it comes to productivity and well-being at the workplace. In this
@@
 continually collected
-biometrics data with which we were able to create a model that is able to
-predict user stress, sleepyness and lack of focus with an up to 84.65\% improvement over the baseline.
+biometric data with which we were able to create a model that is able to
+predict user stress, sleepyness, and lack of focus with an up to 84.65\% improvement over the baseline.
 

### 2020-02-14 UbiCompSubmission/bp_studyMethod.tex
Submission to IMWUT/UbiComp 2020

@@
 \subsubsection{Computer Interaction Data}
-To gain a better understanding of our participants day-to-day work activities, we installed an open source computer interaction monitor~\footnote{reference omitted for double-blind review} \todo{One of the conditions to submit to Ubicomp is to NOT omit references due to anonymization purposes.
-So either we add the reference (which will be fine as long as we don't say we made it, or completely remove this reference)}.
+To gain a better understanding of our participants day-to-day work activities, we used an open source computer interaction monitor~\cite{Meyer17}.
 The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity.

### 2020-03-25 HICSSSubmission/bp_intro_v4.tex
adding folder for HICSS submission

--- 2020-02-06 UbiCompSubmission/bp_intro_v4.tex
@@
 longest in-situ study performed in a real-world 
-environment of knowledge workers to analyze an extensive collection of biometric signals with experience samples examining real-time prediction of stress in the workplace~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}.
-Previous studies are predominantly controlled lab studies which used only a subset of the relevant biometric signals we captured, have been much more limited in duration, or are not focused on a real-world environment0.
+environment of knowledge workers analyzing an extensive collection of biometric signals with experience samples examining real-time prediction of stress in the workplace~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}.
+Previous studies are predominantly controlled lab studies which used only a subset of the relevant biometric signals we captured, much more limited in duration, or not focused on a real-world environment.
 
@@
 our study perform various job functions for a research and
-development group within a single large corporation. For this study they wore a novel, state-of-the-art
+development group within a single large corporation. For this study, they wore a novel, state-of-the-art
 biometric armband sensor\footnote{Biovotion Everion
@@
 \begin{itemize}
+	\vspace{-3mm}
 	\item A qualitative examination of how knowledge worker stress presents and fluctuates in the wild based on an eight-week field study with 14 office workers.
+	\vspace{-3mm}
 	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' stress, focus and awakeneess in the workplace based on a machine learning model trained on biometric data and experience samples.
 	%\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
+	\vspace{-3mm}
 	\item A discussion on the impact of applying this research to improve

### 2020-03-25 HICSSSubmission/bp_relatedWork.tex
adding folder for HICSS submission

--- 2020-02-04 UbiCompSubmission/bp_relatedWork.tex
@@
 
-The study and prediction approach on which we report in this paper is related
+The study and prediction approach used in this paper is related
 to previous studies of productivity-related indicators and studies
-using biometerics to predict these indicators. We consider
+using biometerics to predict them. We consider
 related work in each of these categories in turn.
@@
 
+\vspace{-8mm}
+
 \subsubsection{Stress}
+
 Much previous work relates to identifying and mitigating stress in an office environment.
@@
 
+\vspace{-8mm}
+
 \subsubsection{Focus}
+
 Focus refers to the allocation of limited cognitive processing resources~\cite{Anderson04}.
@@
 Flow occurs when a person is focused on an activity that requires high challenge and high use of the person's skills, whereas mindfulness is characterized by being aware of fine detail, affording the capacity to discover and manage unexpected events.
+
+\vspace{-8mm}
 

### 2020-03-26 HICSSSubmission/bp_studyMethod.tex
moved related work to the end, still need to cut 3 pages

--- 2020-02-14 UbiCompSubmission/bp_studyMethod.tex
@@
 \subsection{Participants}
-We recruited group of 14 professionals via personal contacts from a large power and automation company.
+We recruited a group of 14 professionals via personal contacts from a large power and automation company.
 The group is diverse in terms of age, work experience, gender, and work responsibilities.
@@
 Figure~\ref{everion} illustrates Biovotion's Everion, which we used to track the biometric signals of the study participants.
-The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/categories/201633909-Everion-Device} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements for shorter periods of time or capturing a smaller number of measurements.
+The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/articles/213613165} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements for shorter periods of time or capturing a smaller number of measurements.
 

### 2020-03-26 HICSSSubmission/bp_discussion.tex
changed a couple of pics to png because it was messing the document, and slashed a big part of related work, and added spacing so the hicss template works

--- 2020-02-04 UbiCompSubmission/bp_discussion.tex
@@
 \subsection{Implications for Workplaces}
+
 Being able to accurately recognize periods of high-stress in knowledge
@@
 \subsection{The Effect of Tai Chi on Stress}
+
 As described in our study procedure,
@@
 While participants were not required to take a Tai Chi class, they all took one per week except for two participants that opted out of the last two weeks of classes.
-Four participants explicitly stated that they liked the Tai Chi sessions or that they were “excellent”.
+Four participants explicitly stated that they liked the Tai Chi sessions or that they were ``excellent''.
 These classes were performed once per week with a duration of one hour.
@@
 \subsection{Ground Truth and Self-Reporting}
+
 Studying mental states, such as stress, awakeness and focus,
@@
 highly individual nature of reporting aspects such as
-stress~\cite{Hernandez11,Hovsepian15}. In addition, in contexts such
+stress~\cite{Hernandez11,Hovsepian15}.
+
+In addition, in contexts such
 as the workplace, as in our study,
@@
 \subsection{Imbalanced Data}
+
 Study participants provided highly
@@
 \subsection{Predicting Stress with Computer Interaction Data}
+
 Knowledge workers, a focus of our work and study, often spend

### 2020-03-26 HICSSSubmission/bp_relatedWork.tex
changed a couple of pics to png because it was messing the document, and slashed a big part of related work, and added spacing so the hicss template works

@@
 to previous studies of productivity-related indicators and studies
-using biometerics to predict them. We consider
-related work in each of these categories in turn.
+using biometrics to predict them.
 
@@
 
-Hovsepian et al.~\cite{Hovsepian15} have worked to obtain a standard for continuous stress assessment.
-They used sensors to conduct a seven-day lab study with 26 participants, as well as a field study with 20 participants.
-They found that their model showed significant improvement over simple heart rate variability measurements.
-However, this model requires the use of a suite of invasive sensors that would be impractical for a study the length of ours.
-
+Hovsepian et al.~\cite{Hovsepian15} have worked to obtain a standard for continuous stress assessment. Their model showed significant improvement over simple heart rate variability measurements.
 Hernandez et al.~\cite{Hernandez14} investigated the use of pressure-sensitive keyboard and a capacitative mouse as non-intrusive means for measuring computer users stress levels.
@@
 They found that workers exposed to open-office noise showed aftereffects that indicate motivational deficits.
-The population for their experiment comprised 40 female clerical workers, who were randomly assigned to a control condition or to three-hour low-intensity noise room designed to simulate typical open-office noise levels.
-
 Stress in the workplace and its effects on service providers has been analyzed in call centers~\cite{Hernandez11} and in the context of the perceived imbalance between resources and demands~\cite{cherniss80}.
+Kocielnik et al.~\cite{kocielnik_smart_2013} developed a framework for unobtrusive and continuous measurement of stress in real life conditions.
+They observed that the data they collected reflected well the participants perceptions of their own stress levels.
 These studies considered several factors such as personality traits, career-related goals and attitudes, and life outside of work, and these factors were correlated with stress levels and burnout.
-
-Kocielnik et al.~\cite{kocielnik_smart_2013} developed a framework for unobtrusive and continuous measurement of stress in real life conditions.
-They equipped university staff members with a wristband sensor and combined this data with information from the participants calendars over the course of four weeks.
-They observed that the data they collected reflected well the participants perceptions of their own stress levels.
-However this work focused mostly on providing retrospective information to users so they can work to improve their own stress balance.
-They did not attempt to make observations about the big picture of participants stress profiles or make predictions in real time.
 
@@
 Mark et al.~\cite{mark2014bored} studied engagement in workplace activities by analyzing the digital activity of 32 information workers in situ for 5 days to understand how attentional states change with context.
-They found that boredom is highest in the early afternoon and focus peaks in the middle of the afternoon.
-They also found that doing work that requires focus correlates with stress, while rote work correlates with happiness.
+They found that doing work that requires focus correlates with stress, while rote work correlates with happiness.
 
-Interruptions in the office are a common barrier keeping workers from sustaining focus on their work related activities, particularly when the interruptions occur at inopportune moments.
-Such interruptions may include emails, alerts, or interactions with co-workers\cite{gonzalez2004constant,chong2006interruptions,shamsi07}.
+Interruptions in the office are a common barrier keeping workers from sustaining focus including emails, alerts, or interactions with co-workers\cite{gonzalez2004constant,chong2006interruptions,shamsi07}.
 Interruptions in inopportune times can have negative effects that range from higher error rate and lower overall performance
@@
 
-Other studied constructs that relate to focus in the workplace include cognitive absorption, cognitive engagement, flow, and mindfulness.
 Cognitive absorption describes periods of time in which a person experiences total immersion in an activity.
-This state is also accompanied by a sense of deep enjoyment, a feeling of control, curiosity, and not realizing the passing of time.
 It has been associated with ease of use and perceived usefulness of information technology~\cite{agarwal00}.
@@
 Flow~\cite{Csikszentmihalyi90}, and mindfulness~\cite{Weick06,dane11} are psychological states that describe periods of prolonged attention and total immersion in an activity.
-Flow occurs when a person is focused on an activity that requires high challenge and high use of the person's skills, whereas mindfulness is characterized by being aware of fine detail, affording the capacity to discover and manage unexpected events.
 
@@
 
-Sleepiness (lack of awakeness) and its associated risk of serious injury to passengers has been studied in the context of automobile 
+Sleepiness (lack of awakeness) has been studied in the context of automobile 
 accidents~\cite{connor02,Nordbakke07}.
 These studies show a strong association between the level of acute driver sleepiness and the risk of injury crash.
-Connor et al.~\cite{connor02} conducted a population-based case study using the Stanford sleepiness scale, which is similar to a seven-point Likert scale and describes seven different levels of sleepiness from ``Could not stay awake, sleep onset was imminent'' (1) to ``Felt active, wide awake'' (7).
+Connor et al.~\cite{connor02} measure sleepiness using the Stanford sleepiness scale, similar to the seven-point Likert scale used in this study.
+
 Nordbakke and Sagberg~\cite{Nordbakke07} show that drivers are well aware of  various factors influencing the risk of falling asleep while driving.
 Drivers also have good knowledge of the most effective measures to prevent falling asleep at the wheel.
-However, most of drivers continue driving even when recognizing sleepiness signals, due to the desire to arrive at a reasonable time, the length of the drive, or pre-planed commitments.
+However, most of drivers continue driving even when recognizing sleepiness signals.
 
 \subsection{Biometrics}
-Our study investigates the prediction of multiple productivity-related indicators (stress, focus, and awakeness) using two different types of measurements, biometric signals and computer interaction, over eight weeks in a real-life office setting.
+
 Existing work~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015} analyzes a broad array of biometric signals and correlates them with individual's cognitive states and processes.
-For example, Zuger et al.~\cite{zuger18} used biometrics to sense interruptibility in the office~\cite{zuger18}.
-%and found correlations between biometric signals and predicting code quality~\cite{Muller16}, to assess task difficulty in software development~\cite{fritz2014using}, to assess developer's mental load and perceived difficulty while working on small code snippets, and
+Zuger et al.~\cite{zuger18} used biometrics to sense interruptibility in the office~\cite{zuger18}.
 Biometric signals have also been studied in the context of technology users.
-For example, Parnin~\cite{Parnin2011} analyzes electromyography to measure sub-vocal utterances, and how these might be correlated with the programmer's perceived difficulty of programming tasks.
-Similarly, biometrics have been used to measure code difficulty by using biometric sensors~\cite{fritz2014using}
+Parnin~\cite{Parnin2011} analyzes electromyography to measure sub-vocal utterances, and how these might be correlated with the programmer's perceived difficulty of programming tasks.
+Similarly, biometrics have been used to measure code difficulty~\cite{fritz2014using}
 and using Near Infrared Spectroscopy to measure developer's cerebral blood flow~\cite{Nakagawa2014}.
@@
 Eye tracking technology~\cite{Bednarik2006,Crosby1990,Rodeghero2014} and brain activity~\cite{Ikutani2014,Siegmund2014} 
-have been used in previous studies to analyze different tasks in an office environment.
+have been used in previous studies to analyze tasks in an office environment.
 Eye tracking has been used to analyze memory load and processing load by inspecting task-evoked pupillary response and pupil size~\cite{beatty82}.
@@
 attentional demand and working memory load~\cite{Smith05,sterman93}.
+
 In contrast to studies that use eye tracking or EEG, we focused on a less invasive technology that can be applied in a real world scenario.
-Similarly, we considered adding additional metrics such as Depression, Anxiety and Stress Scale (DASS) and Perceived Stress Scale (PSS)~\cite{Ferdous15} but opted for the ones presented in the paper after several discussions with experts in the area, and after piloting surveys to maxime participant compliance.
+Similarly, we considered adding additional metrics such as Depression, Anxiety and Stress Scale (DASS) and Perceived Stress Scale (PSS)~\cite{Ferdous15} but opted for the ones presented in the paper after several discussions with experts in the area, and after piloting surveys to maximize participant compliance.

### 2020-03-26 HICSSSubmission/bp_results_1.tex
changed a couple of pics to png because it was messing the document, and slashed a big part of related work, and added spacing so the hicss template works

--- 2020-02-06 UbiCompSubmission/bp_results_1.tex
@@
 
-\section{Observed Trends Over Time in Stress Levels}
+\section{Observed Trends in Stress Levels}
 \label{stressTrends}
@@
 \subsection{Baseline Stress Levels}
+
 Common amongst all participants was a trend to select one stress rating
@@
   \centering
-      \includegraphics[width=0.5\textwidth]{dailystress_s1.pdf}
-      \includegraphics[width=0.5\textwidth]{dailystress_s3.pdf}
+      \includegraphics[width=0.4\textwidth]{dailystress_s1.png}
+      \includegraphics[width=0.4\textwidth]{dailystress_s3.png}
   \caption{The day-to-day stress levels reported by two participants (S1 and S3) are shown.
@@
 The x-axes are slightly different for the 2 participants as they did not begin and end the study on exactly the same days.}
+   \label{fig:dailyStress}
    \vspace*{-2mm}
-   \label{fig:dailyStress}
 \end{figure}
@@
 \subsection{Stressful Days Tend to Cluster}
+
 Accounting for the variance between participants perceived stress
@@
 \subsection{Extreme Changes in Stress Levels are Rare}
+
 After accounting for each participant's perceived stress baseline, we
@@
 \subsection{Explaining Stress Fluctuations}
+
 In an attempt to explain some of the stress that our participants were experiencing, we created a linear mixed model with the self-reported daily stress levels as dependent variable and the participants as random effects.
@@
 \subsubsection{Feature Extraction}
+
 We extracted features from the biometric data to provide as input to machine 
@@
 \subsubsection{Response Transformations}
+
 Table~\ref{responseDistribution} illustrates the distribution of responses from each participant for each of the three survey questions (listed in Section~\ref{sec:Surveys}).
@@
 \subsubsection{Oversampling}
+
 Even after binarizing the responses as described in the previous section, we found the distribution of responses was still quite imbalanced for many of our participants.
@@
 \subsection{Individual Classifiers}
+
 %\noindent\textit{Results of Individual Classifiers}\\
@@
 \subsection{Feature Selection and Importance}
+
 %\noindent\textit{Feature Selection and Importance}\\
@@
 \subsection{Individual vs.\ General Model}
+
 %\noindent\textit{Individual vs.\ General Model}\\

### 2020-03-26 HICSSSubmission/bp_studyMethod.tex
changed a couple of pics to png because it was messing the document, and slashed a big part of related work, and added spacing so the hicss template works

@@
 \section{Field Study}
+
 We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress over time and the feasibility of predicting stress, focus, and awakeness based on biometric signals.
@@
 \subsection{Participants}
+
 We recruited a group of 14 professionals via personal contacts from a large power and automation company.
@@
 \subsection{Procedure}
+
 We performed a study over the course of eight weeks.
@@
 \subsubsection{Biometric Sensors}
+
 Figure~\ref{everion} illustrates Biovotion's Everion, which we used to track the biometric signals of the study participants.
@@
 \subsubsection{Computer Interaction Data}
+
 To gain a better understanding of our participants day-to-day work activities, we used an open source computer interaction monitor~\cite{Meyer17}.
@@
 \label{sec:Surveys}
+
 Following guidelines from previous studies~\cite{Lalle16,Panwar18,Luo18} and 

### 2020-03-26 HICSSSubmission/bp_threats.tex
adding folder for HICSS submission
changed a couple of pics to png because it was messing the document, and slashed a big part of related work, and added spacing so the hicss template works

--- 2019-09-20 bp_threats.tex
@@
 \section{Threats to Validity}
+
 There are numerous threats to validity to our study.
@@
 we collected a data for an eight week period, which is
-400\% longer than previous studies~\cite{zuger18,Muller16}.
+400\% longer than the longest previous studies~\cite{zuger18,Muller16}.
 

### 2020-03-26 HICSSSubmission/bp_discussion.tex
reducing text

@@
 
+\vspace{-4mm}
 Humans experience stress, focus and awakeness in different ways.  In
@@
 
+\vspace{-4mm}
 Being able to accurately recognize periods of high-stress in knowledge
@@
 
+\vspace{-4mm}
 As described in our study procedure,
@@
 
+\vspace{-4mm}
 Studying mental states, such as stress, awakeness and focus,
@@
 
+\vspace{-4mm}
 Study participants provided highly
@@
 
+\vspace{-4mm}
 Knowledge workers, a focus of our work and study, often spend

### 2020-03-26 HICSSSubmission/bp_intro_v4.tex
moved related work to the end, still need to cut 3 pages
reducing text

@@
 \section{Introduction}
+
+\vspace{-4mm}
 \textit{`The most valuable asset of a 21st-century institution (whether 
@@
 experience samples on stress, and two related human aspects: focus and
-awakeness. We collected participant reports on focus because the
+awakeness (i.e. wakefulness). We collected participant reports on focus because the
 challenges that contribute to stress levels can also make it hard for
@@
 The main contributions of our work are:
-\begin{itemize}
-	\vspace{-3mm}
+\begin{itemize}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
 	\item A qualitative examination of how knowledge worker stress presents and fluctuates in the wild based on an eight-week field study with 14 office workers.
-	\vspace{-3mm}
 	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' stress, focus and awakeneess in the workplace based on a machine learning model trained on biometric data and experience samples.
 	%\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
-	\vspace{-3mm}
 	\item A discussion on the impact of applying this research to improve

### 2020-03-26 HICSSSubmission/bp_relatedWork.tex
reducing text

@@
 
+\vspace{-4mm}
 The study and prediction approach used in this paper is related
@@
 
+\vspace{-4mm}
 Our study focuses on the productivity-related indicators of stress, focus and awakeness.
@@
 
-\vspace{-8mm}
-
 \subsubsection{Stress}
 
+\vspace{-8mm}
 Much previous work relates to identifying and mitigating stress in an office environment.
@@
 
-\vspace{-8mm}
-
 \subsubsection{Focus}
 
+\vspace{-8mm}
 Focus refers to the allocation of limited cognitive processing resources~\cite{Anderson04}.
@@
 
-\vspace{-8mm}
-
 \subsubsection{Awakeness}
 
+\vspace{-8mm}
 Sleepiness (lack of awakeness) has been studied in the context of automobile 
@@
 
+\vspace{-4mm}
 Existing work~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015} analyzes a broad array of biometric signals and correlates them with individual's cognitive states and processes.

### 2020-03-26 HICSSSubmission/bp_results_1.tex
reducing text

@@
 
+\vspace{-4mm}
 To gain insights into how knowledge workers experience stress over an
@@
 
-Common amongst all participants was a trend to select one stress rating
+\vspace{-4mm}
+Common among all participants was a trend to select one stress rating
 far more frequently than any other. We will refer to this value as the
@@
       \includegraphics[width=0.4\textwidth]{dailystress_s3.png}
-  \caption{The day-to-day stress levels reported by two participants (S1 and S3) are shown.
-The y-axis represents values on the 5-point Likert scale we asked participants to respond with ranging from 1/Not at all stressed to 5/Extremely stressed.
-The x-axis represents the day of the study on which the value was recorded (from 0-45).
-The x-axes are slightly different for the 2 participants as they did not begin and end the study on exactly the same days.}
+  \caption{The day-to-day stress levels reported by two participants. Y-axis represents values on the 5-point Likert scale. The x-axis represents the day of the study.}
    \label{fig:dailyStress}
@@
 
+\vspace{-4mm}
 Accounting for the variance between participants perceived stress
@@
 
-\subsection{Extreme Changes in Stress Levels are Rare}
+\subsection{Extreme Changes are Rare}
 
+\vspace{-4mm}
 After accounting for each participant's perceived stress baseline, we
@@
 participants. This suggests that some people may be less resilient to
-the stress of the workplace than others. For most participants,
+the stress of the workplace. For most participants,
 extremely stressful days were few and far between.
@@
 
+\vspace{-4mm}
 In an attempt to explain some of the stress that our participants were experiencing, we created a linear mixed model with the self-reported daily stress levels as dependent variable and the participants as random effects.
@@
 This indicates that aspects such as the time spent on a computer or the non work-related web browsing do not have a significant impact on the stress level.
-Furthermore, this points to the need for additional instrumentation and data collection, if we are to succesfully understand and predict stress in the workplace.
+Furthermore, this points to the need for additional instrumentation and data collection, if we are to successfully understand and predict stress in the workplace.
 
@@
 
+\vspace{-4mm}
 To investigate whether stress, focus and awakeness can be predicted in
@@
 
+\vspace{-4mm}
 In this section we describe how we preprocessed the collected data for use in training and testing machine learning models.
@@
 
+\vspace{-8mm}
 We linked the collected biometric data and survey responses for each participant.
@@
 Reasons for a survey response to lack associated biometric data include:
-\begin{itemize}
+\begin{itemize}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
 \item The participant not wearing the Everion in the hour before before beginning the survey
@@
       \includegraphics[width=0.5\textwidth]{DuringTheDay.png}
-  \caption{The figure shows the biometric data available per each participant. Green sections represent survey responses with biometric
+  \caption{Biometric data available per each participant. Green sections represent survey responses with biometric
   data available, red sections responses with no biometric data available}
@@
 
+\vspace{-8mm}
 We extracted features from the biometric data to provide as input to machine 
@@
 
+\vspace{-8mm}
 Table~\ref{responseDistribution} illustrates the distribution of responses from each participant for each of the three survey questions (listed in Section~\ref{sec:Surveys}).
@@
 
+\vspace{-8mm}
 Even after binarizing the responses as described in the previous section, we found the distribution of responses was still quite imbalanced for many of our participants.
@@
       \includegraphics[width=0.5\textwidth]{distributiontable.pdf}
-  \caption{The distribution of the responses of each participant to the three questions asked during the day are shown.
-Each bar in the histograms represent one of the 5 values on the 5-point Likert scale we asked participants to respond with, where the far left side of the histograms are 1/Not at all, and the far right sides are 5/Extremely}
+  \caption{Distribution of the responses of each participant.
+Each bar in the histograms represent one of the 5 values on the 5-point Likert scale, the far left side of the histograms are 1/Not at all, and the far right sides are 5/Extremely}
    \label{responseDistribution}
@@
 
-\subsection{Selecting a Classifier Algorithm}
+\subsection{Selecting a Classifier}
 
+\vspace{-4mm}
 There are many different algorithms that can be used to build a classifier. To select an
@@
 
+\vspace{-4mm}
 %\noindent\textit{Results of Individual Classifiers}\\
@@
 
+\vspace{-4mm}
 %\noindent\textit{Feature Selection and Importance}\\
@@
 
+\vspace{-4mm}
 %\noindent\textit{Individual vs.\ General Model}\\

### 2020-03-26 HICSSSubmission/bp_studyMethod.tex
reducing text

@@
 
+\vspace{-4mm}
 We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress over time and the feasibility of predicting stress, focus, and awakeness based on biometric signals.
@@
 
+\vspace{-4mm}
 We recruited a group of 14 professionals via personal contacts from a large power and automation company.
@@
 All participants work primarily in an office environment, though half of the participants spend at least 10\% (and up to 50\%) of their time in a laboratory environment.
-Office workers are a population that generalizes to a variety of contexts, and including part-time laboratory workers guarantees that our participants have varying work patterns that include different levels of computer usage, as well as different levels of activity in both individual and collaborative tasks.
+Office workers are a population that generalizes to a variety of contexts.
+Our participants have varying work patterns including varying levels of computer usage, and activity in both individual and collaborative tasks.
 
@@
 
+\vspace{-4mm}
 We performed a study over the course of eight weeks.
@@
 
-After the initial setup, participants were asked to fill out 3 surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}).
-At the end of the eight weeks, we collected the biometric sensors and performed a short follow-up interview on the study and the participants' experiences.
+After the initial setup, participants were asked to fill out three surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}).
+At the end of the eight weeks, we collected the biometric sensors and performed a short follow-up interview on the participants' experiences.
 
@@
 a good link to mindfulness and an intervention that might alleviate
-stress. By recording attendence of participants at the Tai Chi
+stress. By recording attendance of participants at the Tai Chi
 sessions, we are able to analyze the impact of the sessions
-on the results. As a brief summary, the impact of the
-sessions was minimal; a a result, we present an analysis
-of the effects in the Discussion section of the paper.
+on the results.
 
@@
 
+\vspace{-4mm}
 In this section we describe the two datasets that we collected from each study participant.
@@
 
+\vspace{-8mm}
 Figure~\ref{everion} illustrates Biovotion's Everion, which we used to track the biometric signals of the study participants.
@@
       \includegraphics[width=0.4\textwidth]{Everion.jpg}
-  \caption{We used Biovotion's Everion to collect biometric measurements from the participants}
+  \caption{Biovotion's Everion}
    \label{everion}
@@
 
-Table~\ref{signals} lists the biometrics measurements that we collected using the Everion.
+Table~\ref{signals} lists the biometric measurements that we collected using the Everion.
 Each measurement is collected once per second, and each recorded observation has an associated timestamp and quality rating.
-Data collected by the Everion are uploaded to a server, from which we downloaded the data for use in our study.
+Data collected by the Everion is uploaded to a server, from which we downloaded the data for our study.
 
@@
 \end{tabular}
-\caption{Biometric measurements captured by the Everion, organized by category and with references to previous works using similar data}
+\caption{Biometric measurements captured by the Everion}
 \label{signals}
@@
 
+\vspace{-8mm}
 To gain a better understanding of our participants day-to-day work activities, we used an open source computer interaction monitor~\cite{Meyer17}.
@@
 
+\vspace{-8mm}
 Following guidelines from previous studies~\cite{Lalle16,Panwar18,Luo18} and 
@@
 The same survey was sent each time:
-\begin{enumerate}
+\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
 \item How awake are you right now?
@@
 detailed below:
-\begin{enumerate}
+\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
 \item How awake have you been today?

### 2020-03-29 IJHCSSubmission/bp_conclusion.tex
put the paper in ijhcs template and added references to the papers suggested by ubicomp reviewers

--- 2020-02-14 UbiCompSubmission/bp_conclusion.tex
@@
 the study support and extend previous work. Based on
-twice collected daily survey responses, we observed that
+twice-collected daily survey responses, we observed that
 although participants sometimes saw periods of sustained stress,
@@
 biometric data with which we were able to create a model that is able to
-predict user stress, sleepyness, and lack of focus with an up to 84.65\% improvement over the baseline.
+predict user stress, sleepiness, and lack of focus with an up to 84.65\% improvement over the baseline.
 

### 2020-03-29 IJHCSSubmission/bp_data.tex
put the paper in ijhcs template and added references to the papers suggested by ubicomp reviewers

--- 2019-09-20 bp_results_1.tex
@@
 %!TEX root = bioPrediction_main.tex
-%\input{bp_results_1}
-%%\input{bp_results_2}
-
-%\section{Analysis and Results}
-
-%% To evaluate the efficacy of continuously predicting a knowledge worker's stress, focus, and awakeness in the workplace, we trained and tested machine learning classifiers using a leave-one-out cross validation.
-For prediction, we used the features extracted from the collected data as the input data, and binarized the participants' self-reported responses on stress, focus, and awakeness into two classes each (e.g., 'stressed' and 'not stressed') to use as output measures.
-
-\section{Observed Trends Over Time in Stress Levels}
-\label{stressTrends}
-
-To gain insights into how knowledge workers experience stress over an
-extended period of work, we examined the end of day survey responses
-collected from each participant to see if any identifiable trends
-emerged. We used data from 13 of the 14 participants - we excluded one participant
-from this analysis as they experienced atypical stress
-levels in the latter half of the study due to factors outside of our
-control.
-
-\subsection{Baseline Stress Levels}
-Common amongst all participants was a trend to select one stress rating
-far more frequently than any other. We will refer to this value as the
-participant's baseline stress level. All but one participant reported
-their perceived stress level for the day as their baseline stress
-level more than 50\% of the time. In total, the baseline values made
-up 65\% of the reported values collected from
-participants. Interestingly, while participants sometimes saw periods
-of sustained increases in stress, lasting as many as 6 consecutive
-workdays in the most extreme case, participants would always return to
-their baseline stress level at some point.
-
-\begin{figure}
-  \centering
-      \includegraphics[width=0.5\textwidth]{dailystress_s1.pdf}
-      \includegraphics[width=0.5\textwidth]{dailystress_s3.pdf}
-  \caption{The day-to-day stress levels reported by two participants (S1 and S3) are shown.
-The y-axis represents values on the 5-point Likert scale we asked participants to respond with ranging from 1/Not at all stressed to 5/Extremely stressed.
-The x-axis represents the day of the study on which the value was recorded (from 0-45).
-The x-axes are slightly different for the 2 participants as they did not begin and end the study on exactly the same days.}
-   \vspace*{-2mm}
-   \label{fig:dailyStress}
-\end{figure}
-
-The baseline stress level varied significantly between participants. 7
-participants (54\%) reported feeling average stress levels most frequently
-(rating 3 on our scale), while 5 (38\%) reported feeling little stress
-(rating 2) and 1 (8\%) reported feeling no stress at all (rating 1).
-
-Figure \ref{fig:dailyStress} illustrates these points, showing both participants tendency to report and return to baseline stress levels
-
-\subsection{Stressful Days Tend to Cluster}
-Accounting for the variance between participants perceived stress
-baselines, we consider a stressful day to be one that represents a
-deviation of 1 or more stress levels above the participant's
-baseline. Of the 93 stressful days we observed in total, we found that
-39 (41\%) of these days occurred in groupings of two or more
-consecutive stressful workdays. The most common size of these groups
-was two workdays, while the largest group we observed was six
-workdays.  The day after a stressful day is much more likely to be a
-stressful day as compared to any other day with a 0.55 average increase
-over baseline, compared to 0.02 average increase over baseline.
-
-\subsection{Extreme Changes in Stress Levels are Rare}
-After accounting for each participant's perceived stress baseline, we
-examined the frequency of deviations from the baseline. We found that
-participants were far more likely to report a stress level that was
-within 1 point of their baseline, than to report a stress level 2 or
-more points away. These extreme deviations represented only 15\% of
-all reported values which differed from the participants baseline. As
-well, the majority (78\%) of these deviations came from just two
-participants. This suggests that some people may be less resilient to
-the stress of the workplace than others. For most participants,
-extremely stressful days were few and far between.
-
-\subsection{Explaining Stress Fluctuations}
-We created a linear mixed model to attempt to explain some of the
-stress that our participants were experiencing. We experimented with
-multiple independent variables, including those derived from computer
-interaction data such as total time spent actively on the computer,
-percentage of time idle, and amount of time spent on non work-related
-web browsing. We also looked at the day of the week and proximity to
-beginning or end of month as possible explanatory variables. However,
-ultimately we were unable to find any variable which showed a
-significant correlation with our participants perceived stress
-levels. This points to the need for additional instrumentation if we
-are to succesfully understand and predict stress in the workplace.
-
-\section{Predicting Stress in the Moment}
-% random forest best
-\label{secOverallAccuracy}
-
-To investigate whether stress, focus and awakeness can be predicted in
-the moment based on biometric measures.
-we investigated classifiers trained
-for each individual and across all participants. We report on the effectiveness of
-these classifiers and the features that are important in prediction of stress, focus
-and awakeness.
 
@@
 \end{table}
-
-\begin{table}[h]
-	\begin{centering}
-	\small\addtolength{\tabcolsep}{-1pt}
-    \begin{tabular}{llll}
-      \hline
-      Variable & K & \# Estimators & Minimum Samples Split \\
-      \hline
-      Stress & 300 & 100 & 4\\
-      Focus & 200 & 50 & 4\\
-      Awakeness & 800 & 100 & 4\\
-      \hline
-    \end{tabular}
-    \caption{The hyperparameters selected by grid search analysis to tune our random forest models. K refers to the number of features selected.}    \label{tab:hyperparams}
-    \end{centering}
-\end{table}
-
-\subsection{Selecting a Classifier Algorithm}
-
-There are many different algorithms that can be used to build a classifier. To select an
-algorithm, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11} and performed a grid search analysis to determine the optimal hyperparameters for each classifier.
-Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, and a multilayer perceptron neural network.
-The optimal values for random forest and the three output measures are listed in Table \ref{tab:hyperparams}.
-For the remainder of this paper, we refer to random forest classifiers trained with these hyperparameters.%\\[-0.1cm]
-% for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)
-
-\begin{table*}[h]
-  \centering
-  \includegraphics[width=1.0\textwidth]{rq1performance.pdf}
-  \caption{Results of predictions using the individual models.
-The distribution columns show a bar chart of the response distribution (negative/positive) for each of the three variables.
-The baseline row represents the averaged results of our baesline classifier.
-The general row shows the averaged results of our models trained on all participants.}\label{tab:accuracy}%\todo{revise caption}}
-  \vspace*{-4mm}
-\end{table*}
-
-\subsection{Individual Classifiers}
-%\noindent\textit{Results of Individual Classifiers}\\
-Since peoples' experience of stress, focus, and awakeness (as well as
-their physiological manifestations) can vary substantially
-(e.g.,~\cite{Hernandez11}), we first trained individual classifiers
-for each participant (rather than a general one for all
-participants). The results of our analysis are reported in
-Table~\ref{tab:accuracy}. For our analysis, we report values of
-accuracy, one of the most commonly used metric to compare performance,
-as well as precision and recall of the classes of interest:
-'stressed', 'not focused', and 'not awake'. Since the imbalance in the
-data can lead to high accuracy values if a classifier always just
-predicts the most likely/frequent class while ignoring the class of
-higher importance and interest, precision and recall of the class of
-interest are also important to
-consider~\cite{yap2014,bhattacharyya_data_2011,Hernandez11}.
-For some users (i.e., S1.
-as seen In Table \ref{responseDistribution}), the imbalance in their data was so extreme that even after adjusting by oversampling it was impossible to create a reasonable classifier.
-These scenarios are difficult to predict as any classifier will not have enough variance in its training data for the 'stressed' situation to adequately distinguish it from the non-stressed case.
-
-% However, the imbalance in the data can lead to a classifier with high accuracy if it always just predicts the more likely class while ignoring the class of higher importance and interest in our case (i.e.
-stressed, not focused, not awake).
-Therefore, we further report the precision and recall for the three classes 'stressed', 'not focused' and 'not awake'.
-
-Overall, we were able to use extracted physiological features to
-predict all three aspects with reasonable accuracy, precision, and
-recall, as well as to improve upon the baseline---a stratified random
-classifier that randomly chooses one of the two classes with a bias
-towards the larger class. While the individually trained classifiers
-improved on average across all participants upon the baseline in all
-cases excepting recall of 'not focused', the improvement was
-substantially higher for awakeness (85\% improvement in precision,
-62\% in recall, and 7\% in accuracy) than for stress or focus. Also,
-the performance of the individually trained classifiers varied greatly
-across participants. While some participants showed a large
-improvement, for others the baseline performed much better than the
-individually trained classifier. For instance, for predicting
-'stressed', the individual classifiers improved upon the baseline for
-S4, S6, S8, S11, S12, and S14 with a maximum improvement of 88.4\% in
-precision and 184.6\% in recall for S12, while they did worse for S1,
-S3, S5, S7, S9, S10, and S13, and in the worst cases did not correctly
-predict a single instance of 'stressed'. Typically users that have the
-lowest precision and recall values are those where the data is the
-most unbalanced.\\[-0.1cm]
-% We attribute these discrepancies to the large differences in response distributions between participants, as well as to the subjectivity of self-reporting.
-%
-%(improvement: 8\% precision, 1\% recall, 8\% accuracy), the individual performance varied greatly among participants.
-Some participants showed negligible difference in comparison to the baseline classifier, e.g.
-subject ..., while others showed a large improvement, e.g.
-subject ...
-%\begin{table}[h!]
-%\begin{centering}
-%\begin{tabular}{lll}
-%\hline
-%Participant & Recall Improvement (\%) & Precision Improvement (\%)\\
-%\hline
-%S12 & 184.6 & 88.4 \\
-%S6 & 66.7 & 85.8 \\
-%S11 & 50.0 & 44.4 \\
-%S8 & 32.4 & 26.7 \\
-%S14 & 14.7 & 9.8 \\
-%S4 & 5.6 & 5.5 \\
-%S9 & -55.4 & -46.8 \\
-%S3 & -90.0 & -82.1 \\
-%S1 & -100.0 & -100.0 \\
-%S5 & -100.0 & -100.0 \\
-%S7 & -100.0 & -100.0 \\
-%S10 & -100.0 & -100.0 \\
-%S13 & -100.0 & -100.0\\
-%S2 & - & -\\
-%\hline
-%\end{tabular}
-%\caption{Percentage improvement in recall and precision for stress, using our approach compared to the baseline, on an individual level.
-For participant 2, the baseline achieved a precision and recall of 0, thus the improvement is undefined}
-%\end{centering}
-%\label{tab:indImprovement}
-%\end{table}
-
-\subsection{Feature Selection and Importance}
-%\noindent\textit{Feature Selection and Importance}\\
-There are a large variety of features that can be (and have been) calculated in previous research for each of the basic measurements listed in Table~\ref{signals}, such as the mean, standard deviation, maximum, and interquartile range.
-In addition, each of these metrics can be combined with the various time windows captured of a basic measurement, resulting in a large feature space.
-To reduce the feature space, we experimented with multiple feature selection methods, including selecting the top k highest correlated features by various metrics such as mutual information, Pearson's correlation coefficient, ANOVA's F-value, as well as wrapper methods such as recursive feature elimination, optimizing mean decrease accuracy by iteratively permuting features, and only selecting features that exceed a certain Gini importance threshold.
-We found that all methods produced similar results with respect to accuracy, precision, and recall for the individual models.
-Ultimately, we chose to use the top k features with the highest ANOVA F-value, as it is relatively simple and efficient to calculate.
-The values of k used were selected by grid search analysis, and are shown in Table \ref{tab:hyperparams}.
-
-%Respiration rate is also highly important for stress (#3 , 14.4%) but I'm not sure what previous works say about correlation with stress
-Overall, the features that were selected as the important ones for the individual models based on the random forest algorithm varied greatly across participants.
-Yet, some feature categories were considered to be important more frequently than others.
-Table~\ref{tab:featureImportance} shows the averaged Gini importance for the feature categories used for predicting stress.
-Particularly important for stress were the feature categories heart rate variability (18.3\%) and skin temperature (15.7\%), both of which have been shown in several previous studies to be  indicators for stress~\cite{dishman2000stress,mcduff16,kataoka00}.
-For predicting focus, the feature categories for blood pulse wave (14.2\%) and heart rate variability (13\%) showed to be the most important categories, while for awakeness the most important ones were heart rate variability (13.6\%) and blood pulse wave (13.1\%).\\[-0.1cm]
-
-\begin{table}[h!]
-  \begin{centering}
-  \begin{tabular}{llll}
-    \hline
-    Feature Category & Stress & Focus & Awakeness\\
-    \hline
-    Heart Rate Variability & 18.3\% & 13\% & 13.6\%\\
-    Blood Pulse Wave & 10\% & 14.2\% & 13.1\%\\
-    Heart Rate & 8.7\% & 12.6\% & 10.3\%\\
-    Skin Temperature & 15.7\% & 9.8\% & 10\%\\
-    Galvanic Skin Response & 6.6\% & 8.2\% & 5.1\%\\
-    Respiration Rate & 14.8\% & 12.7\% & 10\%\\
-    Oxygen Saturation & 5.6\% & 4.3\% & 2\%\\ 
-    Energy Expenditure & 6\% & 7.7\% & 4.8\%\\
-    Activity & 4.6\% & 7.8\% & 7.6\%\\
-    Steps & 1.7\% & 0.8\% & 0.9\%\\
-    Time of Day & 0\% & 0.1\% & 0.5\%\\
-    \hline
-  \end{tabular}
-  \caption{The averaged Gini importance of each feature category, per response variable.}
-  \label{tab:featureImportance}
-  \end{centering}
-  \vspace*{-2mm}
-\end{table}
-
-\subsection{Individual vs.\ General Model}
-%\noindent\textit{Individual vs.\ General Model}\\
-Individual models are trained specifically for each individual and thus require a data collection period before they are capable of making accurate predictions.
-On the other hand, the idea of general models is to be able to train them on already collected data and then  to be able to apply them even to new and unseen individuals, thus overcoming the cold-start problem.
-Given the large individual differences in biometrics, training a general model to achieve an adequate accuracy for new individuals is not necessarily possible.
-
-To examine the performance of a general model for our participants, we trained three general models, one for focus, one for awakeness and one for stress.
-We roughly followed the same procedure as for the individual models.
-Due to the larger amount of data available in the general case, we used the more common random undersampling, which randomly selects elements in the majority class to exclude from the dataset, instead of random oversampling to balance the distribution of the dataset.
-The models were trained on the datasets of 13 of the 14 participants, and then evaluated on the dataset of the last, repeating this process for all 14 participants.
-
-The bottom row of Table \ref{tab:accuracy} presents the averaged performance results for this approach in terms of accuracy, precision, and recall.
-Although the averaged precision and recall are comparable or better than those of the averaged individual results, this was at the cost of a large decrease in overall accuracy.
-Upon closer investigation into the performance of the general model when testing on each participant, we found that individually trained models for each participant performed much better than a general model trained over all participants.
-Using stress as an example, for participant S12, for whom we saw the greatest increase compared to the baseline in individual models, the general model was unable to predict a single instance of 'stressed' correctly.
-This is consistent with our expectations because biometric features are highly specific to individuals.
-
-%\todo{put the content of the following sentence somewhere into the discussion; We attribute these discrepancies to the large differences in response distributions between participants, as well as to the subjectivity of self-reporting.}

### 2020-03-29 IJHCSSubmission/bp_intro_v2.tex
put the paper in ijhcs template and added references to the papers suggested by ubicomp reviewers

--- 2019-05-31 BiometricsPaper/bp_intro_v3.tex
@@
 \section{Introduction}
-\textit{`The most valuable asset of a 21st-century institution (whether business or non-business) will be its knowledge workers and their productivity'}~\cite{drucker1999knowledge}.
-Yet, in today's workplaces, knowledge workers and their time are not necessarily treated as the most valuable asset, and they constantly face challenges, such as a high work fragmentation, continuous disruptions and distractions, highly complex and demanding tasks, and long working hours~\cite{gonzalez2004constant,mark2008cost,czerwinski04diary}.
-These challenges make it difficult for a knowledge worker to stay focused and can impact both the quality of work performed and productivity.
-For instance, the continuous disruptions that knowledge workers face in office work environments can lead to a higher error rate and a lower performance~\cite{bailey2001effects,mark2008cost}.
-In some cases,
-the consequences are even more severe; in control room
-settings, the consequences can be accidents that cost millions of dollars in losses. 
+Ideally businesses believe that \textit{`The most valuable asset of a 21st-century institution (whether business or non-business) [is] its knowledge workers and their productivity'}~\cite{drucker1999knowledge}.
+In practice these valuable assets are as mercilessly optimized as assembly lines during the industrial age.
+Knowledge workers' desks occupy increasingly small footprints, optimizing a company's cost per square unit.
+Their shifts, especially in factories, command centers, or web service organizations, span twenty four hours a day, optimizing uptime.
+
+In these settings they face a variety of challenges.
+Their squeezed workspaces lead to continuous disruptions and distractions, on top of the high work fragmentation and highly complex tasks typical of modern office work~\cite{gonzalez2004constant,zuger2015interruptibility,zuger18,mark2014bored}.
+%This leads to an up to 80\% of accidents being attributed, at least in part, to actions or omissions of workers\cite{hse99}.
+Their shift work--and the required overnight shifts that accompany it--naturally causes sleepines.
+These challenges combine to increase stress and decrease focus, which in turn impacts worker productivity.
+For instance, the continuous disruptions that knowledge workers face can lead to higher error rates, lower overall performance~\cite{bailey2001effects,mark2008cost}, and, in a control room setting, could easily cause an accident that costs millions of dollars.
 %when a sleepy operator might, for example, miss early warning signs and cause...
 %\todo{Nick: can you find a good example here and citation?}. 
-%Similarly, stress at the workplace is a growing concern at the workplace and one of the most common work-related health problem that can lead to fatigue, burnout and various other illnesses, resulting in work absence and a high productivity loss~\cite{hockey1997stress,setz2010stress,wrs2010}.
-Less immediate, but just as insidious, stress at the workplace is a growing concern and one of the most common work-related health problems.
+Less immediate but just as insidious, stress at the workplace is a growing concern and one of the most common work-related health problems.
 It leads to fatigue, burnout and various other illnesses, ultimately resulting in work absences and marked productivity losses~\cite{hockey1997stress,setz2010stress,wrs2010}.
 
-Being able to continuously measure human aspects --- such as focus, awakeness, and stress --- in the workplace bears huge potential for better supporting knowledge workers, combatting workplace challenges, and improving knowledge worker productivity.
-Such measures could be used to provide better management of disruptions~\cite{iqbal2005effectiveness,zuger2017reducing}, to automatically adjust lighting to reduce sleepiness among control room operators, or to continuously monitor stress levels to trigger pro-active interventions and reduce health problems.
+The first step in combating these common workplace challenges is to be able to measure their effect on workers.
+Yet measuring human aspects--focus, stress, and awakeness--has traditionally been done manually, making it difficult to efficiently apply Taylorism's scientific management.
+And yet the benefits of doing so are clear; measuring human aspects could provide better management of disruptions~\cite{iqbal2005effectiveness,zuger2017reducing}, automatically adjust lighting to circadian cycles to reduce sleepiness among shift workers, or continuously monitor stress levels to trigger pro-active interventions.
 
-Fortunately, with recent advances in sensing technologies there is increasing hope that we can accurately and automatically measure such human aspects.
-A number of studies have already discussed and investigated the use of biometrics to measure aspects such as a person's cognitive and emotional states~\cite{beatty82,Kramer90,Rowe98,richter1998psychophysiological,Wilson2002}.
-Studies have also looked more specifically at using biometrics to measure stress~\cite{dishman2000stress,Setz2010}, or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004emotion}.
+Fortunately, with recent advances in sensing technologies there is increasing hope that we can accurately and automatically measure human aspects.
+A number of recent studies have discussed and investigated the use of biometrics to measure aspects such as a person's cognitive and emotional states~\cite{beatty82,picard1995affective,Kramer90,Rowe98,richter1998psychophysiological,Wilson2002}.
+Specifically, studies have looked at using biometrics to measure stress~\cite{dishman2000stress,Setz2010}, or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004}.
 Yet, very few studies measure productivity-related factors in the workplace over an extended period of time, and little is known about measuring the more abstract concept of focus.
-Given the importance of focus, described as a combination of engagement and challenge in a work task by Mark et al.~\cite{mark2014bored}, to an office workers' productivity, this factor also deserves further study.
+Given the importance of focus, described as a combination of engagement and challenge in a work task by Mark et al.~\cite{mark2014bored}, to an office workers' productivity, this factor deserves further study.
 
-The objective of our research is to develop continuous (and automatic) measures of focus, awakeness, and stress in the workplace.
+Thus, the objective of our research is to continuously (and automatically) measure focus, awakeness, and stress in the workplace.
 These measures are targeted with an eye towards improving knowledge workers' productivity and well-being in the future.
-For instance, by automatically protecting the worker from audible interruptions during a high focus period by showering him/her with white noise or by recommending stress-reducing interventions during an extended period of high stress.
+For instance, by automatically protecting the worker from audible interruptions during a high focus period by showering him/her with white noise or by recommending stress-reducing interventions such as Tai Chi during an extended period of high stress.
 
@@
 Using machine learning, we created classifiers and analyzed their ability to predict the three productivity-related human aspects.
-In addition, we compared biometric features with computer interaction features in their predictive power, we analyzed how the combination
-of these two training sets compares to the performance of each
-set individually.
-
-The results of our analysis show that biometrics of a single minimally invasive sensor can be used to predict productivity-related indicators accurately, with the abstract concept of focus (predictably) being the hardest to detect.
-The results also show that knowledge workers' self-reported levels of stress, focus and awakeness and their physiological manifestation and prediction can vary a lot between individuals.
-Our results further show that overall combining biometric features with computer interaction information
-results in an improvement
-over each of the training sets individually.
-The main contributions of our work are:
-\begin{itemize}
-	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' awakeness, stress, and focus in the workplace based on an eight-week field study with 14 office workers.
-	\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
-	\item A discussion on the impact of applying this research to possible target workspaces, besides a reflection on aspects that can be further improved in future studies.
-    \end{itemize}
-
-%Ideally businesses believe that \textit{`The most valuable asset of a 21st-century institution (whether business or non-business) [is] its knowledge workers and their productivity'}~\cite{drucker1999knowledge}.
-In practice these valuable assets are as mercilessly optimized as assembly lines during the industrial age.
-Knowledge workers' desks occupy increasingly small footprints, optimizing a company's cost per square unit.
-Their shifts, especially in factories, command centers, or web service organizations, span twenty four hours a day, optimizing uptime.
-%
-%
-%
-%In these settings they face a variety of challenges.
-Their squeezed workspaces lead to continuous disruptions and distractions, on top of the high work fragmentation and highly complex tasks typical of modern office work~\cite{gonzalez2004constant,zuger2015interruptibility,zuger18,mark2014bored}.
-%This leads to an up to 80\% of accidents being attributed, at least in part, to actions or omissions of workers\cite{hse99}.
-%Their shift work--and the required overnight shifts that accompany it--naturally causes sleepines.
-These challenges combine to increase stress and decrease focus, which in turn impacts worker productivity.
-For instance, the continuous disruptions that knowledge workers face can lead to higher error rates, lower overall performance~\cite{bailey2001effects,mark2008cost}, and, in a control room setting, could easily cause an accident that costs millions of dollars.
-%%when a sleepy operator might, for example, miss early warning signs and cause...
-%%\todo{Nick: can you find a good example here and citation?}. 
-%Less immediate but just as insidious, stress at the workplace is a growing concern and one of the most common work-related health problems.
-It leads to fatigue, burnout and various other illnesses, ultimately resulting in work absences and marked productivity losses~\cite{hockey1997stress,setz2010stress,wrs2010}.
-%
-%The first step in combating these common workplace challenges is to be able to measure their effect on workers.
-Yet measuring human aspects--focus, stress, and awakeness--has traditionally been done manually, making it difficult to efficiently apply Taylorism's scientific management.
-And yet the benefits of doing so are clear; measuring human aspects could provide better management of disruptions~\cite{iqbal2005effectiveness,zuger2017reducing}, automatically adjust lighting to circadian cycles to reduce sleepiness among shift workers, or continuously monitor stress levels to trigger pro-active interventions.
-%
-%Fortunately, with recent advances in sensing technologies there is increasing hope that we can accurately and automatically measure human aspects.
-A number of recent studies have discussed and investigated the use of biometrics to measure aspects such as a person's cognitive and emotional states~\cite{beatty82,picard1995affective,Kramer90,Rowe98,richter1998psychophysiological,Wilson2002}.
-Specifically, studies have looked at using biometrics to measure stress~\cite{dishman2000stress,Setz2010}, or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004}.
-Yet, very few studies measure productivity-related factors in the workplace over an extended period of time, and little is known about measuring the more abstract concept of focus.
-Given the importance of focus, described as a combination of engagement and challenge in a work task by Mark et al.~\cite{mark2014bored}, to an office workers' productivity, this factor deserves further study.
-%
-%
-%Especially with the advances in sensing technologies, there is an increasing number of studies that discussed and investigated the use of biometrics to measure aspects such as a person's cognitive and emotional states~\cite{beatty82,picard1995affective,Kramer90,Rowe98,richter1998psychophysiological,Wilson2002}.
-Studies have also looked more specifically at using biometrics to measure stress~\cite{dishman2000stress,Setz2010}, or awakeness / (energetic) arousal~\cite{exler2016tired,mcduff2012affectaura,haag2004}.
-Yet, only very few studies have looked at measuring such productivity-related factors in the workplace over an extended period of time.
-Also, little is known about measuring a more abstract concept of focus that is often used in the context of knowledge workers' productivity and was, for example, described as a combination of engagement and challenge in a work task by Mark et al.~\cite{mark2014bored}.
-
-%Thus, the objective of our research is to continuously (and automatically) measure focus, awakeness, and stress in the workplace.
-These measures are targeted with an eye towards improving knowledge workers' productivity and well-being in the future.
-For instance, by automatically protecting the worker from audible interruptions during a high focus period by showering him/her with white noise or by recommending stress-reducing interventions such as Tai Chi during an extended period of high stress.
-%
-%Our work builds upon previous work and extends it by performing an eight-week study in the workplace with 14 knowledge workers, collecting biometric and computer interaction data as well as experience samples on focus, awakeness, and stress.
-Our participants, who perform various job functions for a research and development group within a single large corporation, wore a single biometric armband sensor\footnote{Biovotion Everion sensor~\cite{everion}} with low invasiveness that captured heart- and skin-related measures.
-This modality was chosen to ease longitudinal deployment in the field.
-Using machine learning, we created classifiers and analyzed their ability to predict the three productivity-related human aspects.
 In addition, we compared biometric features with computer interaction features in their predictive power, we analyzed ..
 and ...
-%
-%The results of our analysis show that biometrics of a single minimally invasive sensor can be used to ...
+
+The results of our analysis show that biometrics of a single minimally invasive sensor can be used to ...
 accurately, with the abstract concept of focus (predictably) being the hardest to detect, yet even its detection within reasonable limits was feasible.
@@
 Our results further show that ...
-%The main contributions of our work are the creation and analysis of measures for the automatic monitoring of knowledge workers' awakeness, stress, and the more abstract concept of focus in the workplace based on an eight-week field study with 14 office workers.
+The main contributions of our work are the creation and analysis of measures for the automatic monitoring of knowledge workers' awakeness, stress, and the more abstract concept of focus in the workplace based on an eight-week field study with 14 office workers.
 a thorough analysis of ...

### 2020-03-29 IJHCSSubmission/bp_relatedWork.tex
put the paper in ijhcs template and added references to the papers suggested by ubicomp reviewers

--- 2020-02-03 UbiCompSubmission/bp_relatedWork.tex
@@
 
-The study and prediction approach on which we report in this paper is related
+The study and prediction approach used in this paper is related
 to previous studies of productivity-related indicators and studies
-using biometerics to predict these indicators. We consider
+using biometrics to predict them. We consider
 related work in each of these categories in turn.
@@
 
+Zaman et al.~\cite{Zaman19} developed a study where they measure stress and productivity in short sessions with 63 participants using biometric sensors (thermal facial camera, wrist EDA, chest breathing sensor, and facial camera).
+Different from this study, our corpus is gathered from real-world office workers.
+Likewise, our study spans for eight weeks allowing us to analyze the behavior or our participants through the weeks and providing a more extensive in-depth analysis of our participants over time.
+
+Mirjafari et al.~\cite{Mirjafari19} use mobile sensing data to gather data and build machine learning models to group workers based on their workplace performance metrics.
+Similarly, Sarsenbayeva et al.~\cite{Sarsenbayeva19} study how stress affects mobile interaction reducing completion time and accuracy during target acquisition tasks.
+
 Hovsepian et al.~\cite{Hovsepian15} have worked to obtain a standard for continuous stress assessment.
@@
 
-Hernandez et al.~\cite{Hernandez14} investigated the use of pressure-sensitive keyboard and a capacitative mouse as non-intrusive means for measuring computer users stress levels.
+Hernandez et al.~\cite{Hernandez14} investigated the use of pressure-sensitive keyboard and a capacitive mouse as non-intrusive means for measuring computer users stress levels.
 They found participants exhibited significantly increased typing pressure and mouse contact when in stressful conditions.
@@
 Kocielnik et al.~\cite{kocielnik_smart_2013} developed a framework for unobtrusive and continuous measurement of stress in real life conditions.
-They equipped university staff members with a wristband sensor and combined this data with information from the participants calendars over the course of four weeks.
+They equipped university staff members with a wristband sensor and combined this data with information from the participants' calendars over the course of four weeks.
 They observed that the data they collected reflected well the participants perceptions of their own stress levels.
-However this work focused mostly on providing retrospective information to users so they can work to improve their own stress balance.
+However, this work focused mostly on providing retrospective information to users so they can work to improve their own stress balance.
 They did not attempt to make observations about the big picture of participants stress profiles or make predictions in real time.
@@
 
-Interruptions in the office are a common barrier keeping workers from sustaining focus on their work related activities, particularly when the interruptions occur at inopportune moments.
+Interruptions in the office are a common barriers keeping workers from sustaining focus on their work related activities, particularly when the interruptions occur at inopportune moments.
 Such interruptions may include emails, alerts, or interactions with co-workers\cite{gonzalez2004constant,chong2006interruptions,shamsi07}.
@@
 In contrast to studies that use eye tracking or EEG, we focused on a less invasive technology that can be applied in a real world scenario.
+Similarly, we considered adding additional metrics such as Depression, Anxiety and Stress Scale (DASS) and Perceived Stress Scale (PSS)~\cite{Ferdous15} but opted for the ones presented in the paper after several discussions with experts in the area, and after piloting surveys to maximize participant compliance.

### 2020-03-29 IJHCSSubmission/bp_results_1.tex
put the paper in ijhcs template and added references to the papers suggested by ubicomp reviewers

--- 2020-02-06 UbiCompSubmission/bp_results_1.tex
@@
   \centering
-      \includegraphics[width=0.5\textwidth]{dailystress_s1.pdf}
-      \includegraphics[width=0.5\textwidth]{dailystress_s3.pdf}
+      \includegraphics[width=0.7\textwidth]{dailystress_s1.pdf}
+      \includegraphics[width=0.7\textwidth]{dailystress_s3.pdf}
   \caption{The day-to-day stress levels reported by two participants (S1 and S3) are shown.
@@
 more points away. These extreme deviations represented only 15\% of
-all reported values which differed from the participants baseline. As
+all reported values, which differed from the participants baseline. As
 well, the majority (78\%) of these deviations came from just two
@@
 \begin{itemize}
-\item The participant not wearing the Everion in the hour before before beginning the survey
+\item The participant not wearing the Everion in the hour before beginning the survey
 \item The Everion not recording data in the hour before the participant began the survey (e.g., due to low battery)
@@
   \centering
-      \includegraphics[width=0.5\textwidth]{DuringTheDay.png}
+      \includegraphics[width=0.8\textwidth]{DuringTheDay.png}
   \caption{The figure shows the biometric data available per each participant. Green sections represent survey responses with biometric
@@
   \centering
-      \includegraphics[width=0.5\textwidth]{distributiontable.pdf}
+      \includegraphics[width=0.8\textwidth]{distributiontable.pdf}
   \caption{The distribution of the responses of each participant to the three questions asked during the day are shown.
@@
    \label{responseDistribution}
-   \vspace*{-2mm}
+   %\vspace*{-2mm}
 \end{table}
@@
 The distribution columns show a bar chart of the response distribution (negative/positive) for each of the three variables.
-The baseline row represents the averaged results of our baesline classifier.
+The baseline row represents the averaged results of our baseline classifier.
 The general row shows the averaged results of our models trained on all participants.}\label{tab:accuracy}%\todo{revise caption}}
@@
 substantially higher for awakeness (85\% improvement in precision,
-62\% in recall, and 7\% in accuracy) than for stress or focus. Also,
+62\% in recall, and 7\% in accuracy) than for stress or focus. In addition,
 the performance of the individually trained classifiers varied greatly
@@
 S3, S5, S7, S9, S10, and S13, and in the worst cases did not correctly
-predict a single instance of 'stressed'. Typically users that have the
+predict a single instance of 'stressed'. Typically, users that have the
 lowest precision and recall values are those where the data is the

### 2020-03-29 IJHCSSubmission/bp_studyMethod.tex
put the paper in ijhcs template and added references to the papers suggested by ubicomp reviewers

--- 2020-02-14 UbiCompSubmission/bp_studyMethod.tex
@@
 a good link to mindfulness and an intervention that might alleviate
-stress. By recording attendence of participants at the Tai Chi
+stress. By recording attendance of participants at the Tai Chi
 sessions, we are able to analyze the impact of the sessions
 on the results. As a brief summary, the impact of the
-sessions was minimal; a a result, we present an analysis
+sessions was minimal; a result, we present an analysis
 of the effects in the Discussion section of the paper.
@@
   \centering
-      \includegraphics[width=0.4\textwidth]{Everion.jpg}
+      \includegraphics[width=0.5\textwidth]{Everion.jpg}
   \caption{We used Biovotion's Everion to collect biometric measurements from the participants}
@@
 Following guidelines from previous studies~\cite{Lalle16,Panwar18,Luo18} and 
-following the preferrences of extensive user piloting, we sent via 
+following the preferences of extensive user piloting, we sent via 
 text message a survey request to each participant two times per 
@@
 for too much freedom in response time~\cite{Adams18} since this would 
-discurage participation in stressed timeframes and would bias the 
+discourage participation in stressed time frames and would bias the 
 corpus. 

### 2020-03-29 IJHCSSubmission/bp_threats.tex
put the paper in ijhcs template and added references to the papers suggested by ubicomp reviewers

--- 2019-09-20 bp_threats.tex
@@
 
-\textbf{External Validity:}
+\subsection{External Validity:}
 The results of our study may 
@@
 
-Another threat is that our results may notgeneralize
+Another threat is that our results may not generalize
 to a different office environment. We conducted
@@
 These office environments control for a series of
-variables to make them standard world wide such
+variables to make them standard worldwide such
 as controlled temperature and lighting.
 
-\textbf{Internal Validity:}
+\subsection{Internal Validity:}
 This study tries to find correlations between
@@
 It is possible that the amount of data collected
-is not sufficienet
+is not sufficient
 to draw valid conclusions. To address this threat, 
-we collected a data for an eight week period, which is
-400\% longer than previous studies~\cite{zuger18,Muller16}.
+we collected a data for an eight-week period, which is
+400\% longer than the longest previous studies~\cite{zuger18,Muller16}.
 
-\textbf{Construct Validity:}
+\subsection{Construct Validity:}
 A threat to the study is that
@@
 are unrelated biometric signals.
-To mitigate this risk, we used an state-of-the-art
+To mitigate this risk, we used a state-of-the-art
 device that captures a large number of highly accurate biometric

### 2020-03-29 IJHCSSubmission/elsarticle-template.tex
put the paper in ijhcs template and added references to the papers suggested by ubicomp reviewers

%\documentclass[review]{elsarticle}

%\documentclass[preprint,12pt,plainnat]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
   \documentclass[final,1p,times,plainnat]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
%\AtBeginDocument{%
%  \providecommand\BibTeX{{%
%    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

\usepackage{lineno,hyperref}
\modulolinenumbers[5]

\usepackage{balance}       % to better equalize the last page
\usepackage{graphics}      % for EPS, load graphicx instead 
\usepackage[T1]{fontenc}   % for umlauts and other diaeresis
\usepackage{txfonts}
\usepackage{mathptmx}
%\usepackage[pdflang={en-US},pdftex]{hyperref}
\usepackage{color}
\usepackage{booktabs}
\usepackage{textcomp}
\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}

\renewcommand{\textrightarrow}{$\rightarrow$}

\usepackage{booktabs} % For formal tables
\usepackage{comment}
\usepackage{tabularx}

% Some optional stuff you might like/need.
\usepackage{microtype}        % Improved Tracking and Kerning

\usepackage{todonotes}

\presetkeys{todonotes}{inline}{}
% for commenting and making it visible in the pdf (easy to hide after)
\newboolean{hidecomments}
\setboolean{hidecomments}{false}
%\nochangebars
\ifthenelse{\boolean{hidecomments}}
{\newcommand{\cb}[2]{}}
{\newcommand{\cb}[2]{
		\fbox{\bfseries\sffamily\scriptsize#1}
		{\sf\small$\blacktriangleright$ %$\RHD$
			{#2} $\blacktriangleleft$}}} %\LHD$}}}%
\newcommand\tf[1]{\cb{TF}{\textcolor{green}{#1}}}
\newcommand\cs[1]{\cb{CS}{\textcolor{blue}{#1}}}
\newcommand\ms[1]{\cb{MS}{\textcolor{red}{#1}}}
\newcommand\tfC[1]{\todo[color=green,inline]{Thomas: #1}}

\journal{International Journal of Human-Computer Studies}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frontmatter}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Observing and Predicting Knowledge Worker Stress in the Wild}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Authors will be visible after double blind review}
%\vspace{110cm}
%\vspace{18ex}
%}

\begin{comment}

\author{Mauricio Soto}
\affiliation{%
\institution{ABB Future Labs}
\city{Raleigh}
\state{North Carolina}
}
\email{mauricio.soto@us.abb.com}

\author{Chris Satterfield}
\affiliation{%
\institution{The University of British Columbia}
\city{Vancouver}
\state{British Columbia}
}
\email{cds00@cs.ubc.ca}

\author{Thomas Fritz}
\affiliation{%
\institution{University of Zurich}
\city{Zurich}
\state{Switzerland}
}
\email{fritz@ifi.uzh.ch}

\author{Gail C. Murphy}
\affiliation{%
\institution{The University of British Columbia}
\city{Vancouver}
\state{British Columbia}
}
\email{murphy@cs.ubc.ca}

\author{David Shepherd}
\affiliation{%
\institution{Virginia Commonwealth University}
\city{Richmond}
\state{Virginia}
}
\email{shepherdd@vcu.edu}

\author{Nicholas Kraft}
\affiliation{%
\institution{UserVoice}
\city{Raleigh}
\state{North Carolina}
}
\email{kraft.nicholas.a@gmail.com} \\}

\end{comment}

%% Group authors per affiliation:
%\author{Elsevier\fnref{myfootnote}}
%\address{Radarweg 29, Amsterdam}
%\fntext[myfootnote]{Since 1880.}

%% or include affiliations in footnotes:
%\author[mymainaddress,mysecondaryaddress]{Elsevier Inc}
%\ead[url]{www.elsevier.com}

%\author[mysecondaryaddress]{Global Customer Service
%\corref{mycorrespondingauthor}}
%\cortext[mycorrespondingauthor]{Corresponding author}
%\ead{support@elsevier.com}

%\address[mymainaddress]{1600 John F Kennedy Boulevard, Philadelphia}
%\address[mysecondaryaddress]{360 Park Avenue South, New York}

\begin{abstract}
Knowledge workers face many challenges in the workplace: work is 
fragmented, disruptions are constant, tasks are complex, and work hours can be 
long. These challenges can impact knowledge workers' stress, focus and awakeness, and in turn their interaction with the digital environment, the quality of work 
performed and their productivity in general. We report on a field study with 14
knowledge workers over an eight-week period in which we investigated, using
experience sampling, how the workers experience stress over time. During this field study, we also collected biometric data, which we then used to investigate if it is possible to predict such productivity-related indicators as stress, focus and awakeness, in the moment.
We observed and report on various trends in knowledge worker stress levels
over several weeks and show that machine learning models can be built to
predict stress, focus and awakeness from biometric data.
\end{abstract}

\begin{keyword}
Biometrics, Stress, Awakeness, Focus, Computer Interaction, Ubiquitous Computing, Empirical Study, User Centered Design
\end{keyword}

\end{frontmatter}

%\linenumbers

%\maketitle

\input{bp_intro_v4}
\input{bp_studyMethod}
%\input{bp_data}
\input{bp_results_1}
%\input{bp_results_2}
\input{bp_discussion}
\input{bp_relatedWork}
\input{bp_threats}
\input{bp_conclusion}

\balance{}

% REFERENCES FORMAT
% References must be the same font size as other body text.
%\bibliographystyle{ACM-Reference-Format}
\bibliography{bp_bibliography}

\end{document}\endinput

### 2020-08-07 IJHCSSubmission/bp_conclusion.tex
Updates from Overleaf

@@
 knowledge workers in their workplace over an eight week period to
-better understand how workers experience stress over time and to
-examine the ability of biometrics to predict these
-productivity-related aspects. The longitudinal and in-situ placement of
+better understand how workers experience \rev{these productivity-related human aspects} over time and to
+examine the ability of biometrics to predict these aspects. The longitudinal and in-situ placement of
 the study support and extend previous work. Based on
@@
 biometric data with which we were able to create a model that is able to
-predict user stress, sleepiness, and lack of focus with an up to 84.65\% improvement over the baseline.
+predict user stress, sleepiness, and lack of focus with \rev{small improvements in accuracy (from 3.1\% to 7.2\%, depending on the aspect in question), and moderate improvements in precision (25.9\% - 52.4\%) in comparison to a stratified random classifier.
+While the precision and recall scores we report are still low overall, this is a difficult problem to solve and our improvements indicate the potential for future researchers to build upon.}
 

### 2020-08-07 IJHCSSubmission/bp_discussion.tex
Updates from Overleaf

--- 2020-02-04 UbiCompSubmission/bp_discussion.tex
@@
 %!TEX root = bioPrediction_main.tex
-\section{Discussion}
+\section{Discussion}\label{sec:Discussion}
 
@@
 As described in our study procedure,
-due to our focus on stress, we offered participants an opportunity to be part of Thai Chi classes.
+due to our focus on stress, we offered participants an opportunity to be part of Tai Chi classes.
 The two primary reasons for this intervention was to keep participants motivated to continue the self-reports over the long study period, and to offer a technique that might help reduce stress.
@@
 the individual models and undersampling in case of a general model
-where more data was available. Especially in light of this imbalance
+where more data was available. \rev{Rebalancing the dataset using such techniques
+is a common and effective practice~\cite{branco2016survey}.
+Alternative techniques such as SMOTE~\cite{chawla2002smote} exist and can perform better than those we employed, however they are impractical considering the limited amount of data we have to work with.
+Oversampling the dataset as we did may also lead to an increased risk of model overfitting.
+However, we believe the benefits of rebalancing the dataset outweigh this possible downside.}
+
+Especially in light of the imbalance
 in the data, the results we achieved with our models are

### 2020-08-07 IJHCSSubmission/bp_relatedWork.tex
Updates from Overleaf

@@
 
+%\gm{Move to section 2 to help address reviewers comments about relationship between factors}
+
 The study and prediction approach used in this paper is related
-to previous studies of productivity-related indicators and studies
+to previous studies of \rev{stress, focus, and awakeness} and studies
 using biometrics to predict them. We consider
 related work in each of these categories in turn.
-
-\subsection{Productivity-related Indicators}
-
-Our study focuses on the productivity-related indicators of stress, focus and awakeness.
 
@@
 
-\subsubsection{Stress}
+\subsection{Stress}
 Much previous work relates to identifying and mitigating stress in an office environment.
@@
 
-\subsubsection{Focus}
+\subsection{Focus}
 Focus refers to the allocation of limited cognitive processing resources~\cite{Anderson04}.
@@
 
-\subsubsection{Awakeness}
+\subsection{Awakeness}
 
@@
 \subsection{Biometrics}
-Our study investigates the prediction of multiple productivity-related indicators (stress, focus, and awakeness) using two different types of measurements, biometric signals and computer interaction, over eight weeks in a real-life office setting.
+Our study investigates the prediction of multiple productivity-related indicators (stress, focus, and awakeness) using two different types of measurements, and biometric signals over eight weeks in a real-life office setting.
 Existing work~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015} analyzes a broad array of biometric signals and correlates them with individual's cognitive states and processes.

### 2020-08-07 IJHCSSubmission/bp_results_1.tex
Updates from Overleaf

@@
 
-\section{Observed Trends Over Time in Stress Levels}
+\section{Observed Trends Over Time in Stress \rev{and Awakeness} Levels}
 \label{stressTrends}
 
-To gain insights into how knowledge workers experience stress over an
+To gain insights into how knowledge workers experience stress \rev{and awakeness} over an
 extended period of work, we examined the end of day survey responses
 collected from each participant to see if any identifiable trends
-emerged. We used data from 13 of the 14 participants - we excluded one participant
+emerged. \rev{As we noted in the last section, we did not ask
+participants about focus in the end of the day surveys as focus is an 
+aspect relevant at a particular moment in time rather than an aspect
+for an extended period of work.} We used data from 13 of the 14 participants - we excluded one participant
 from this analysis as they experienced atypical stress
@@
 
-\subsection{Baseline Stress Levels}
+\subsection{Stress Levels}
+
+\rev{We describe characteristics noted in the reporting of the knowledge workers
+about the stress they experienced over the duration of the study.}
+
+\begin{figure}[h!]
+  \centering
+      \includegraphics[width=0.7\textwidth]{s1_stress.pdf}
+      \includegraphics[width=0.7\textwidth]{s3_stress.pdf}
+  \caption{The day-to-day stress levels reported by two participants (S1 and S3) are shown.
+The y-axis represents values on the 5-point Likert scale we asked participants to respond with ranging from 1/Not at all stressed to 5/Extremely stressed.
+The x-axis represents the day of the study on which the value was recorded (from 0-45).
+\rev{Some gaps are present in the chart for S1 as they did not report their stress level on those days.}}
+   \vspace*{-2mm}
+   \label{fig:dailyStress}
+\end{figure}
+
+\subsubsection{Baseline Stress Levels}
 Common amongst all participants was a trend to select one stress rating
@@
 
-\begin{figure}
-  \centering
-      \includegraphics[width=0.7\textwidth]{dailystress_s1.pdf}
-      \includegraphics[width=0.7\textwidth]{dailystress_s3.pdf}
-  \caption{The day-to-day stress levels reported by two participants (S1 and S3) are shown.
-The y-axis represents values on the 5-point Likert scale we asked participants to respond with ranging from 1/Not at all stressed to 5/Extremely stressed.
-The x-axis represents the day of the study on which the value was recorded (from 0-45).
-The x-axes are slightly different for the 2 participants as they did not begin and end the study on exactly the same days.}
-   \vspace*{-2mm}
-   \label{fig:dailyStress}
-\end{figure}
-
 The baseline stress level varied significantly between participants. 7
@@
 (rating 2) and 1 (8\%) reported feeling no stress at all (rating 1).
+Figure \ref{fig:dailyStress} illustrates these points using \rev{ data from two participants, showing the tendency to report and return to baseline stress levels, as well
+as a distinct difference in baseline stress level (rating 2 for S1 vs rating 3 for S3).}
 
-Figure \ref{fig:dailyStress} illustrates these points, showing both participants tendency to report and return to baseline stress levels, as well
-as a distinct difference in baseline stress level (rating 2 for S1 vs rating 3 for S3).
-
-\subsection{Stressful Days Tend to Cluster}
+\subsubsection{Stressful Days Tend to Cluster}
 Accounting for the variance between participants perceived stress
@@
 
-\subsection{Extreme Changes in Stress Levels are Rare}
+\subsubsection{Extreme Changes in Stress Levels are Rare}
 After accounting for each participant's perceived stress baseline, we
@@
 
-\subsection{Explaining Stress Fluctuations}
-In an attempt to explain some of the stress that our participants were experiencing, we created a linear mixed model with the self-reported daily stress levels as dependent variable and the participants as random effects.
+\subsection{Awakeness}
+\rev{We applied the same analyses described above to the self reported awakeness levels of our participants.
+Compared with the reported stress levels, we observed the same trend of participants reporting one baseline far more commonly than any other.
+We did not find there to be a significant correlation between reported stress and awakeness levels.
+Overall, participants awakeness levels fluctuated significantly less than their stress levels (73.1\% of reports were at the baseline level, compared to 65.1\% for stress, p < 0.05).
+Participants were unlikely to experience days with heightened (above baseline) awakeness.
+Such days made up only 6.0\% of the total observed work days across all participant.
+Most of these days came from one participant, S2, who reported 15 heightened awakeness days compared to the next highest, S13 with 4.
+Large deviations (>1 point deviation) from each participants baseline stress levels were extremely uncommon, accounting for only 9 (2.1\%) of our total observations.
+Similarly to what we observed with high stress days, low awakeness days frequently came in clusters of two or three days in a row.
+Given that the immediately preceding day was a low awakeness day, a given day was 228.2\% more likely to be a low awakeness day than when considering any day at random (p < 0.0001).}
+
+\begin{table}[]
+    \centering
+    \begin{tabular}{|c|c c|c c|}
+    \hline
+        &
+        \multicolumn{2}{c|}{Stress} & 
+        \multicolumn{2}{c|}{Awakeness} \\
+        \hline
+        \textbf{Explanatory Variable} & $\mathbf{R^{2}}$ & \textbf{p-value} &  $\mathbf{R^{2}}$ & \textbf{p-value}\\
+        \hline
+        Day of Week - Monday & -0.033 & 0.745 & -0.058 & 0.0.745\\
+        Day of Week - Tuesday & 0.028 & 0.773 & 0.036 & 0.662 \\
+        Day of Week - Wednesday & 0.163 & 0.100 & -0.018 & 0.830 \\
+        Day of Week - Thursday & 0.022 & 0.821 & 0.076 & 0.927 \\
+        Day of Week - Friday & 0.082 & 0.473 & -0.178 & 0.025 \\
+        Proximity to Month End & 0.044 & 0.687 & -0.082 & 0.374 \\
+        \hline
+    \end{tabular}
+    \caption{Correlation coefficient and associated p-values for the explanatory variables we examined in our linear mixed model analysis.}
+    \label{tab:mixedModel}
+\end{table}
+
+\subsection{Explaining Fluctuations}
+In an attempt to explain some of the \rev{fluctuations in stress and awakeness} that our participants were experiencing, we created a linear mixed model with the self-reported daily stress \rev{and awakeness} levels as dependent variables and the participants as random effects.
 We experimented with day of the week and proximity to beginning or end of month as possible explanatory variables.
-We also calculated daily features from the gathered computer interaction data where available---namely total time spent actively on the computer, percentage of time idle, and amount of time spent on non work-related web browsing per day.
-Ultimately, the analysis showed that none of the variables that we examine had a significant explanatory power with respect to our participants perceived stress levels.
-This indicates that aspects such as the time spent on a computer or the non work-related web browsing do not have a significant impact on the stress level.
-Furthermore, this points to the need for additional instrumentation and data collection, if we are to successfully understand and predict stress in the workplace.
+Ultimately, the analysis showed that none of the variables that we examined had a significant explanatory power with respect to our participants perceived stress levels.
+\rev{For awakeness, we found  that there was a small ($R^2$: -0.178) yet significant decrease in awakeness levels on Fridays in particular.
+Table \ref{tab:mixedModel} shows some of the detailed results of these analyses.} \rev{These results show the difficulty of explaining a person's stress and awakeness via simplistic measures, and point to the need for additional instrumentation and data collection if we are to successfully understand and make predictions about these human aspects in the workplace.}
 
-\section{Predicting Stress in the Moment}
+\section{\rev{Predicting Stress, Focus and Awakeness in the Moment}}
 % random forest best
@@
 for each individual and across all participants. We report on the effectiveness of
-these classifiers and the features that are important in prediction of stress, focus
+these classifiers and the features that are important in predicting stress, focus
 and awakeness.
@@
 
-In this section we describe how we preprocessed the collected data for use in training and testing machine learning models.
+\gcm{To ready the collected data for use in training and testing machine learning models, we performed several steps.}
 
+\rev{
+\subsubsection{Data Cleaning}
+All data recorded by the Everion is associated with a quality score ranging from 0-1 that is calculated using proprietary methods. In accordance with the recommendations of Biovotion,
+to prevent our results from being effected by erroneous data we set a quality threshold of 0.5 and discarded any data gathered which had a quality rating below this threshold. 
+}
 \subsubsection{Data Linking}
@@
 We linked the collected biometric data and survey responses for each participant.
-Linking the data is necessary to construct training and test datasets for use in creating and evaluation machine learning models.
+Linking the data is necessary to construct training and test datasets for use in creating and evaluating machine learning models.
 
-Our linking approach is as follows.
-From the start time of each survey response, we look back one hour for available biometric data.
-For each minute in that hour-long time window, we check for biometric data to associate with the survey response.
-For example, if a participant started a survey response at 11:05am, we look for biometric data in the time frame 10:05am to 11:05am.
-If biometric data is available in the hour-long time window, we consider the survey response to have associated biometric data.
-Otherwise, we exclude the survey response from our dataset.
+\rev{To link the data, we look back one hour from the start time of each survey response for
+available biometric data.
+For example, if a participant started a survey response at 11:05am, we look for biometric data from between 10:05am to 11:05am.
+If no biometric data was recorded in the hour time window, we exclude the survey response from the dataset.
+Otherwise, we consider the survey response to have associated biometric data.}
 
-Reasons for a survey response to lack associated biometric data include:
+There are several reasons for a survey response to lack associated biometric data:
 \begin{itemize}
-\item The participant not wearing the Everion in the hour before beginning the survey
-\item The Everion not recording data in the hour before the participant began the survey (e.g., due to low battery)
-\item Biometric data not being uploaded successfully to the server
+\item The participant was not wearing the Everion in the hour before beginning the survey.
+\item The Everion was not recording data in the hour before the participant began the survey (e.g., due to low battery).
+\item Biometric data was not being uploaded successfully to the server.
 %\item Compatibility issues between the sensor and the OS tracking the participant's data
@@
 From the start time of each survey response, we look back the amount of time 
-that corresponds to each time window, and we create features for all of the 
+that corresponds to each time window and we create features for all of the 
 biometric data available in that time window. For example, if a participant 
@@
 features using all of the available biometric data from 10:35am to 11:05am. 
+\rev{If there is a large portion of data missing (either because of a recording issue or because of low quality data) from the time window considered, then the time window is marked as missing.
+In this case, features are imputed based on the mean of other samples of the same feature for that participant.
+This is an effective and commonly used technique, which is preferable to the alternative of deletion as it preserves our already small sample size~\cite{hawthorne2005imputing}.}
 For each time window, we calculate 10 statistical measurements from the 
@@
 asked. These features are created to predict the responses described by the 
-ground truth.
+ground truth. \rev{To attempt to account for inter-participant differences, we normalized all features on a per-participant basis.}
 
@@
 Most participants did not use all five points of the five-point Likert scale in their responses, and the distributions tend to skew toward one side or the other, depending on the question.
-Thus, we binarized the survey data into a two-point scale to give the machine learning models the best possible chance to make useful predictions.
-The two points in the binary scale represent negative or positive responses for each of the three human aspects of interests (e.g., not stressed or stressed).
+\rev{Given
+this distribution and based on our earlier observation that the participants tended to adhere to a baseline reporting level for stress and awakeness, we elected to simplify the problem from 5 classes to 2.
+This transformation enables us to more easily represent patterns in the data, such as when a participant fluctuates from a normal to high stress level.
+To perform this transformation, we began for
+each participant by calculating the median response value for each question.
+We classified each response below the median as 0 ('negative') and each response above the median as 1 ('positive').
+The distribution for the stress question skewed left, so we included the median values in the 'positive' class, while the distributions for focus and awakeness skewed right, so we included those median values in the 'negative' class.}
 
-We binarized the survey responses as follows.
-For each participant, we calculated the median response value for each question.
-We classified each response below the median as 0 ('negative') and each response above the median as 1 ('positive').
-The distribution for the stress question skewed left, so we included the median values in the 'positive' class, while the distributions for focus and awakeness skewed right, so we included those median values in the 'negative' class.
+With this method, we transformed the survey data into a two-point scale, representing  negative or positive responses for each of the three human aspects of interests (e.g., not stressed or stressed).
 
@@
 This can be seen in the distribution columns in Table \ref{tab:accuracy}.
-To combat this, we applied random oversampling to our training sets, which artificially rebalances the dataset by creating randomly replicated data in the minority class.
-This has been a commonly used technique in previous studies on unbalanced datasets \cite{chawla2004,yap2014}.
+To mitigate this effect, we applied random oversampling to our training sets, which artificially rebalances the dataset by creating randomly replicated data in the minority class.
+This technique has commonly been used  in previous studies on unbalanced datasets \cite{chawla2004,yap2014}.
 
@@
 
+\begin{comment}
 \begin{table}[h]
@@
 \end{table}
+\end{comment}
 
@@
 There are many different algorithms that can be used to build a classifier. To select an
-algorithm, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11} and performed a grid search analysis to determine the optimal hyperparameters for each classifier.
+algorithm, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11},  \rev{evaluating each one by using leave-one-out cross validation}.
 Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, and a multilayer perceptron neural network.
-The optimal values for random forest and the three output measures are listed in Table \ref{tab:hyperparams}.
-For the remainder of this paper, we refer to random forest classifiers trained with these hyperparameters.%\\[-0.1cm]
+For the remainder of this paper, \rev{we refer to a random forest classifier}.%\\[-0.1cm]
 % for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)
@@
   \centering
-  \includegraphics[width=1.0\textwidth]{rq1performance.pdf}
+  \includegraphics[width=1.0\textwidth]{rq1performance_v2.pdf}
   \caption{Results of predictions using the individual models.
-The distribution columns show a bar chart of the response distribution (negative/positive) for each of the three variables.
-The baseline row represents the averaged results of our baseline classifier.
+\rev{The distribution columns show the proportion of the minority class out of the total number of responses for each of the three variables.} The baseline rows represents the averaged results of our baseline classifiers.
 The general row shows the averaged results of our models trained on all participants.}\label{tab:accuracy}%\todo{revise caption}}
@@
 their physiological manifestations) can vary substantially
-(e.g.,~\cite{Hernandez11}), we first trained individual classifiers
+(e.g.,~\cite{Hernandez11}), we first trained \rev{and evaluated individual classifiers
 for each participant (as opposed to a general one for all
-participants). The results of our analysis are reported in
+participants) using leave-one-out cross validation.} The results of our analysis are reported in
 Table~\ref{tab:accuracy}. For our analysis, we report values of
@@
 consider~\cite{yap2014,bhattacharyya_data_2011,Hernandez11}.
+
+\rev{Beside our results, for the purposes of comparison we also present two commonly used baselines in this table - a majority classifier, which always predicts the larger of the two classes considered, and a stratified random classifier which randomly chooses between the two classes, but with a proportional bias towards the larger class.}
+
 For some users (i.e., S1.
@@
 predict all three aspects with reasonable accuracy, precision, and
-recall, as well as to improve upon the baseline---a stratified random
-classifier that randomly chooses one of the two classes with a bias
-towards the larger class. While the individually trained classifiers
+recall.
+\rev{We present a comparison between the averaged results of our individual classifiers and those of the baseline stratified random classifer in the ``Improvement Over Random'' row of Table \ref{tab:accuracy}.
+This is calculated as the difference between the overall average and the baseline results, divided by the baseline results (for example, $\frac{Acc_{Overall} - Acc_{Random}}{Acc_{Random}}$).
+We do not compare our results with the majority classifier directly as this classifier achieved precision and recall scores of zero, making a meaningful comparison impossible.
+The improvement percentages demonstrate that the predictions made by our classifiers are much better than random after correcting for the imbalance in our dataset.}
+
+While the individually trained classifiers
 improved on average across all participants upon the baseline in all
 cases excepting recall of 'not focused', the improvement was
-substantially higher for awakeness (85\% improvement in precision,
-62\% in recall, and 7\% in accuracy) than for stress or focus. In addition,
+substantially higher for awakeness (\rev{52.4\%} improvement in precision,
+\rev{40.8\%} in recall, and \rev{6.5\%} in accuracy) than for stress or focus. In addition,
 the performance of the individually trained classifiers varied greatly
@@
 'stressed', the individual classifiers improved upon the baseline for
-S4, S6, S8, S11, S12, and S14 with a maximum improvement of 88.4\% in
-precision and 184.6\% in recall for S12, while they did worse for S1,
+S4, S6, S8, S11, S12, and S14 with a maximum improvement of \rev{152.0\%} in
+precision and \rev{111.1}\% in recall for S12, while they did worse for S1,
 S3, S5, S7, S9, S10, and S13, and in the worst cases did not correctly
@@
 We found that all methods produced similar results with respect to accuracy, precision, and recall for the individual models.
-Ultimately, we chose to use the top k features with the highest ANOVA F-value, as it is relatively simple and efficient to calculate.
-The values of k used were selected by grid search analysis, and are shown in Table \ref{tab:hyperparams}.
+Ultimately, we \rev{elected not to utilize any feature selection in order to simplify our approach, as we found there to be minimal differences in performance between the techniques, and the random forest algorithm is capable of handling datasets with many features.}
 
@@
 Table~\ref{tab:featureImportance} shows the averaged Gini importance for the feature categories used for predicting stress.
-Particularly important for stress were the feature categories heart rate variability (18.3\%) and skin temperature (15.7\%), both of which have been shown in several previous studies to be  indicators for stress~\cite{dishman2000stress,mcduff16,kataoka00}.
-For predicting focus, the feature categories for blood pulse wave (14.2\%) and heart rate variability (13\%) showed to be the most important categories, while for awakeness the most important ones were heart rate variability (13.6\%) and blood pulse wave (13.1\%).\\[-0.1cm]
+\rev{Heart rate variability proved to be an important measure for all of the aspects of interest, ranking as the most important feature category for both stress and awakeness, and the second most important one for focus.
+This is not surprising, as heart rate variability and skin temperature have been shown in several previous studies to be possible indicators for stress levels~\cite{dishman2000stress,mcduff16,kataoka00}.
+We also found blood pulse wave to be an important indicator for both focus and awakeness, but less important for stress, while respiration rate was important in stress and focus but not awakeness.
+Besides these mentioned feature categories, there was great variation in which measures were important to which aspect.
+This shows that there is a clear benefit to having multiple biometric streams available for predicting these aspects.}\\[-0.1cm]
 
@@
 
-The bottom row of Table \ref{tab:accuracy} presents the averaged performance results for this approach in terms of accuracy, precision, and recall.
-Although the averaged precision and recall are comparable or better than those of the averaged individual results, this was at the cost of a large decrease in overall accuracy.
+The bottom \rev{3 rows} of Table \ref{tab:accuracy} present the averaged performance results for this approach in terms of accuracy, precision, and recall\rev{, as well as the results of baseline stratified random and majority classifiers following the same leave-one-group-out cross-validation procedure.} Although the averaged precision and recall are comparable or better than those of the averaged individual results, this was at the cost of a large decrease in overall accuracy.
 Upon closer investigation into the performance of the general model when testing on each participant, we found that individually trained models for each participant performed much better than a general model trained over all participants.

### 2020-08-07 IJHCSSubmission/bp_studyMethod.tex
Updates from Overleaf

@@
 \section{Field Study}
-We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress over time and the feasibility of predicting stress, focus, and awakeness based on biometric signals.
+We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress\rev{, focus and awakeness} over time and the feasibility of predicting stress, focus, and awakeness \rev{using} biometric signals.
 
 \subsection{Participants}
-We recruited group of 14 professionals via personal contacts from a large power and automation company.
+We recruited a group of 14 professionals via personal contacts from a large power and automation company.
 The group is diverse in terms of age, work experience, gender, and work responsibilities.
 All participants work primarily in an office environment, though half of the participants spend at least 10\% (and up to 50\%) of their time in a laboratory environment.
-Office workers are a population that generalizes to a variety of contexts, and including part-time laboratory workers guarantees that our participants have varying work patterns that include different levels of computer usage, as well as different levels of activity in both individual and collaborative tasks.
+Office workers are a population that generalizes to a variety of contexts, including part-time laboratory workers, and guarantees that our participants have varying work patterns that include different levels of computer usage, as well as different levels of activity in both individual and collaborative tasks.
 
@@
 We performed a study over the course of eight weeks.
-This study includes the collection of the participant's biometric data, computer interaction data, and self-reporting surveys.
-We informed participants about the study purpose and procedure, handed out biometric sensors, and introduced and explained the self-reporting to the participants.
-For 10 of the participants, we further installed a computer activity tracker on their company-issued laptops (note: four participants from the original sample declined for privacy reasons).
+This study includes the collection of the participant's \rev{biometric data and self-reporting surveys during their work}.
+\rev{We also collected computer interaction data, but given privacy concerns of some participants chose not to use this data (see Section~\ref{sec:computer_interaction_data}}).
+We informed participants about the study purpose and procedure, handed out biometric sensors, introduced and explained the self-reporting to the participants, \rev{handed out consent form and ensured informed consent for the study.}%\tf{and asked them to sign a consent form}.%\tfC{DAVE: do you still have these?}.
+\gm{Can we state that it was reviewed by a University ethics board?}
 
-After the initial setup, participants were asked to fill out 3 surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}).
+After the initial setup, participants were asked to fill out three surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}) \rev{and wear the biometric sensor during their work hours}.
 At the end of the eight weeks, we collected the biometric sensors and performed a short follow-up interview on the study and the participants' experiences.
@@
 on the results. As a brief summary, the impact of the
-sessions was minimal; a result, we present an analysis
-of the effects in the Discussion section of the paper.
+sessions was minimal; we present an analysis
+of the effects in Section~\ref{sec:Discussion}.
+
+%\tfC{What's missing is when we asked participant to wear it.
+Chris: do you know what the usual time window of the Everion data was?
+we could state that here or in section 2.3.
+this would mitigate the question on shower usage since I think we only asked them to wear it during working hours, but I'm not sure anymore :(.}
 
@@
 \subsection{Data Collection}
-
-In this section we describe the two datasets that we collected from each study participant.
+\label{sec:DataCol}
+We collected two datasets  from each study participant.
 
@@
 The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/articles/213613165} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements for shorter periods of time or capturing a smaller number of measurements.
+\rev{A comparative study has  shown that the Everion can be used as a valid proxy for HRV metrics for knowledge workers~\cite{barrios2019a}.}
 
@@
 \begin{center}
-\small\addtolength{\tabcolsep}{-1pt}
+\small\addtolength{\tabcolsep}{10pt}%-1pt}
 \begin{tabular}{l l}
@@
 \hspace{3mm}Blood pulse wave & (No Unit)  \\
-\hspace{3mm}Heart rate variability (RMSSD) & Milliseconds (ms) \\
+\hspace{3mm}Heart rate variability: RMSSD  & Milliseconds (ms) \\
+%\multicolumn{2}{l}{\hspace{4mm}\textit{(RMSSD: root mean square of successive heartbeat interval differences)}} \\
 \hspace{3mm}Blood oxygenation  & Percent (\%)  \\
@@
 \end{tabular}
-\caption{Biometric measurements captured by the Everion, organized by category and with references to previous works using similar data}
+\caption{Biometric measurements captured by the Everion, organized by category and with references to previous works using similar data.
+\rev{\textit{(RMSSD denotes the root mean square of successive heartbeat interval differences)}}}
 \label{signals}
@@
 
-\subsubsection{Computer Interaction Data}
-To gain a better understanding of our participants day-to-day work activities, we used an open source computer interaction monitor~\cite{Meyer17}.
+\subsubsection{Computer Interaction Data}\label{sec:computer_interaction_data}
+To gain a better understanding of our participants day-to-day work activities, we \rev{asked participants to install} an open source computer interaction monitor~\cite{Meyer17}.
 The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity.
 Four of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13).
+% Gail: I commented out Thomas' comment as I don't think we need to address it
+%\tfC{Probably needs to state something about how it also captured non-work related browsing (see section 3.4)}
 %Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.
@@
 
+\rev{We did not ask our participants about their focus levels in the end of the day survey as focus
+is more of an in-the-moment aspect than stress and awakeness.}
+
 %\subsection{Intervention}

### 2020-08-07 IJHCSSubmission/bp_threats.tex
Updates from Overleaf

@@
 
+\rev{Due to the small amount of data available to us for the purposes of this study, there is a risk
+that our models may be overfitting to some degree.
+Such overfitting indicates that the results presented in our work may be less than optimal, however any overfitting applies strictly to the testing data and given that there is no overlap between the data used for training and testing, we do not believe this invalidates our results.
+We leave further optimization of our approach to future researchers.}
+
 \subsection{Construct Validity:}
@@
 
-We also perform a grid analysis in our machine learning model
-to pick the number of features that is more predictive
-of the productivity-related indicators, 
-therefore removing the biometric
-features that may be unrelated to the indicators.
+\rev{A
+future more thorough statistical analysis of the relationships between
+the aspects of interest, such as stress, and the physiological
+data may provide deeper insights into the data and
+how it might be used in prediction.}

### 2020-08-07 IJHCSSubmission/elsarticle-template.tex
Updates from Overleaf

@@
 			{#2} $\blacktriangleleft$}}} %\LHD$}}}%
-\newcommand\tf[1]{\cb{TF}{\textcolor{green}{#1}}}
-\newcommand\cs[1]{\cb{CS}{\textcolor{blue}{#1}}}
+\newcommand\tf[1]{\cb{TF}{\textcolor{blue}{#1}}}
+\newcommand\cs[1]{\cb{CS}{\textcolor{cyan}{#1}}}
 \newcommand\ms[1]{\cb{MS}{\textcolor{red}{#1}}}
+\newcommand\gm[1]{\cb{GCM}{\textcolor{magenta}{#1}}}
 \newcommand\tfC[1]{\todo[color=green,inline]{Thomas: #1}}
+\newcommand{\rev}[1]{\textcolor{blue}{#1}}
 
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Observing and Predicting Knowledge Worker Stress in the Wild}
+\title{\rev{Observing and Predicting Knowledge Worker Stress, Focus and Awakeness in the Wild}}
 
@@
 
-%% Group authors per affiliation:
-%\author{Elsevier\fnref{myfootnote}}
-%\address{Radarweg 29, Amsterdam}
-%\fntext[myfootnote]{Since 1880.}
-
-%% or include affiliations in footnotes:
-%\author[mymainaddress,mysecondaryaddress]{Elsevier Inc}
-%\ead[url]{www.elsevier.com}
-
-%\author[mysecondaryaddress]{Global Customer Service
-%\corref{mycorrespondingauthor}}
-%\cortext[mycorrespondingauthor]{Corresponding author}
-%\ead{support@elsevier.com}
-
-%\address[mymainaddress]{1600 John F Kennedy Boulevard, Philadelphia}
-%\address[mysecondaryaddress]{360 Park Avenue South, New York}
-
 \begin{abstract}
@@
 knowledge workers over an eight-week period in which we investigated, using
-experience sampling, how the workers experience stress over time.
-During this field study, we also collected biometric data, which we then used to investigate if it is possible to predict such productivity-related indicators as stress, focus and awakeness, in the moment.
+experience sampling, how the workers experience stress, \rev{focus and awakeness} over time.
+During this field study, we also collected biometric data \rev{including heart- and skin-related measures}, which we then used to investigate if it is possible to predict stress, focus and awakeness, in the moment.
 We observed and report on various trends in knowledge worker stress levels
-over several weeks and show that machine learning models can be built to
-predict stress, focus and awakeness from biometric data.
+over several weeks and show that machine learning models can be built
+\rev{from the data of a single minimally invasive sensor}
+to
+predict stress, focus and awakeness \rev{accurately, with the abstract concept of focus
+the hardest to predict.}
 \end{abstract}
@@
 \begin{keyword}
-Biometrics, Stress, Awakeness, Focus, Computer Interaction, Ubiquitous Computing, Empirical Study, User Centered Design
+Biometrics, \tf{Psycho-Physiological Data}, Stress, Awakeness, Focus, Computer Interaction, Ubiquitous Computing, Empirical Study, User Centered Design
 \end{keyword}
@@
 \input{bp_intro_v4}
+\input{bp_relatedWork}
 \input{bp_studyMethod}
@@
 \input{bp_discussion}
-\input{bp_relatedWork}
 \input{bp_threats}

### 2020-08-07 UbiCompSubmission/bp_results.tex
Updates from Overleaf

%!TEX root = bioPrediction_main.tex
% \section{Analysis and Results}

% \begin{table*}[h]
%   \centering
%   \includegraphics[width=1.0\textwidth]{rq1performance.png}
%   \caption{Results of leave one out evaluation on individual models. The overall and baseline rows contain weighted averages of the individual models, to account for the large imbalance in the data collected
%   }
%    \label{rq1performance}
% \end{table*}

% To evaluate the ability of predicting a knowledge worker's stress, focus, and awakeness in the workplace, we trained and tested machine learning classifiers using a leave-one-out cross validation. In particular, we used the features extracted from the collected data as the input data and binarized the participants' experience samples on stress, focus and awakeness into two classes each (e.g., 'stressed' and 'not stressed').

% % random forest best
% In a first step, we compared multiple classifiers using the popular machine learning library scikit-learn \todo{add ref back in} and performing a grid search analysis to determine the optimal hyperparameters for each classifier. Our analysis showed that random forest outperfoms all other classifiers, including Naive Bayes, decision trees, support vector machine, random forest, and a multilayer perceptron neural network. The optimal values for random forest and the three output measures are listed in Table \todo{fill table below (see comment for values)}. For the remainder of this paper, we will be referring to random forest classifiers trained with these hyperparameters.\\[-0.1cm]
% % for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)

% \begin{table}[H]
% 	\begin{centering}
%     \begin{tabular}{llll}
%       \hline
%       Variable & K & Number of 		Estimators & Minimum Samples to Split \\
%       \hline
%     \end{tabular}
%     \end{centering}
% \end{table}

% \subsection{Predicting Stress, Focus \& Awakeness} \label{secOverallAccuracy}

% Since peoples experience of stress, focus and awakeness as well as their physiological manifestation can vary a lot~\cite{}\todo{find good references here, including https://pdfs.semanticscholar.org/d17d/930333da55860189205707252eef6be6c663.pdf}, we trained individual classifiers for each participant. The results of our analysis are reported in Table~\ref{rq1performance}. For our analysis, we report values of accuracy, one of the most commonly used metric to compare performance, as well as precision and recall of the classes of interest: 'stressed', 'not focused', and 'not awake'. Since the imbalance in the data can lead to high accuracy values if a classifier always just predicts the more likely class while ignoring the class of higher importance and interest, precision and recall of the class of interest can be more adequate metrics in this case~\cite{yap2014,bhattacharyya_data_2011}\todo{also cite: https://pdfs.semanticscholar.org/d17d/930333da55860189205707252eef6be6c663.pdf}. 
% % However, the imbalance in the data can lead to a classifier with high accuracy if it always just predicts the more likely class while ignoring the class of higher importance and interest in our case (i.e. stressed, not focused, not awake). Therefore, we further report the precision and recall for the three classes 'stressed', 'not focused' and 'not awake'.

% Overall, we were able to use the extracted physiological features to predict all three aspects with reasonable accuracy, precision and recall, and improve upon the baseline---a stratified random classifier that randomly chooses one of the two classes with a bias towards the larger class. While the average improvement over the baseline across all participants was quite substantial, especially for awakeness (58\% improvement in precision,  82\% in recall, and 7\% in accuracy) and stress (improvement: 8\% precision, 1\% recall, 8\% accuracy), the individual performance varied greatly among participants. Some participants showed negligible difference in comparison to the baseline classifier, e.g. subject ..., while others showed a large improvement, e.g. subject ...
% % We attribute these discrepancies to the large differences in response distributions between participants, as well as to the subjectivity of self-reporting.
%  \todo{connection to percentage improvement table}

% \begin{table}[h!]
% \begin{centering}
% \begin{tabular}{lll}
% \hline
% Participant & Recall Improvement (\%) & Precision Improvement (\%)\\
% \hline
% S12 & 184.6 & 88.4 \\
% S6 & 66.7 & 85.8 \\
% S11 & 50.0 & 44.4 \\
% S8 & 32.4 & 26.7 \\
% S14 & 14.7 & 9.8 \\
% S4 & 5.6 & 5.5 \\
% S9 & -55.4 & -46.8 \\
% S3 & -90.0 & -82.1 \\
% S1 & -100.0 & -100.0 \\
% S5 & -100.0 & -100.0 \\
% S7 & -100.0 & -100.0 \\
% S10 & -100.0 & -100.0 \\
% S13 & -100.0 & -100.0\\
% S2 & - & -\\
% \hline
% \end{tabular}
% \caption{Percentage improvement in recall and precision for stress, using our approach compared to the baseline, on an individual level. For participant 2, the baseline achieved a precision and recall of 0, thus the improvement is undefined}
% \end{centering}
% \label{tableIndividualImprovement}
% \end{table}

\subsection{Minimum Number of Training Samples} \label{secLearningCurve}

The samples described in this study are costly to obtain since each participant
has to wear a biometric sensors for days and answer
surveys several times a day. Therefore, in the 
interest of minimizing the number of samples required to 
train a model for the 
model to perform adequately, we analyzed the learning
curve of the model with different sets of varying number of 
training samples.

To calculate the learning curve, we start by training the model with the
data obtain from only one survey response, then 
using two responses, then three, and we continue this process
increasing the number of training samples. 
More specifically, we evaluate
the performance of the model using a leave-one-out 
cross validation mechanism, and per each fold of the 
training corpus we select  \textit{n} instances, 
where \textit{n} is the increasing number of training samples.

We are interested in analyzing the highest 
number possible of training samples (survey responses) while
maintaining generalizability to a distinct set of workers.
The participants in this study had different levels of responsiveness to the surveys ranging from 10 to 76 as shown
in Figure~\ref{responseDistribution}. This means that
for the first 10 training sets we could use the
answers of all 14 participants. But for the set of 11 training
samples we could only use 13 participants; continuing
this reasoning, when we reach the set of 76 training samples
would only contain responses from one participant.

We therefore considered for this section of the study the highest number of training samples
with associated biometric data features
where at least ten participants responded as a 
compromise between maintaining a large number of participants for generalizability purposes, but removing the participants that
had the least responsive rate to maximize the number of
training sample sets in the analysis.

As a result of this, we obtained a corpus of analysis of at least 34 survey responses per
each participant. We did not consider for this section
the participants for which 
the number of survey responses were less than 34 (four participants). This way we guarantee that for all the 
analyzed training sets have the same number of participants.

The spectrum of possible training samples analyzed ranges
from 1 to 33 survey responses with their associated 
biometric features.
For each of these values, we performed a leave-one-out
cross validation training on random samples of increasing 
number of training instances, and evaluating
on the left out fold consisting of one response. We then calculate the average
through all the folds of the leave-one-out cross validation.

We are interested in predicting the class for which we have the least
number of instances (e.g. when the participant is stressed, not
awake, or not focused). Even though some participants have a very high
number of overall responses, a very small number of them fit this minority class. Therefore we finally weight each participants' model performance by the percentage of minority class instances
each participant responded.

\begin{figure}
  \centering
          \includegraphics[width=0.5\textwidth]{20180912AwakenessLC2.png}
  \caption{
Performance of the per user trained model measured in precision, recall,
  and f-score. An upwards trending line showing the performance of fscore through the
  different number of training samples is indicated as a dotted red line
  }
   \label{learningCurvet}
\end{figure}

Figure~\ref{learningCurvet} 
shows the performance of a random forest model
trained with the biometric data and predicting the 
left out fold through different numbers training samples.
The data has been binarized as described previously and
the model was trained per each user,
using the survey responses and associated biometric
data of that user. We then evaluated the prediction power for
each of the survey responses of each user, and averaged
the performances for all users.

We show precision, recall, and f-score
to capture the completeness and soundness of our approach 
when predicting the relevant classes, where the relevant
classes are the class for which we have the least number 
of gathered instances (e.g., stressed, not awake, not focused).
This graph shows an overall upwards trend represented by the 
f-score trending line (dotted red) which indicates
that there is a correlation between number of instances trained
on and improvement of the approach's performance.
The trending line shows that there is an improvement of ~9\%
f-score
between training the model with one instance and training
it with 33 instances. Similar results were obtained
when predicting focus and stress.

When contrasting
models trained and tested on the 
biometric data of each user individually, against
a general model trained and tested with the data of all users. 
We can conclude
that for up to 33 training samples analyzed in this study,
the former has better performance when predicting the
responses of each user than when training 
a model with the data of several different users. 
This is partially due to the 
variability across users and the subjective nature
of the responses (e.g.,where slightly awake for one person
may be not awake for another).

\begin{figure}
  \centering
      \includegraphics[width=0.5\textwidth]{20180914Awakeness5PointScaleOnly15Lines.png}
  \caption{
Predicting awakeness in a 5-point Likert scale using root mean squared error 
(lower is better)
which indicates how 
  far from the correct answer are the predictions of the model.
  The error decreases as more training instances are added
  }
   \label{learningCurveb}
\end{figure}

Figure~\ref{learningCurveb} 
was created predicting the values in the 5-point
Likert scale, different from binarized responses
as the previous results.
Since the 5-point likert scale is a spectrum
of values and not a binary response, 
the performance of the approach is 
measured using the root mean squared error (lower is better),
which represents how distant is the 
predicted value from the actual value in a 
the scale.

Figure~\ref{learningCurveb} shows the individual
performance results for each user as straight lines
while the dotted line shows the average of these
performances per user.
The error decreases in an inversely proportional manner
to the number of training samples. The
more training samples add the less likely the 
model will predict a wrong value.
The difference between training with 1 sample (0,49) and training
with 33 samples (0,41) shows a gentle downwards trend in the 
average of all
users.
It is also interesting to notice how the performance results
vary substantially between participants. This means
that for some of the participants the predictions made
by the model are very close most of the time 
to the answer the participant responded
in the survey.

\subsection{Minimum Time Window}
\label{secMinimumTW}
Gathering biometric data is time consuming.
Recognizing which is the minimum time window necessary
to properly predict future responses can make the data
gathering process much faster.
The time windows analyzed refer to the portion of 
time to analyze previous to the start of each survey as 
depicted in Figure~\ref{timeWindows}. These time
windows overlap each with the next (e.g., the 
20 second time window contains all the biometric
data from the 10 second time window, and the
3 hour time window contains the biometric data from
all the previous ones).

\begin{figure}
  \centering
      \includegraphics[width=0.4\textwidth]{timeWindows.png}
  \caption{A set of different time windows to gather biometric data previous to each survey response was chosen for analysis. The 16 time windows selected are: 10sec, 20sec, 30sec, 45sec, 1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, 3hour}
   \label{timeWindows}
\end{figure}

To determine the optimal time window to analyze biometric
data to predict stress, focus, and awakeness, 
we divided our corpus into the features associated with 
each particular time windows under analysis (16 different
time windows).
We then performed a leave-one-out cross validation
using only the features related to each time window.
Since we are only considering a subset of the features
created for each time window, the 
parameter optimization regarding best number of features 
selected, used in the general model, was
not able to be applied in this section.
We followed the procedure described before by weighting
the model of each user by the number of minority
class instances and aggregating the models' weights
according to their percentage in the overall corpus.

\begin{figure}
  \centering
      \includegraphics[width=0.5\textwidth]{20180912AwakenessTWBars.png}
  \caption{Performance of a model trained per user predicting
  awakeness using a random forest using different time windows
  }
   \label{timeWindowsPandR}
\end{figure}

Figure~\ref{timeWindowsPandR} shows the precision, recall,
and f-score when predicting awakeness through 16 
different time windows. 
Different from previous analysis where we considered the
full set of features, in this section we consider only the 
features from a single time window. Overall a very
gentle upwards trend is shown with one point
of inflexion in 45 seconds. The performance
through the different time windows stabilizes close 
to 180 seconds.
The best precision, recall, and f-score 
results from the
leave-one-out cross validation are shown in 
the time window corresponding to 10,800 seconds
(3 hours).

We also compared the performance of a model created per user against
a model created across users.
To calculate the performance of the model created per user,
for each user we create a model using the features 
related to each time window for the selected user only, 
while in the model created with data across users,
we trained the model with the data related to each
particular time window of all users.
our results show that the personalized model created
for each user outperforms in every case
the model created the with data across all users.

This study
shows that in general larger time windows predict stress,
focus, and awakeness more precisely than 
shorter time windows. We can also notice
a very gentle upwards trend
with a inflexion points at 45 seconds and 180.
There is a peak of performance in 45 seconds, 
time window that could be used if time is a scarce resource.
The performance stabilizes around 180 time window
which indicates that collecting data for a longer time
period will not increase significantly the performance of the approach.
The difference in performance between the 
lowest performing and the highest performing time window is 
less than a 10\% improvement.

\subsection{Computer Interaction Data} \label{secCI}

To collect computer interaction data, we used the open source PersonalAnalytics project\footnote{https://github.com/sealuzh/PersonalAnalytics}  \cite{meyer18} to track participant's mouse and keyboard activity, as well as details about their active window. The specifics of the features tracked are listed in Table \ref{tracker}. The tracker was installed on the computers of 10 of the 14 participants, with participants S6, S10, S12, and S13 opting out of this part of the study.

\begin{table}
\begin{center}
\begin{tabular}{l l}
\hline

Feature collected by tool& Description \\ 
\hline
Total keystrokes per min& Sum of all types of keystrokes \\ 
Normal keystrokes per min&F[h] Not backspace and navigation \\ 
Backspace keystrokes per min& Backspace keystrokes \\ 
Navigation keystrokes per min& Arrow key keystrokes \\ 
Total clicks per min& Sum of all click types \\ 
Other clicks per min& Not right and left clicks \\ 
Left clicks per min& Left clicks \\ 
Right clicks per min& Right clicks \\ 
Scrolled distance per min& Scrolled distance in pixels \\ 
Moved distance per min& Mouse movements in pixels \\ 
Activity switches per min& Browser window title changes \\ 
Category switches per min& Activity performed category \\ 
\hline
\end{tabular}
\caption{Features collected per user by the computer interaction tracker~\cite{meyer18}}
\label{tracker}
\end{center}
\end{table}

To create the computer interaction features we selected the aforementioned
16 time windows and scaled the values according to the time window.

We used these computer interaction features to create two new feature sets for each participant: one with only computer interaction features, and one with computer interaction features plus biometric features. Table \ref{ciPerformance} shows the scores that we achieved. We found that in all cases, the computer interaction based model was able to better the results our biometric models. Further, we found that the combined model was the most effective model for predicting stress and awakeness overall, but performed slightly worse than the model using only computer interaction features for focus. Since we consider precision and recall to be the most important statistics when interpreting our results, we compared the models against each other by averaging these two statistics.

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\hline
Model/Feature Set & Accuracy & Precision & Recall\\
\hline
\textbf{Stress}\\
\hspace{3mm}Biometrics only & 0.775 & 0.270 &	0.260\\
\hspace{3mm}C.I. only & 0.698 & 0.290 & 0.272\\
\hspace{3mm}Biometrics + C.I. & 0.712 & 0.317 & 0.286\\
\hline
\textbf{Focus}\\
\hspace{3mm}Biometrics only & 0.716 & 0.251 & 0.256\\
\hspace{3mm}C.I. only & 0.742 & 0.332 & 0.342\\
\hspace{3mm}Biometrics + C.I. & 0.745 & 0.340 & 0.316\\
\hline
\textbf{Awakeness}\\
\hspace{3mm}Biometrics only & 0.808 & 0.269 & 0.314\\
\hspace{3mm}C.I. only & 0.758 & 0.425 & 0.362\\
\hspace{3mm}Biometrics + C.I. & 0.791 & 0.390 & 0.404\\
\hline
\end{tabular}
\caption{A comparison of the results achieved using each of the 3 different feature sets. These values are weighted averages of the individual results from each participant, with more weight given to participants with a larger number of responses. Computer Interactions is abbreviated as C.I. here for readability}
\label{ciPerformance}
\end{center}
\end{table}

As with the biometric models, the individual performance of both the computer interaction only models and the combined models varied quite a bit between participants. Using stress as an example again, in the computer interaction models 5 of the 10 participants saw improvements compared to the baseline, with a maximum improvement of 128\% was seen in precision, and 78\% in recall. In the combined model, referring to stress again, 5 of the 10 participants saw improvements compared to the baseline, with a maximum improvement of 117\% in precision and 95\% in recall. Neither model was capable of correctly predicting any instances of 'stressed' for participant S7.

Since the number of features changes  depending on which feature set is used, we adjusted the feature selection parameter for each of the computer interaction and combined computer interaction/biometric models. The values reported in this section were achieved using the optimal feature selection parameters we found, which are shown in Table \ref{ciFeatureSelection}.

\begin{table}
\begin{center}
\begin{tabular}{lc}
\hline
Model/Feature Set & Number of Features Selected\\
\hline
\textbf{Stress}\\
\hspace{3mm}C.I. & 400\\
\hspace{3mm}Biometrics + C.I. & 800\\
\hline
\textbf{Focus}\\
\hspace{3mm}C.I. & 20\\
\hspace{3mm}Biometrics + C.I. & 300\\
\hline
\textbf{Awakeness}\\
\hspace{3mm}C.I. & All\\
\hspace{3mm}Biometrics + C.I. & 50\\
\hline

\end{tabular}
\caption{The optimal number of features we found to select for each of the model/feature set combinations. Computer interactions is abbreviated as C.I.}
\label{ciFeatureSelection}
\end{center}
\end{table}

### 2020-08-07 UbiCompSubmission/bp_threats.tex
Extra revissions
Updates from Overleaf

--- 2019-09-20 bp_threats.tex
@@
 
-\textbf{Internal Validity:}
+\subsection{Internal Validity:}
 This study tries to find correlations between
@@
 we collected a data for an eight week period, which is
-400\% longer than previous studies~\cite{zuger18,Muller16}.
+400\% longer than the longest previous studies~\cite{zuger18,Muller16}.
 
-\textbf{Construct Validity:}
+\subsection{Construct Validity:}
 A threat to the study is that

### 2020-08-12 IJHCSSubmission/bp_conclusion.tex
Updates from Overleaf

@@
 knowledge workers in their workplace over an eight week period to
-better understand how workers experience \rev{these productivity-related human aspects} over time and to
+better understand how workers experience \rev{these human aspects} over time and to
 examine the ability of biometrics to predict these aspects. The longitudinal and in-situ placement of
 the study support and extend previous work. Based on
-twice-collected daily survey responses, we observed that
-although participants sometimes saw periods of sustained stress,
-they would always return to a baseline stress level for them
+\rev{daily collected} survey responses, we observed that
+although participants sometimes saw periods of sustained stress \rev{or sleepiness},
+they would always return to \rev{their} baseline \rev{reporting} level
 at some point. We also observed that stress levels seldom spiked,

### 2020-08-12 IJHCSSubmission/bp_discussion.tex
Updates from Overleaf

@@
 be seen in Table~\ref{responseDistribution}. While some of the
-imbalance is expected due to certain classes, such as 'not stressed',
+imbalance is expected due to certain classes, such as `not stressed',
 being more common in the workplace, this imbalance also provides

### 2020-08-12 IJHCSSubmission/bp_relatedWork.tex
Updates from Overleaf

@@
 This approach is impractical for use over prolonged periods of time, as in our eight-week study.
-Further, this approach is imprecise because of the delay from the stress stimulation to the stress response, which 
-may take from minutes to hours~\cite{Chandola10,Hellhammer09}.
+Further, this approach is imprecise because of the delay from the stress stimulation to the stress response, which may take from minutes to hours~\cite{Chandola10,Hellhammer09}.
 
-The second approach, which we have chosen for our study, is to measure autonomic nervous system (ANS) activity by analyzing biometric signals of the human body, such as blood pressure, heartbeat, and temperature~\cite{kataoka00,Eekelen04,valentini10}.
+The second approach, which we have chosen for our study, is to measure autonomic nervous system (ANS) activity by analyzing biometric signals of the human body, such as blood pressure, heartbeat, and skin temperature~\cite{kataoka00,Eekelen04,valentini10}.
 In particular, changes in heart rate variability are associated with cognitive and emotional stress~\cite{mcduff16,dishman2000stress}.
@@
 
-Zaman et al.~\cite{Zaman19} developed a study where they measure stress and productivity in short sessions with 63 participants using biometric sensors (thermal facial camera, wrist EDA, chest breathing sensor, and facial camera).
+Zaman et al.~\cite{Zaman19} performed a study in which they measure stress and productivity in short sessions with 63 participants using biometric sensors (thermal facial camera, wrist EDA, chest breathing sensor, and facial camera).
 Different from this study, our corpus is gathered from real-world office workers.
-Likewise, our study spans for eight weeks allowing us to analyze the behavior or our participants through the weeks and providing a more extensive in-depth analysis of our participants over time.
-
-Mirjafari et al.~\cite{Mirjafari19} use mobile sensing data to gather data and build machine learning models to group workers based on their workplace performance metrics.
-Similarly, Sarsenbayeva et al.~\cite{Sarsenbayeva19} study how stress affects mobile interaction reducing completion time and accuracy during target acquisition tasks.
+Also, our study spans for eight weeks allowing us to analyze the behavior or our participants through the weeks and providing a more extensive in-depth analysis of our participants over time.
 
@@
 
-Hernandez et al.~\cite{Hernandez14} investigated the use of pressure-sensitive keyboard and a capacitive mouse as non-intrusive means for measuring computer users stress levels.
+Hernandez et al.~\cite{Hernandez14} investigated the use of a pressure-sensitive keyboard and a capacitive mouse as non-intrusive means for measuring computer users' stress levels.
 They found participants exhibited significantly increased typing pressure and mouse contact when in stressful conditions.
@@
 All of these studies were performed in a controlled lab setting over a short duration and the results have yet to be replicated in the field.
-
-Evans and Johnson~\cite{evans00} investigated the correlation between noise in the workplace and stress levels.
-They found that workers exposed to open-office noise showed aftereffects that indicate motivational deficits.
-The population for their experiment comprised 40 female clerical workers, who were randomly assigned to a control condition or to three-hour low-intensity noise room designed to simulate typical open-office noise levels.
-
-Stress in the workplace and its effects on service providers has been analyzed in call centers~\cite{Hernandez11} and in the context of the perceived imbalance between resources and demands~\cite{cherniss80}.
-These studies considered several factors such as personality traits, career-related goals and attitudes, and life outside of work, and these factors were correlated with stress levels and burnout.
 
@@
 They did not attempt to make observations about the big picture of participants stress profiles or make predictions in real time.
+Also using a wristband, Hernandez et al.~\cite{Hernandez11} studied the stress level during calls in a call center over a seven day period.
+They collected skin conductance measures and examined the interpersonal variability of reporting stress.
 
@@
 The eight week duration gives us authority to speak on the nature of day-to-day fluctuations in knowledge worker stress levels.
+
+\rev{Rather than trying to measure or predict stress, several studies have induced stress and examined its effect on work performance, motivation and others.
+For instance, Sarsenbayeva et al.~\cite{Sarsenbayeva19} induced stress and studied how it affects mobile interaction, showing that it can reduce completion time and accuracy during target acquisition tasks.
+Evans and Johnson~\cite{evans00} investigated the correlation between noise in the workplace and stress levels.
+They found that workers exposed to open-office noise showed aftereffects that indicate motivational deficits but found no difference in cortisol levels.
+The population for their experiment comprised 40 female clerical workers, who were randomly assigned to a control condition or to three-hour low-intensity noise room designed to simulate typical open-office noise levels.}
+
+\rev{Stress in the workplace and its effects on service providers has also been analyzed in call centers~\cite{Hernandez11} and in the context of the perceived imbalance between resources and demands~\cite{cherniss80}.
+These studies considered several factors such as personality traits, career-related goals and attitudes, as well as life outside of work, and examined their correlation with stress levels and burnout.}
+
+\rev{Finally, there is also some work that already used proprietary stress measures for other types of classification.
+For example, Mirjafari et al.~\cite{Mirjafari19} used a proprietary stress measure by Garmin together with other data collected through the sensing of mobile devices and built machine learning models to differentiate between low and high performing workers.}
+
+%\tfC{this is very different}
+%Mirjafari et al.~\cite{Mirjafari19} used mobile devices over a long period to gather data, including a proprietary stress level measure, and built machine learning models to group workers based on their workplace performance metrics.
+%Sarsenbayeva et al.~\cite{Sarsenbayeva19} induced stress and studied how it affects mobile interaction reducing completion time and accuracy during target acquisition tasks.
+
+%Evans and Johnson~\cite{evans00} investigated the correlation between noise in the workplace and stress levels.
+They found that workers exposed to open-office noise showed aftereffects that indicate motivational deficits but found no difference in cortisol levels.
+The population for their experiment comprised 40 female clerical workers, who were randomly assigned to a control condition or to three-hour low-intensity noise room designed to simulate typical open-office noise levels.
+
+%\tfC{following stuff needs to be rewritten, it's just wrong in the first place}
+%When it comes to stress in the workplace, there have also been several studies to analyze its effects. For instance, 
+%Stress in the workplace and its effects on service providers has been analyzed in call centers~\cite{Hernandez11} 
+%and in the context of the perceived imbalance between resources and demands~\cite{cherniss80}.
+These studies considered several factors such as personality traits, career-related goals and attitudes, and life outside of work, and these factors were correlated with stress levels and burnout.
 
@@
 Focus refers to the allocation of limited cognitive processing resources~\cite{Anderson04}.
-Mark et al.~\cite{mark2014bored} studied engagement in workplace activities by analyzing the digital activity of 32 information workers in situ for 5 days to understand how attentional states change with context.
-They found that boredom is highest in the early afternoon and focus peaks in the middle of the afternoon.
+Mark et al.~\cite{mark2014bored} studied \rev{attentional states, including focus, for workplace activities by analyzing the digital activity of 32 information workers in situ for 5 days.} They found that boredom is highest in the early afternoon and focus peaks in the middle of the afternoon.
 They also found that doing work that requires focus correlates with stress, while rote work correlates with happiness.
@@
 This chain is composed by stages of preparation, diversion, resumption and recovery that result in time away from an ongoing task.
+\rev{Since interruptions can have a large impact on the focus and productivity of office workers, several studies have examined the prediction of interruptibility---the availability for interruptions---using a variety of features, including computer interaction and biometrics~\cite{fogarty2005examining, iqbal2008effects, chen2004using, bailey2008understanding, zuger2015interruptibility, zuger18}.
+Most of these studies were again conducted for small and controlled tasks over shorter periods of time.}
 
@@
 \subsection{Biometrics}
-Our study investigates the prediction of multiple productivity-related indicators (stress, focus, and awakeness) using two different types of measurements, and biometric signals over eight weeks in a real-life office setting.
-Existing work~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015} analyzes a broad array of biometric signals and correlates them with individual's cognitive states and processes.
+Our study investigates the prediction of stress, focus, and awakeness using two different types of measurements, and biometric signals over eight weeks in a real-life office setting.
+Existing work~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}\rev{---some of it already mentioned above---}analyzes a broad array of biometric signals and correlates them with individual's cognitive states and processes.
 For example, Zuger et al.~\cite{zuger18} used biometrics to sense interruptibility in the office~\cite{zuger18}.
@@
 Eye tracking technology~\cite{Bednarik2006,Crosby1990,Rodeghero2014} and brain activity~\cite{Ikutani2014,Siegmund2014} 
-have been used in previous studies to analyze different tasks in an office environment.
+have been used in previous studies to examine different tasks in an office environment.
 Eye tracking has been used to analyze memory load and processing load by inspecting task-evoked pupillary response and pupil size~\cite{beatty82}.
@@
 In contrast to studies that use eye tracking or EEG, we focused on a less invasive technology that can be applied in a real world scenario.
-Similarly, we considered adding additional metrics such as Depression, Anxiety and Stress Scale (DASS) and Perceived Stress Scale (PSS)~\cite{Ferdous15} but opted for the ones presented in the paper after several discussions with experts in the area, and after piloting surveys to maximize participant compliance.
+Similarly, we considered adding additional metrics such as the Depression, Anxiety and Stress Scale (DASS) and the Perceived Stress Scale (PSS)~\cite{Ferdous15} but opted for the ones presented in the paper after several discussions with experts in the area, and after piloting surveys to maximize participant compliance.

### 2020-08-12 IJHCSSubmission/bp_results_1.tex
Updates from Overleaf

@@
 emerged. \rev{As we noted in the last section, we did not ask
-participants about focus in the end of the day surveys as focus is an 
+participants about focus in the end of  day surveys as focus is an 
 aspect relevant at a particular moment in time rather than an aspect
@@
 \subsection{Stress Levels}
-
-\rev{We describe characteristics noted in the reporting of the knowledge workers
-about the stress they experienced over the duration of the study.}
+Overall, we identified three prominent characteristics in the stress levels.
+%\rev{We describe characteristics noted in the reporting of the knowledge workers about the stress they experienced over the duration of the study.}
 
@@
 \subsubsection{Baseline Stress Levels}
+%\noindent\textit{\textbf{Baseline Stress Levels.}}
 Common amongst all participants was a trend to select one stress rating
@@
 Figure \ref{fig:dailyStress} illustrates these points using \rev{ data from two participants, showing the tendency to report and return to baseline stress levels, as well
-as a distinct difference in baseline stress level (rating 2 for S1 vs rating 3 for S3).}
+as a distinct difference in baseline stress level (rating 2 for S1 vs rating 3 for S3). Gaps in the chart for S1 represent days for which the participant did not report their stress level.}
 
@@
 Most of these days came from one participant, S2, who reported 15 heightened awakeness days compared to the next highest, S13 with 4.
-Large deviations (>1 point deviation) from each participants baseline stress levels were extremely uncommon, accounting for only 9 (2.1\%) of our total observations.
+Large deviations (>1 point deviation) from each participants baseline awakeness levels were extremely uncommon, accounting for only 9 (2.1\%) of our total observations.
 Similarly to what we observed with high stress days, low awakeness days frequently came in clusters of two or three days in a row.
@@
     \centering
-    \begin{tabular}{|c|c c|c c|}
+\rev{\begin{tabular}{|c|c c|c c|}
     \hline
@@
         \hline
-        Day of Week - Monday & -0.033 & 0.745 & -0.058 & 0.0.745\\
+        Day of Week - Monday & -0.033 & 0.745 & -0.058 & 0.497\\
         Day of Week - Tuesday & 0.028 & 0.773 & 0.036 & 0.662 \\
@@
     \caption{Correlation coefficient and associated p-values for the explanatory variables we examined in our linear mixed model analysis.}
-    \label{tab:mixedModel}
+    \label{tab:mixedModel}}
 \end{table}
@@
 \rev{For awakeness, we found  that there was a small ($R^2$: -0.178) yet significant decrease in awakeness levels on Fridays in particular.
-Table \ref{tab:mixedModel} shows some of the detailed results of these analyses.} \rev{These results show the difficulty of explaining a person's stress and awakeness via simplistic measures, and point to the need for additional instrumentation and data collection if we are to successfully understand and make predictions about these human aspects in the workplace.}
+Table \ref{tab:mixedModel} shows some of the detailed results of these analyses.} \rev{These results show the difficulty of explaining a person's stress and awakeness via simple measures, and point to the need for additional instrumentation and data collection if we are to successfully understand and make predictions about these human aspects in the workplace.}
 
@@
 
-\gcm{To ready the collected data for use in training and testing machine learning models, we performed several steps.}
+\rev{To ready the collected data for use in training and testing machine learning models, we performed several steps.}
 
@@
 features using all of the available biometric data from 10:35am to 11:05am. 
-\rev{If there is a large portion of data missing (either because of a recording issue or because of low quality data) from the time window considered, then the time window is marked as missing.
+\rev{If there is a large portion ($\geq$50\%) of data missing (either because of a recording issue or because of low quality data) from the time window considered, then the time window is marked as missing.
 In this case, features are imputed based on the mean of other samples of the same feature for that participant.
@@
 
+\begin{table}
+  \centering
+      \includegraphics[width=0.8\textwidth]{distributiontable.pdf}
+  \caption{The distribution of the responses of each participant to the three questions asked during the day are shown.
+Each bar in the histograms represent one of the 5 values on the 5-point Likert scale we asked participants to respond with, where the far left side of the histograms are 1/Not at all, and the far right sides are 5/Extremely}
+   \label{responseDistribution}
+   %\vspace*{-2mm}
+\end{table}
+
 \subsubsection{Response Transformations}
@@
 each participant by calculating the median response value for each question.
-We classified each response below the median as 0 ('negative') and each response above the median as 1 ('positive').
-The distribution for the stress question skewed left, so we included the median values in the 'positive' class, while the distributions for focus and awakeness skewed right, so we included those median values in the 'negative' class.}
+We classified each response below the median as 0 (`negative') and each response above the median as 1 (`positive').
+The distribution for the stress question skewed left, so we included the median values in the `positive' class (i.e.
+`stressed'), while the distributions for focus and awakeness skewed right, so we included those median values in the `negative' class (i.e.
+`not focused', `not awake').}
 
@@
 This technique has commonly been used  in previous studies on unbalanced datasets \cite{chawla2004,yap2014}.
-
-\begin{table}
-  \centering
-      \includegraphics[width=0.8\textwidth]{distributiontable.pdf}
-  \caption{The distribution of the responses of each participant to the three questions asked during the day are shown.
-Each bar in the histograms represent one of the 5 values on the 5-point Likert scale we asked participants to respond with, where the far left side of the histograms are 1/Not at all, and the far right sides are 5/Extremely}
-   \label{responseDistribution}
-   %\vspace*{-2mm}
-\end{table}
 
@@
 as well as precision and recall of the classes of interest:
-'stressed', 'not focused', and 'not awake'. Since the imbalance in the
+`stressed', `not focused', and `not awake'. Since the imbalance in the
 data can lead to high accuracy values if a classifier always just
@@
 This is calculated as the difference between the overall average and the baseline results, divided by the baseline results (for example, $\frac{Acc_{Overall} - Acc_{Random}}{Acc_{Random}}$).
-We do not compare our results with the majority classifier directly as this classifier achieved precision and recall scores of zero, making a meaningful comparison impossible.
+We do not compare our results with the majority classifier directly as this classifier achieved precision and recall scores of zero when predicting stress, lack of focus, and lack of awakeness, making a meaningful comparison impossible.
 The improvement percentages demonstrate that the predictions made by our classifiers are much better than random after correcting for the imbalance in our dataset.}
@@
 improved on average across all participants upon the baseline in all
-cases excepting recall of 'not focused', the improvement was
+cases except in recall of `not focused', the improvement was
 substantially higher for awakeness (\rev{52.4\%} improvement in precision,
@@
 individually trained classifier. For instance, for predicting
-'stressed', the individual classifiers improved upon the baseline for
+`stressed', the individual classifiers improved upon the baseline for
 S4, S6, S8, S11, S12, and S14 with a maximum improvement of \rev{152.0\%} in
@@
 S3, S5, S7, S9, S10, and S13, and in the worst cases did not correctly
-predict a single instance of 'stressed'. Typically, users that have the
+predict a single instance of `stressed'. Typically, users that have the
 lowest precision and recall values are those where the data is the
-most unbalanced.\\[-0.1cm]
+most imbalanced.\\[-0.1cm]
 % We attribute these discrepancies to the large differences in response distributions between participants, as well as to the subjectivity of self-reporting.
@@
 We found that all methods produced similar results with respect to accuracy, precision, and recall for the individual models.
-Ultimately, we \rev{elected not to utilize any feature selection in order to simplify our approach, as we found there to be minimal differences in performance between the techniques, and the random forest algorithm is capable of handling datasets with many features.}
+Ultimately, we \rev{elected not to utilize any feature selection in order to simplify our approach, as we found there to be minimal differences in performance between the techniques, and the random forest algorithm is capable of (and robust for) handling datasets with many features.}
 
@@
 We also found blood pulse wave to be an important indicator for both focus and awakeness, but less important for stress, while respiration rate was important in stress and focus but not awakeness.
-Besides these mentioned feature categories, there was great variation in which measures were important to which aspect.
-This shows that there is a clear benefit to having multiple biometric streams available for predicting these aspects.}\\[-0.1cm]
+Besides these mentioned feature categories, there was great variation in which measures were important to which of the three aspects.
+This shows that there is a clear benefit to having multiple biometric streams available for predicting stress, focus and awakeness.}\\[-0.1cm]
 
@@
 Due to the larger amount of data available in the general case, we used the more common random undersampling, which randomly selects elements in the majority class to exclude from the dataset, instead of random oversampling to balance the distribution of the dataset.
-The models were trained on the datasets of 13 of the 14 participants, and then evaluated on the dataset of the last (leave-one-group-out cross-validation), repeating this process for all 14 participants.
+The models were trained on the datasets of 13 of the 14 participants, and then evaluated on the dataset of the last (leave-one-participant-out cross-validation), repeating this process for all 14 participants.
 
@@
 Upon closer investigation into the performance of the general model when testing on each participant, we found that individually trained models for each participant performed much better than a general model trained over all participants.
-Using stress as an example, for participant S12, for whom we saw the greatest increase compared to the baseline in individual models, the general model was unable to predict a single instance of 'stressed' correctly.
+Using stress as an example, for participant S12, for whom we saw the greatest increase compared to the baseline in individual models, the general model was unable to predict a single instance of `stressed' correctly.
 This is consistent with our expectations because biometric features are highly specific to individuals.

### 2020-08-12 IJHCSSubmission/bp_studyMethod.tex
Updates from Overleaf

@@
 \section{Field Study}
-We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress\rev{, focus and awakeness} over time and the feasibility of predicting stress, focus, and awakeness \rev{using} biometric signals.
+We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress \rev{and awakeness} over time and the feasibility of predicting stress, focus, and awakeness \rev{using} biometric signals.
 
@@
 The average participant age is ~40, with 5 in the age range 25-34, 7 in the age range 35-44, and 2 in the age range 55+.
-The average number of years of professional experience of the participants is ~12, with 2 having less than 5 years, 10 having 5-15 years, and 2 having more than 25 years.
+The average number of years of professional experience of the participants is 12, with 2 having less than 5 years, 10 having 5-15 years, and 2 having more than 25 years.
 All participants work for a research organization within the company, but their job functions span line management, laboratory science, scientific research, technology evaluation, and software development.
@@
 We performed a study over the course of eight weeks.
-This study includes the collection of the participant's \rev{biometric data and self-reporting surveys during their work}.
-\rev{We also collected computer interaction data, but given privacy concerns of some participants chose not to use this data (see Section~\ref{sec:computer_interaction_data}}).
-We informed participants about the study purpose and procedure, handed out biometric sensors, introduced and explained the self-reporting to the participants, \rev{handed out consent form and ensured informed consent for the study.}%\tf{and asked them to sign a consent form}.%\tfC{DAVE: do you still have these?}.
-\gm{Can we state that it was reviewed by a University ethics board?}
+This study includes the collection of the participant's \rev{biometric data and self-reporting surveys during their work hours}.
+\rev{We also collected computer interaction data, but given privacy concerns of some participants chose not to use this data (see Section~\ref{sec:computer_interaction_data})}.
+We informed participants about the study purpose and procedure, handed out biometric sensors, introduced and explained the self-reporting to the participants, \rev{handed out consent forms and ensured informed consent from each participant for the study.}
 
@@
 Data collected by the Everion are uploaded to a server, from which we downloaded the data for use in our study.
+\rev{We chose these biometric measurements based on previous research that has indicated their potential (see references in Table~\ref{signals}) and the availability and feasibility of collecting these measurements with low invasiveness over a long duration.}
 

### 2020-08-12 IJHCSSubmission/bp_threats.tex
Updates from Overleaf

--- 2020-03-29 IJHCSSubmission/bp_threats.tex
@@
 
-\subsection{External Validity:}
+\subsection{External Validity}
 The results of our study may 
@@
 
-\subsection{Internal Validity:}
+\subsection{Internal Validity}
 This study tries to find correlations between
-biometric features and different productivity-related indicators (stress, focus, and awakeness).
+biometric features and \rev{the human aspects of} stress, focus, and awakeness.
 Nonetheless, biometric signals are influenced by far more
@@
 Therefore, trying to draw a strong causality between the biometric
-features and the productivity-related aspects would be inaccurate.
+features and the aspects would be inaccurate.
 To mitigate this risk,  we collected the data
@@
 
-\subsection{Construct Validity:}
+\rev{Due to the small amount of data available to us for the purposes of this study, there is a risk
+that our models may be overfitting to some degree.
+Such overfitting indicates that the results presented in our work may be less than optimal, however any overfitting applies strictly to the training data and given that there is no overlap between the data used for training and testing, we do not believe this invalidates our results.
+We leave further optimization of our approach to future researchers.}
+
+\subsection{Construct Validity}
 A threat to the study is that
@@
 biometrics that historically have shown correlation with 
-productivity aspects from each user.
-In an effort to maintain this research applicable to real-world environments, we picked this already existing device, even when, as a trade-off, we could not capture more descriptive and more intrusive signals such as SDNN, SCL, SCR, eye tracking, or brain activity.
+\rev{the studied human aspects of stress, focus and awakeness}.
+In an effort to maintain this research applicable to real-world environments, we picked \rev{the already existing Everion device}, even when, as a trade-off, we could not capture more descriptive and more intrusive signals such as SDNN, SCL, SCR, eye tracking, or brain activity.
 
-We also perform a grid analysis in our machine learning model
-to pick the number of features that is more predictive
-of the productivity-related indicators, 
-therefore removing the biometric
-features that may be unrelated to the indicators.
+\rev{A future more thorough statistical analysis of the relationships between
+the aspects of interest, such as stress, and the physiological
+data may further provide deeper insights into the data and
+how it might be used in prediction.}

### 2020-08-12 IJHCSSubmission/elsarticle-template.tex
Updates from Overleaf

@@
 			{#2} $\blacktriangleleft$}}} %\LHD$}}}%
-\newcommand\tf[1]{\cb{TF}{\textcolor{blue}{#1}}}
+\newcommand\tf[1]{\cb{TF}{\textcolor{green}{#1}}}
 \newcommand\cs[1]{\cb{CS}{\textcolor{cyan}{#1}}}
@@
 knowledge workers over an eight-week period in which we investigated, using
-experience sampling, how the workers experience stress, \rev{focus and awakeness} over time.
+experience sampling, how the workers experience stress, \rev{awakeness} over time.
 During this field study, we also collected biometric data \rev{including heart- and skin-related measures}, which we then used to investigate if it is possible to predict stress, focus and awakeness, in the moment.
-We observed and report on various trends in knowledge worker stress levels
-over several weeks and show that machine learning models can be built
+We observed and report on various trends in knowledge worker stress \rev{and awakeness} levels
+over several weeks\rev{, finding that people tend to have certain ``baseline'' levels for these aspects and stressful days or days with low awakeness tend to cluster.
+We further show} that machine learning models can be built
 \rev{from the data of a single minimally invasive sensor}
 to
-predict stress, focus and awakeness \rev{accurately, with the abstract concept of focus
-the hardest to predict.}
+predict stress, focus and awakeness.
+\rev{Overall, we found that our models were capable of large improvements in precision and recall in comparison to a random classifier for stress (25.9\% increase over random for precision, 4.2\% for recall) and awakeness (52.4\% increase in precision, 40.8\% in recall).
+The abstract concept of focus proved to be
+the hardest to predict (26.0\% increase in precision, 27.8\% decrease in recall).} 
 \end{abstract}
@@
 \begin{keyword}
-Biometrics, \tf{Psycho-Physiological Data}, Stress, Awakeness, Focus, Computer Interaction, Ubiquitous Computing, Empirical Study, User Centered Design
+Biometrics, \rev{Psycho-Physiological Data}, Stress, Awakeness, Focus, Computer Interaction, Ubiquitous Computing, Empirical Study, User Centered Design
 \end{keyword}

### 2020-10-01 IJHCSSubmission/bp_conclusion.tex
removing the past revision blue marks and starting with new ones for this coming revision

--- 2020-03-29 IJHCSSubmission/bp_conclusion.tex
@@
 knowledge workers in their workplace over an eight week period to
-better understand how workers experience stress over time and to
-examine the ability of biometrics to predict these
-productivity-related aspects. The longitudinal and in-situ placement of
+better understand how workers experience these human aspects over time and to
+examine the ability of biometrics to predict these aspects. The longitudinal and in-situ placement of
 the study support and extend previous work. Based on
-twice-collected daily survey responses, we observed that
-although participants sometimes saw periods of sustained stress,
-they would always return to a baseline stress level for them
+daily collected survey responses, we observed that
+although participants sometimes saw periods of sustained stress or sleepiness,
+they would always return to their baseline reporting level
 at some point. We also observed that stress levels seldom spiked,
@@
 biometric data with which we were able to create a model that is able to
-predict user stress, sleepiness, and lack of focus with an up to 84.65\% improvement over the baseline.
+predict user stress, sleepiness, and lack of focus with small improvements in accuracy (from 3.1\% to 7.2\%, depending on the aspect in question), and moderate improvements in precision (25.9\% - 52.4\%) in comparison to a stratified random classifier.
+While the precision and recall scores we report are still low overall, this is a difficult problem to solve and our improvements indicate the potential for future researchers to build upon.
 

### 2020-10-01 IJHCSSubmission/bp_discussion.tex
removing the past revision blue marks and starting with new ones for this coming revision

@@
 the individual models and undersampling in case of a general model
-where more data was available. \rev{Rebalancing the dataset using such techniques
+where more data was available. Rebalancing the dataset using such techniques
 is a common and effective practice~\cite{branco2016survey}.
@@
 Oversampling the dataset as we did may also lead to an increased risk of model overfitting.
-However, we believe the benefits of rebalancing the dataset outweigh this possible downside.}
+However, we believe the benefits of rebalancing the dataset outweigh this possible downside.
 

### 2020-10-01 IJHCSSubmission/bp_relatedWork.tex
removing the past revision blue marks and starting with new ones for this coming revision

@@
 The study and prediction approach used in this paper is related
-to previous studies of \rev{stress, focus, and awakeness} and studies
+to previous studies of stress, focus, and awakeness and studies
 using biometrics to predict them. We consider
@@
 
-\rev{Rather than trying to measure or predict stress, several studies have induced stress and examined its effect on work performance, motivation and others.
+Rather than trying to measure or predict stress, several studies have induced stress and examined its effect on work performance, motivation and others.
 For instance, Sarsenbayeva et al.~\cite{Sarsenbayeva19} induced stress and studied how it affects mobile interaction, showing that it can reduce completion time and accuracy during target acquisition tasks.
@@
 They found that workers exposed to open-office noise showed aftereffects that indicate motivational deficits but found no difference in cortisol levels.
-The population for their experiment comprised 40 female clerical workers, who were randomly assigned to a control condition or to three-hour low-intensity noise room designed to simulate typical open-office noise levels.}
+The population for their experiment comprised 40 female clerical workers, who were randomly assigned to a control condition or to three-hour low-intensity noise room designed to simulate typical open-office noise levels.
 
-\rev{Stress in the workplace and its effects on service providers has also been analyzed in call centers~\cite{Hernandez11} and in the context of the perceived imbalance between resources and demands~\cite{cherniss80}.
-These studies considered several factors such as personality traits, career-related goals and attitudes, as well as life outside of work, and examined their correlation with stress levels and burnout.}
+Stress in the workplace and its effects on service providers has also been analyzed in call centers~\cite{Hernandez11} and in the context of the perceived imbalance between resources and demands~\cite{cherniss80}.
+These studies considered several factors such as personality traits, career-related goals and attitudes, as well as life outside of work, and examined their correlation with stress levels and burnout.
 
-\rev{Finally, there is also some work that already used proprietary stress measures for other types of classification.
-For example, Mirjafari et al.~\cite{Mirjafari19} used a proprietary stress measure by Garmin together with other data collected through the sensing of mobile devices and built machine learning models to differentiate between low and high performing workers.}
+Finally, there is also some work that already used proprietary stress measures for other types of classification.
+For example, Mirjafari et al.~\cite{Mirjafari19} used a proprietary stress measure by Garmin together with other data collected through the sensing of mobile devices and built machine learning models to differentiate between low and high performing workers.
 
@@
 Focus refers to the allocation of limited cognitive processing resources~\cite{Anderson04}.
-Mark et al.~\cite{mark2014bored} studied \rev{attentional states, including focus, for workplace activities by analyzing the digital activity of 32 information workers in situ for 5 days.} They found that boredom is highest in the early afternoon and focus peaks in the middle of the afternoon.
+Mark et al.~\cite{mark2014bored} studied attentional states, including focus, for workplace activities by analyzing the digital activity of 32 information workers in situ for 5 days.
+They found that boredom is highest in the early afternoon and focus peaks in the middle of the afternoon.
 They also found that doing work that requires focus correlates with stress, while rote work correlates with happiness.
@@
 This chain is composed by stages of preparation, diversion, resumption and recovery that result in time away from an ongoing task.
-\rev{Since interruptions can have a large impact on the focus and productivity of office workers, several studies have examined the prediction of interruptibility---the availability for interruptions---using a variety of features, including computer interaction and biometrics~\cite{fogarty2005examining, iqbal2008effects, chen2004using, bailey2008understanding, zuger2015interruptibility, zuger18}.
-Most of these studies were again conducted for small and controlled tasks over shorter periods of time.}
+Since interruptions can have a large impact on the focus and productivity of office workers, several studies have examined the prediction of interruptibility---the availability for interruptions---using a variety of features, including computer interaction and biometrics~\cite{fogarty2005examining, iqbal2008effects, chen2004using, bailey2008understanding, zuger2015interruptibility, zuger18}.
+Most of these studies were again conducted for small and controlled tasks over shorter periods of time.
 
@@
 Our study investigates the prediction of stress, focus, and awakeness using two different types of measurements, and biometric signals over eight weeks in a real-life office setting.
-Existing work~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}\rev{---some of it already mentioned above---}analyzes a broad array of biometric signals and correlates them with individual's cognitive states and processes.
+Existing work~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}---some of it already mentioned above--- analyzes a broad array of biometric signals and correlates them with individual's cognitive states and processes.
 For example, Zuger et al.~\cite{zuger18} used biometrics to sense interruptibility in the office~\cite{zuger18}.

### 2020-10-01 IJHCSSubmission/bp_results_1.tex
removing the past revision blue marks and starting with new ones for this coming revision

@@
 
-\section{Observed Trends Over Time in Stress \rev{and Awakeness} Levels}
+\section{Observed Trends Over Time in Stress and Awakeness Levels}
 \label{stressTrends}
 
-To gain insights into how knowledge workers experience stress \rev{and awakeness} over an
+To gain insights into how knowledge workers experience stress and awakeness over an
 extended period of work, we examined the end of day survey responses
 collected from each participant to see if any identifiable trends
-emerged. \rev{As we noted in the last section, we did not ask
+emerged. As we noted in the last section, we did not ask
 participants about focus in the end of  day surveys as focus is an 
 aspect relevant at a particular moment in time rather than an aspect
-for an extended period of work.} We used data from 13 of the 14 participants - we excluded one participant
+for an extended period of work. We used data from 13 of the 14 participants - we excluded one participant
 from this analysis as they experienced atypical stress
@@
 The x-axis represents the day of the study on which the value was recorded (from 0-45).
-\rev{Some gaps are present in the chart for S1 as they did not report their stress level on those days.}}
+Some gaps are present in the chart for S1 as they did not report their stress level on those days.}
    \vspace*{-2mm}
@@
 (rating 2) and 1 (8\%) reported feeling no stress at all (rating 1).
-Figure \ref{fig:dailyStress} illustrates these points using \rev{ data from two participants, showing the tendency to report and return to baseline stress levels, as well
-as a distinct difference in baseline stress level (rating 2 for S1 vs rating 3 for S3). Gaps in the chart for S1 represent days for which the participant did not report their stress level.}
+Figure \ref{fig:dailyStress} illustrates these points using  data from two participants, showing the tendency to report and return to baseline stress levels, as well
+as a distinct difference in baseline stress level (rating 2 for S1 vs rating 3 for S3). Gaps in the chart for S1 represent days for which the participant did not report their stress level.
 
@@
 \subsection{Awakeness}
-\rev{We applied the same analyses described above to the self reported awakeness levels of our participants.
+We applied the same analyses described above to the self reported awakeness levels of our participants.
 Compared with the reported stress levels, we observed the same trend of participants reporting one baseline far more commonly than any other.
@@
 Similarly to what we observed with high stress days, low awakeness days frequently came in clusters of two or three days in a row.
-Given that the immediately preceding day was a low awakeness day, a given day was 228.2\% more likely to be a low awakeness day than when considering any day at random (p < 0.0001).}
+Given that the immediately preceding day was a low awakeness day, a given day was 228.2\% more likely to be a low awakeness day than when considering any day at random (p < 0.0001).
 
@@
     \centering
-\rev{\begin{tabular}{|c|c c|c c|}
+\begin{tabular}{|c|c c|c c|}
     \hline
@@
     \caption{Correlation coefficient and associated p-values for the explanatory variables we examined in our linear mixed model analysis.}
-    \label{tab:mixedModel}}
+    \label{tab:mixedModel}
 \end{table}
@@
 \subsection{Explaining Fluctuations}
-In an attempt to explain some of the \rev{fluctuations in stress and awakeness} that our participants were experiencing, we created a linear mixed model with the self-reported daily stress \rev{and awakeness} levels as dependent variables and the participants as random effects.
+In an attempt to explain some of the fluctuations in stress and awakeness that our participants were experiencing, we created a linear mixed model with the self-reported daily stress and awakeness levels as dependent variables and the participants as random effects.
 We experimented with day of the week and proximity to beginning or end of month as possible explanatory variables.
 Ultimately, the analysis showed that none of the variables that we examined had a significant explanatory power with respect to our participants perceived stress levels.
-\rev{For awakeness, we found  that there was a small ($R^2$: -0.178) yet significant decrease in awakeness levels on Fridays in particular.
-Table \ref{tab:mixedModel} shows some of the detailed results of these analyses.} \rev{These results show the difficulty of explaining a person's stress and awakeness via simple measures, and point to the need for additional instrumentation and data collection if we are to successfully understand and make predictions about these human aspects in the workplace.}
+For awakeness, we found  that there was a small ($R^2$: -0.178) yet significant decrease in awakeness levels on Fridays in particular.
+Table \ref{tab:mixedModel} shows some of the detailed results of these analyses.
+These results show the difficulty of explaining a person's stress and awakeness via simple measures, and point to the need for additional instrumentation and data collection if we are to successfully understand and make predictions about these human aspects in the workplace.
 
-\section{\rev{Predicting Stress, Focus and Awakeness in the Moment}}
+\section{Predicting Stress, Focus and Awakeness in the Moment}
 % random forest best
@@
 
-\rev{To ready the collected data for use in training and testing machine learning models, we performed several steps.}
+\rev{In a machine learning context, data preparation and utilization is an essential part of the proposed solution.
+To prepare} the collected data for use in training and testing \rev{our proposed} machine learning models, we performed \rev{the following} steps.
 
-\rev{
 \subsubsection{Data Cleaning}
@@
 to prevent our results from being effected by erroneous data we set a quality threshold of 0.5 and discarded any data gathered which had a quality rating below this threshold.
-}
+
 \subsubsection{Data Linking}
@@
 
-\rev{To link the data, we look back one hour from the start time of each survey response for
+To link the data, we look back one hour from the start time of each survey response for
 available biometric data.
@@
 If no biometric data was recorded in the hour time window, we exclude the survey response from the dataset.
-Otherwise, we consider the survey response to have associated biometric data.}
+Otherwise, we consider the survey response to have associated biometric data.
 
@@
 features using all of the available biometric data from 10:35am to 11:05am. 
-\rev{If there is a large portion ($\geq$50\%) of data missing (either because of a recording issue or because of low quality data) from the time window considered, then the time window is marked as missing.
+If there is a large portion ($\geq$50\%) of data missing (either because of a recording issue or because of low quality data) from the time window considered, then the time window is marked as missing.
 In this case, features are imputed based on the mean of other samples of the same feature for that participant.
-This is an effective and commonly used technique, which is preferable to the alternative of deletion as it preserves our already small sample size~\cite{hawthorne2005imputing}.}
+This is an effective and commonly used technique, which is preferable to the alternative of deletion as it preserves our already small sample size~\cite{hawthorne2005imputing}.
 For each time window, we calculate 10 statistical measurements from the 
@@
 asked. These features are created to predict the responses described by the 
-ground truth. \rev{To attempt to account for inter-participant differences, we normalized all features on a per-participant basis.}
+ground truth. To attempt to account for inter-participant differences, we normalized all features on a per-participant basis.
 
@@
 Most participants did not use all five points of the five-point Likert scale in their responses, and the distributions tend to skew toward one side or the other, depending on the question.
-\rev{Given
+Given
 this distribution and based on our earlier observation that the participants tended to adhere to a baseline reporting level for stress and awakeness, we elected to simplify the problem from 5 classes to 2.
@@
 `stressed'), while the distributions for focus and awakeness skewed right, so we included those median values in the `negative' class (i.e.
-`not focused', `not awake').}
+`not focused', `not awake').
 
@@
 There are many different algorithms that can be used to build a classifier. To select an
-algorithm, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11},  \rev{evaluating each one by using leave-one-out cross validation}.
+algorithm, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11},  evaluating each one by using leave-one-out cross validation.
 Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, and a multilayer perceptron neural network.
-For the remainder of this paper, \rev{we refer to a random forest classifier}.%\\[-0.1cm]
+For the remainder of this paper, we refer to a random forest classifier.%\\[-0.1cm]
 % for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)
@@
   \caption{Results of predictions using the individual models.
-\rev{The distribution columns show the proportion of the minority class out of the total number of responses for each of the three variables.} The baseline rows represents the averaged results of our baseline classifiers.
+The distribution columns show the proportion of the minority class out of the total number of responses for each of the three variables.
+The baseline rows represents the averaged results of our baseline classifiers.
 The general row shows the averaged results of our models trained on all participants.}\label{tab:accuracy}%\todo{revise caption}}
@@
 their physiological manifestations) can vary substantially
-(e.g.,~\cite{Hernandez11}), we first trained \rev{and evaluated individual classifiers
+(e.g.,~\cite{Hernandez11}), we first trained and evaluated individual classifiers
 for each participant (as opposed to a general one for all
-participants) using leave-one-out cross validation.} The results of our analysis are reported in
+participants) using leave-one-out cross validation. The results of our analysis are reported in
 Table~\ref{tab:accuracy}. For our analysis, we report values of
@@
 
-\rev{Beside our results, for the purposes of comparison we also present two commonly used baselines in this table - a majority classifier, which always predicts the larger of the two classes considered, and a stratified random classifier which randomly chooses between the two classes, but with a proportional bias towards the larger class.}
+Beside our results, for the purposes of comparison we also present two commonly used baselines in this table - a majority classifier, which always predicts the larger of the two classes considered, and a stratified random classifier which randomly chooses between the two classes, but with a proportional bias towards the larger class.
 
@@
 recall.
-\rev{We present a comparison between the averaged results of our individual classifiers and those of the baseline stratified random classifer in the ``Improvement Over Random'' row of Table \ref{tab:accuracy}.
+We present a comparison between the averaged results of our individual classifiers and those of the baseline stratified random classifer in the ``Improvement Over Random'' row of Table \ref{tab:accuracy}.
 This is calculated as the difference between the overall average and the baseline results, divided by the baseline results (for example, $\frac{Acc_{Overall} - Acc_{Random}}{Acc_{Random}}$).
 We do not compare our results with the majority classifier directly as this classifier achieved precision and recall scores of zero when predicting stress, lack of focus, and lack of awakeness, making a meaningful comparison impossible.
-The improvement percentages demonstrate that the predictions made by our classifiers are much better than random after correcting for the imbalance in our dataset.}
+The improvement percentages demonstrate that the predictions made by our classifiers are much better than random after correcting for the imbalance in our dataset.
 
@@
 cases except in recall of `not focused', the improvement was
-substantially higher for awakeness (\rev{52.4\%} improvement in precision,
-\rev{40.8\%} in recall, and \rev{6.5\%} in accuracy) than for stress or focus. In addition,
+substantially higher for awakeness (52.4\% improvement in precision,
+40.8\% in recall, and 6.5\% in accuracy) than for stress or focus. In addition,
 the performance of the individually trained classifiers varied greatly
@@
 `stressed', the individual classifiers improved upon the baseline for
-S4, S6, S8, S11, S12, and S14 with a maximum improvement of \rev{152.0\%} in
-precision and \rev{111.1}\% in recall for S12, while they did worse for S1,
+S4, S6, S8, S11, S12, and S14 with a maximum improvement of 152.0\% in
+precision and 111.1\% in recall for S12, while they did worse for S1,
 S3, S5, S7, S9, S10, and S13, and in the worst cases did not correctly
@@
 We found that all methods produced similar results with respect to accuracy, precision, and recall for the individual models.
-Ultimately, we \rev{elected not to utilize any feature selection in order to simplify our approach, as we found there to be minimal differences in performance between the techniques, and the random forest algorithm is capable of (and robust for) handling datasets with many features.}
+Ultimately, we elected not to utilize any feature selection in order to simplify our approach, as we found there to be minimal differences in performance between the techniques, and the random forest algorithm is capable of (and robust for) handling datasets with many features.
 
@@
 Table~\ref{tab:featureImportance} shows the averaged Gini importance for the feature categories used for predicting stress.
-\rev{Heart rate variability proved to be an important measure for all of the aspects of interest, ranking as the most important feature category for both stress and awakeness, and the second most important one for focus.
+Heart rate variability proved to be an important measure for all of the aspects of interest, ranking as the most important feature category for both stress and awakeness, and the second most important one for focus.
 This is not surprising, as heart rate variability and skin temperature have been shown in several previous studies to be possible indicators for stress levels~\cite{dishman2000stress,mcduff16,kataoka00}.
@@
 Besides these mentioned feature categories, there was great variation in which measures were important to which of the three aspects.
-This shows that there is a clear benefit to having multiple biometric streams available for predicting stress, focus and awakeness.}\\[-0.1cm]
+This shows that there is a clear benefit to having multiple biometric streams available for predicting stress, focus and awakeness.\\[-0.1cm]
 
@@
 
-The bottom \rev{3 rows} of Table \ref{tab:accuracy} present the averaged performance results for this approach in terms of accuracy, precision, and recall\rev{, as well as the results of baseline stratified random and majority classifiers following the same leave-one-group-out cross-validation procedure.} Although the averaged precision and recall are comparable or better than those of the averaged individual results, this was at the cost of a large decrease in overall accuracy.
+The bottom 3 rows of Table \ref{tab:accuracy} present the averaged performance results for this approach in terms of accuracy, precision, and recall, as well as the results of baseline stratified random and majority classifiers following the same leave-one-group-out cross-validation procedure.
+Although the averaged precision and recall are comparable or better than those of the averaged individual results, this was at the cost of a large decrease in overall accuracy.
 Upon closer investigation into the performance of the general model when testing on each participant, we found that individually trained models for each participant performed much better than a general model trained over all participants.

### 2020-10-01 IJHCSSubmission/bp_studyMethod.tex
removing the past revision blue marks and starting with new ones for this coming revision

@@
 \section{Field Study}
-We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress \rev{and awakeness} over time and the feasibility of predicting stress, focus, and awakeness \rev{using} biometric signals.
+We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress and awakeness over time and the feasibility of predicting stress, focus, and awakeness using biometric signals.
 
@@
 We performed a study over the course of eight weeks.
-This study includes the collection of the participant's \rev{biometric data and self-reporting surveys during their work hours}.
-\rev{We also collected computer interaction data, but given privacy concerns of some participants chose not to use this data (see Section~\ref{sec:computer_interaction_data})}.
-We informed participants about the study purpose and procedure, handed out biometric sensors, introduced and explained the self-reporting to the participants, \rev{handed out consent forms and ensured informed consent from each participant for the study.}
+This study includes the collection of the participant's biometric data and self-reporting surveys during their work hours.
+We also collected computer interaction data, but given privacy concerns of some participants chose not to use this data (see Section~\ref{sec:computer_interaction_data}).
+We informed participants about the study purpose and procedure, handed out biometric sensors, introduced and explained the self-reporting to the participants, handed out consent forms and ensured informed consent from each participant for the study.
 
-After the initial setup, participants were asked to fill out three surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}) \rev{and wear the biometric sensor during their work hours}.
+After the initial setup, participants were asked to fill out three surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}) and wear the biometric sensor during their work hours.
 At the end of the eight weeks, we collected the biometric sensors and performed a short follow-up interview on the study and the participants' experiences.
@@
 The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/articles/213613165} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements for shorter periods of time or capturing a smaller number of measurements.
-\rev{A comparative study has  shown that the Everion can be used as a valid proxy for HRV metrics for knowledge workers~\cite{barrios2019a}.}
+A comparative study has  shown that the Everion can be used as a valid proxy for HRV metrics for knowledge workers~\cite{barrios2019a}.
 
@@
 Data collected by the Everion are uploaded to a server, from which we downloaded the data for use in our study.
-\rev{We chose these biometric measurements based on previous research that has indicated their potential (see references in Table~\ref{signals}) and the availability and feasibility of collecting these measurements with low invasiveness over a long duration.}
+We chose these biometric measurements based on previous research that has indicated their potential (see references in Table~\ref{signals}) and the availability and feasibility of collecting these measurements with low invasiveness over a long duration.
 
@@
 \caption{Biometric measurements captured by the Everion, organized by category and with references to previous works using similar data.
-\rev{\textit{(RMSSD denotes the root mean square of successive heartbeat interval differences)}}}
+\textit{(RMSSD denotes the root mean square of successive heartbeat interval differences)}}
 \label{signals}
@@
 \subsubsection{Computer Interaction Data}\label{sec:computer_interaction_data}
-To gain a better understanding of our participants day-to-day work activities, we \rev{asked participants to install} an open source computer interaction monitor~\cite{Meyer17}.
+To gain a better understanding of our participants day-to-day work activities, we asked participants to install an open source computer interaction monitor~\cite{Meyer17}.
 The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity.
@@
 
-\rev{We did not ask our participants about their focus levels in the end of the day survey as focus
-is more of an in-the-moment aspect than stress and awakeness.}
+We did not ask our participants about their focus levels in the end of the day survey as focus
+is more of an in-the-moment aspect than stress and awakeness.
 

### 2020-10-01 IJHCSSubmission/bp_threats.tex
removing the past revision blue marks and starting with new ones for this coming revision

@@
 This study tries to find correlations between
-biometric features and \rev{the human aspects of} stress, focus, and awakeness.
+biometric features and the human aspects of stress, focus, and awakeness.
 Nonetheless, biometric signals are influenced by far more
@@
 
-\rev{Due to the small amount of data available to us for the purposes of this study, there is a risk
+Due to the small amount of data available to us for the purposes of this study, there is a risk
 that our models may be overfitting to some degree.
 Such overfitting indicates that the results presented in our work may be less than optimal, however any overfitting applies strictly to the training data and given that there is no overlap between the data used for training and testing, we do not believe this invalidates our results.
-We leave further optimization of our approach to future researchers.}
+We leave further optimization of our approach to future researchers.
 
@@
 biometrics that historically have shown correlation with 
-\rev{the studied human aspects of stress, focus and awakeness}.
-In an effort to maintain this research applicable to real-world environments, we picked \rev{the already existing Everion device}, even when, as a trade-off, we could not capture more descriptive and more intrusive signals such as SDNN, SCL, SCR, eye tracking, or brain activity.
+the studied human aspects of stress, focus and awakeness.
+In an effort to maintain this research applicable to real-world environments, we picked the already existing Everion device, even when, as a trade-off, we could not capture more descriptive and more intrusive signals such as SDNN, SCL, SCR, eye tracking, or brain activity.
 
-\rev{A future more thorough statistical analysis of the relationships between
+A future more thorough statistical analysis of the relationships between
 the aspects of interest, such as stress, and the physiological
 data may further provide deeper insights into the data and
-how it might be used in prediction.}
+how it might be used in prediction.

### 2020-10-01 IJHCSSubmission/elsarticle-template.tex
removing the past revision blue marks and starting with new ones for this coming revision

--- 2020-08-07 IJHCSSubmission/elsarticle-template.tex
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{\rev{Observing and Predicting Knowledge Worker Stress, Focus and Awakeness in the Wild}}
+\title{Observing and Predicting Knowledge Worker Stress, Focus and Awakeness in the Wild}
 
@@
 %% used to denote shared contribution to the research.
-\author{Authors will be visible after double blind review}
+%\author{Authors will be visible after double blind review}
 %\vspace{110cm}
@@
 
-\begin{comment}
+\author{Mauricio Soto\corref{ma}}
+\ead{mauricio.soto-gonzalez1@hitachi-powergrids.com}
+\address{Hitachi ABB Power Grids, Raleigh, North Carolina}
 
-\author{Mauricio Soto}
-\affiliation{%
-\institution{ABB Future Labs}
-\city{Raleigh}
-\state{North Carolina}
-}
-\email{mauricio.soto@us.abb.com}
-
-\author{Chris Satterfield}
-\affiliation{%
-\institution{The University of British Columbia}
-\city{Vancouver}
-\state{British Columbia}
-}
-\email{cds00@cs.ubc.ca}
+\author{Chris Satterfield\corref{ma}}
+\ead{cds00@cs.ubc.ca}
+\address{The University of British Columbia, Vancouver, British Columbia}
 
 \author{Thomas Fritz}
-\affiliation{%
-\institution{University of Zurich}
-\city{Zurich}
-\state{Switzerland}
-}
-\email{fritz@ifi.uzh.ch}
+\ead{fritz@ifi.uzh.ch}
+\address{University of Zurich, Zurich, Switzerland}
 
 \author{Gail C. Murphy}
-\affiliation{%
-\institution{The University of British Columbia}
-\city{Vancouver}
-\state{British Columbia}
-}
-\email{murphy@cs.ubc.ca}
+\ead{murphy@cs.ubc.ca}
+\address{The University of British Columbia, Vancouver, British Columbia }
 
 \author{David Shepherd}
-\affiliation{%
-\institution{Virginia Commonwealth University}
-\city{Richmond}
-\state{Virginia}
-}
-\email{shepherdd@vcu.edu}
+\ead{shepherdd@vcu.edu}
+\address{Virginia Commonwealth University, Richmond, Virginia}
 
 \author{Nicholas Kraft}
-\affiliation{%
-\institution{UserVoice}
-\city{Raleigh}
-\state{North Carolina}
-}
-\email{kraft.nicholas.a@gmail.com} \\}
+\ead{kraft.nicholas.a@gmail.com}
+\address{UserVoice, Raleigh, North Carolina}
 
-\end{comment}
+\cortext[ma]{Main Author}
 
@@
 knowledge workers over an eight-week period in which we investigated, using
-experience sampling, how the workers experience stress, \rev{focus and awakeness} over time.
-During this field study, we also collected biometric data \rev{including heart- and skin-related measures}, which we then used to investigate if it is possible to predict stress, focus and awakeness, in the moment.
-We observed and report on various trends in knowledge worker stress levels
-over several weeks and show that machine learning models can be built
-\rev{from the data of a single minimally invasive sensor}
+experience sampling, how the workers experience stress \rev{and} awakeness over time.
+During this field study, we also collected biometric data including heart- and skin-related measures, which we then used to investigate if it is possible to predict stress, focus and awakeness, in the moment.
+We observed and report on various trends in knowledge worker stress and awakeness levels
+over several weeks, finding that people tend to have certain ``baseline'' levels for these aspects and stressful days or days with low awakeness tend to cluster.
+We further show that machine learning models can be built
+from the data of a single minimally invasive sensor
 to
-predict stress, focus and awakeness \rev{accurately, with the abstract concept of focus
-the hardest to predict.}
+predict stress, focus and awakeness.
+Overall, we found that our models were capable of large improvements in precision and recall in comparison to a random classifier for stress (25.9\% increase over random for precision, 4.2\% for recall) and awakeness (52.4\% increase in precision, 40.8\% in recall).
+The abstract concept of focus proved to be
+the hardest to predict (26.0\% increase in precision, 27.8\% decrease in recall).
 \end{abstract}
@@
 \begin{keyword}
-Biometrics, \tf{Psycho-Physiological Data}, Stress, Awakeness, Focus, Computer Interaction, Ubiquitous Computing, Empirical Study, User Centered Design
+Biometrics, Psycho-Physiological Data, Stress, Awakeness, Focus, Computer Interaction, Ubiquitous Computing, Empirical Study, User Centered Design
 \end{keyword}

### 2020-10-01 IJHCSSubmission/elsarticle-template.tex
reviewed Intro

@@
 
-\cortext[ma]{Main Author}
+\cortext[ma]{Main Authors}
 
@@
 We observed and report on various trends in knowledge worker stress and awakeness levels
-over several weeks, finding that people tend to have certain ``baseline'' levels for these aspects and stressful days or days with low awakeness tend to cluster.
+over several weeks, finding that people tend to have certain ``baseline'' levels for these aspects\rev{.
+Moreover, we found that days with high levels of stress tend to cluster, similarly as the days with low awakeness}.
 We further show that machine learning models can be built
-from the data of a single minimally invasive sensor
-to
-predict stress, focus and awakeness.
+from the data of a single minimally invasive \rev{device} to
+predict stress, focus, and awakeness.
 Overall, we found that our models were capable of large improvements in precision and recall in comparison to a random classifier for stress (25.9\% increase over random for precision, 4.2\% for recall) and awakeness (52.4\% increase in precision, 40.8\% in recall).
-The abstract concept of focus proved to be
-the hardest to predict (26.0\% increase in precision, 27.8\% decrease in recall).
+The abstract concept of focus proved to be the hardest to predict (26.0\% increase in precision, 27.8\% decrease in recall).
 \end{abstract}

### 2020-10-08 IJHCSSubmission/bp_conclusion.tex
Addressing the reviewers last minor comments

@@
 paper, we presented the results of a study with 14 professional
-knowledge workers in their workplace over an eight week period to
+knowledge workers in their workplace over an \rev{eight-week} period to
 better understand how workers experience these human aspects over time and to

### 2020-10-08 IJHCSSubmission/bp_discussion.tex
Addressing the reviewers last minor comments

@@
 ability to sense and predict stress in the moment could help companies
-to prevent or de-escalate potentially dangerous situations in the
-workplace, e.g. confrontation between co-workers. Building an 
+prevent or de-escalate potentially dangerous situations in the
+workplace \rev{(e.g., confrontation between co-workers)}. Building an 
 understanding of stress and focus over time and in the moment
@@
 any office environment, and thus could impact a large percentage of
-modern workers. As in modern cars, temperature and lighting could be
+modern workers.
+
+As in modern cars, temperature and lighting could be
 regulated on a per-person basis, which would allow the environment to
@@
 due to our focus on stress, we offered participants an opportunity to be part of Tai Chi classes.
-The two primary reasons for this intervention was to keep participants motivated to continue the self-reports over the long study period, and to offer a technique that might help reduce stress.
+The two primary reasons for this intervention \rev{were} to keep participants motivated to continue the self-reports over the long study period, and to offer a technique that might help reduce stress.
 After consulting with other researchers, we decided to offer Tai Chi in the last month of the study, leaving the first month of the study unaltered.
@@
 
-While this was not the focus of our study, we performed a secondary analysis to examine whether the Tai Chi classes had any effect on the participants' stress levels in the work days directly following the Tai Chi session.
+While this was not the focus of our study, we performed a secondary analysis to examine whether the Tai Chi classes had any effect on the participants' stress levels in the workdays directly following the Tai Chi session.
 For this, we build a linear mixed model with the self-reported daily stress level as dependent variables and the participants as random effects.
@@
 Alternative techniques such as SMOTE~\cite{chawla2002smote} exist and can perform better than those we employed, however they are impractical considering the limited amount of data we have to work with.
+
 Oversampling the dataset as we did may also lead to an increased risk of model overfitting.
 However, we believe the benefits of rebalancing the dataset outweigh this possible downside.
-
-Especially in light of the imbalance
+In light of the imbalance
 in the data, the results we achieved with our models are

### 2020-10-08 IJHCSSubmission/bp_relatedWork.tex
Addressing the reviewers last minor comments

@@
 Hernandez et al.~\cite{Hernandez14} investigated the use of a pressure-sensitive keyboard and a capacitive mouse as non-intrusive means for measuring computer users' stress levels.
-They found participants exhibited significantly increased typing pressure and mouse contact when in stressful conditions.
+They found participants\rev{'} exhibited significantly increased typing pressure and mouse contact when in stressful conditions.
 McDuff et al.~\cite{mcduff16} experimented with a camera to measure photoplethysmographic signals indicative of cognitive stress.
@@
 Our study stands out from these works by nature of its length and focus on knowledge worker stress in everyday office life, using unobtrusive measures.
-To the best of our knowledge there is no longitudinal study which attempts to explain and predict knowledge worker stress that comes close to the length of our own study.
+To the best of our knowledge there is no longitudinal study \rev{that} attempts to explain and predict knowledge worker stress that comes close to the length of our own study.
 The eight week duration gives us authority to speak on the nature of day-to-day fluctuations in knowledge worker stress levels.

### 2020-10-08 IJHCSSubmission/bp_results_1.tex
Addressing the reviewers last minor comments

@@
 
-The baseline stress level varied significantly between participants. 7
+The baseline stress level varied significantly between participants. \rev{Seven}
 participants (54\%) reported feeling average stress levels most frequently
-(rating 3 on our scale), while 5 (38\%) reported feeling little stress
-(rating 2) and 1 (8\%) reported feeling no stress at all (rating 1).
+(rating 3 on our scale), while \rev{five} (38\%) reported feeling little stress
+(rating 2) and \rev{one} (8\%) reported feeling no stress at all (rating 1).
 Figure \ref{fig:dailyStress} illustrates these points using  data from two participants, showing the tendency to report and return to baseline stress levels, as well
@@
 baselines, we consider a stressful day to be one that represents a
-deviation of 1 or more stress levels above the participant's
+deviation of \rev{one} or more stress levels above the participant's
 baseline. Of the 93 stressful days we observed in total, we found that
@@
 participants were far more likely to report a stress level that was
-within 1 point of their baseline, than to report a stress level 2 or
+within \rev{one} point of their baseline, than to report a stress level 2 or
 more points away. These extreme deviations represented only 15\% of
-all reported values, which differed from the participants baseline. As
+all reported values, which differed from the \rev{participant's} baseline. As
 well, the majority (78\%) of these deviations came from just two
@@
 \subsection{Awakeness}
-We applied the same analyses described above to the self reported awakeness levels of our participants.
+We applied the same analyses described above to the \rev{self-reported} awakeness levels of our participants.
 Compared with the reported stress levels, we observed the same trend of participants reporting one baseline far more commonly than any other.
 We did not find there to be a significant correlation between reported stress and awakeness levels.
-Overall, participants awakeness levels fluctuated significantly less than their stress levels (73.1\% of reports were at the baseline level, compared to 65.1\% for stress, p < 0.05).
+Overall, \rev{participant's} awakeness levels fluctuated significantly less than their stress levels (73.1\% of reports were at the baseline level, compared to 65.1\% for stress, p < 0.05).
 Participants were unlikely to experience days with heightened (above baseline) awakeness.
-Such days made up only 6.0\% of the total observed work days across all participant.
-Most of these days came from one participant, S2, who reported 15 heightened awakeness days compared to the next highest, S13 with 4.
-Large deviations (>1 point deviation) from each participants baseline awakeness levels were extremely uncommon, accounting for only 9 (2.1\%) of our total observations.
+Such days made up only 6.0\% of the total observed \rev{workdays} across all participant.
+Most of these days came from one participant, S2, who reported 15 heightened awakeness days compared to the next highest, S13 with four.
+Large deviations (>1 point deviation) from each participants baseline awakeness levels were extremely uncommon, accounting for only \rev{nine} (2.1\%) of our total observations.
 Similarly to what we observed with high stress days, low awakeness days frequently came in clusters of two or three days in a row.
-Given that the immediately preceding day was a low awakeness day, a given day was 228.2\% more likely to be a low awakeness day than when considering any day at random (p < 0.0001).
+Given that the immediately preceding day was a low awakeness day\rev{;} a given day was 228.2\% more likely to be a low awakeness day than when considering any day at random (p < 0.0001).
 
@@
     \centering
-\begin{tabular}{|c|c c|c c|}
+\rev{\begin{tabularx}{\textwidth}{|c|c c c|c c c|}
     \hline
         &
-        \multicolumn{2}{c|}{Stress} & 
-        \multicolumn{2}{c|}{Awakeness} \\
+        \multicolumn{3}{c|}{Stress} & 
+        \multicolumn{3}{c|}{Awakeness} \\
         \hline
-        \textbf{Explanatory Variable} & $\mathbf{R^{2}}$ & \textbf{p-value} &  $\mathbf{R^{2}}$ & \textbf{p-value}\\
+        \textbf{Variable} & \centering
+        \textbf{\rev{F.E.E.}}& \textbf{p-value} & \textbf{\rev{C.I. (95\%)}} & \centering \textbf{\rev{F.E.E.}}  & \textbf{p-value} & \textbf{\rev{C.I. (95\%)}}\\
         \hline
-        Day of Week - Monday & -0.033 & 0.745 & -0.058 & 0.497\\
-        Day of Week - Tuesday & 0.028 & 0.773 & 0.036 & 0.662 \\
-        Day of Week - Wednesday & 0.163 & 0.100 & -0.018 & 0.830 \\
-        Day of Week - Thursday & 0.022 & 0.821 & 0.076 & 0.927 \\
-        Day of Week - Friday & 0.082 & 0.473 & -0.178 & 0.025 \\
-        Proximity to Month End & 0.044 & 0.687 & -0.082 & 0.374 \\
+        Monday & \centering -0.033 & 0.745 & \rev{$(-0.234, 0.167)$} & \centering -0.058 & 0.497 & \rev{$(-0.227, 0.110)$}\\
+        Tuesday & \centering 0.028 & 0.773 & \rev{$(-0.164, 0.221)$} & \centering 0.036 & 0.662 & \rev{$(-0.126, 0.198)$}\\
+        Wednesday & \centering 0.163 & 0.100 & \rev{$(-0.031, 0.357)$} & \centering -0.018 & 0.830 & \rev{$(-0.181, 0.145)$}\\
+        Thursday & \centering 0.022 & 0.821 & \rev{$(-0.169, 0.213)$} & \centering 0.076 & 0.927 & \rev{$(-0.085, 0.238)$}\\
+        Friday & \centering 0.082 & 0.473 & \rev{$(-0.142, 0.306)$} & \centering -0.178 & 0.025 & \rev{$(-0.334, -0.022)$}\\
+        Month End & \centering 0.044 & 0.687 & \rev{$(-0.171, 0.259)$} & \centering -0.082 & 0.374 & \rev{$(-0.263, 0.099)$}\\
         \hline
-    \end{tabular}
-    \caption{Correlation coefficient and associated p-values for the explanatory variables we examined in our linear mixed model analysis.}
-    \label{tab:mixedModel}
+    \end{tabularx}
+    \caption{\rev{Fixed effects estimates (F.E.E.), 95\% confidence intervals and associated p-values for the explanatory variables (day of week and proximity to month end) that we examined in our linear mixed model analysis.}}    \label{tab:mixedModel}}
 \end{table}
@@
 Ultimately, the analysis showed that none of the variables that we examined had a significant explanatory power with respect to our participants perceived stress levels.
-For awakeness, we found  that there was a small ($R^2$: -0.178) yet significant decrease in awakeness levels on Fridays in particular.
+For awakeness, we found  that there was a small (\rev{fixed effects estimate}: -0.178) yet significant decrease in awakeness levels on Fridays in particular.
 Table \ref{tab:mixedModel} shows some of the detailed results of these analyses.
@@
   \caption{The distribution of the responses of each participant to the three questions asked during the day are shown.
-Each bar in the histograms represent one of the 5 values on the 5-point Likert scale we asked participants to respond with, where the far left side of the histograms are 1/Not at all, and the far right sides are 5/Extremely}
+Each bar in the histograms represent one of the \ref{five} values on the 5-point Likert scale we asked participants to respond with, where the far left side of the histograms are 1/Not at all, and the far right sides are 5/Extremely}
    \label{responseDistribution}
@@
 Given
-this distribution and based on our earlier observation that the participants tended to adhere to a baseline reporting level for stress and awakeness, we elected to simplify the problem from 5 classes to 2.
+this distribution and based on our earlier observation that the participants tended to adhere to a baseline reporting level for stress and awakeness, we elected to simplify the problem from \rev{five} classes to \rev{two}.
 This transformation enables us to more easily represent patterns in the data, such as when a participant fluctuates from a normal to high stress level.
-To perform this transformation, we began for
-each participant by calculating the median response value for each question.
+To perform this transformation, \rev{we began by calculating the median response value for each participant and each question}.
 We classified each response below the median as 0 (`negative') and each response above the median as 1 (`positive').
@@
 
-There are many different algorithms that can be used to build a classifier. To select an
+\rev{Many different algorithms can be used to build a classifier.} To select an
 algorithm, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11},  evaluating each one by using leave-one-out cross validation.
@@
 
-Beside our results, for the purposes of comparison we also present two commonly used baselines in this table - a majority classifier, which always predicts the larger of the two classes considered, and a stratified random classifier which randomly chooses between the two classes, but with a proportional bias towards the larger class.
+\rev{Besides} our results, for the purposes of comparison\rev{,} we also present two commonly used baselines in this table - a majority classifier, which always predicts the larger of the two classes considered, and a stratified random classifier which randomly chooses between the two classes, but with a proportional bias towards the larger class.
 
 For some users (i.e., S1.
-as seen in Table \ref{responseDistribution}), the imbalance in their data was so extreme that even after adjusting by oversampling it was impossible to create a reasonable classifier.
-These scenarios are difficult to predict as any classifier will not have enough variance in its training data for the 'stressed' situation to adequately distinguish it from the non-stressed case.
+as seen in Table \ref{responseDistribution}), the imbalance in their data was so extreme that even after adjusting by oversampling\rev{, we were not able} to create a reasonable classifier.
+These scenarios are difficult to predict\rev{,} as any classifier will not have enough variance in its training data for the 'stressed' situation to adequately distinguish it from the non-stressed case.
 
@@
 recall.
-We present a comparison between the averaged results of our individual classifiers and those of the baseline stratified random classifer in the ``Improvement Over Random'' row of Table \ref{tab:accuracy}.
+We present a comparison between the averaged results of our individual classifiers and those of the baseline stratified random \rev{classifier} in the ``Improvement Over Random'' row of Table \ref{tab:accuracy}.
 This is calculated as the difference between the overall average and the baseline results, divided by the baseline results (for example, $\frac{Acc_{Overall} - Acc_{Random}}{Acc_{Random}}$).
-We do not compare our results with the majority classifier directly as this classifier achieved precision and recall scores of zero when predicting stress, lack of focus, and lack of awakeness, making a meaningful comparison impossible.
+We do not compare our results with the majority classifier directly as this classifier achieved precision and recall scores of zero when predicting stress, lack of focus, and lack of awakeness, making a meaningful comparison \rev{unfeasible}.
 The improvement percentages demonstrate that the predictions made by our classifiers are much better than random after correcting for the imbalance in our dataset.
@@
 
-The bottom 3 rows of Table \ref{tab:accuracy} present the averaged performance results for this approach in terms of accuracy, precision, and recall, as well as the results of baseline stratified random and majority classifiers following the same leave-one-group-out cross-validation procedure.
+The bottom \rev{three} rows of Table \ref{tab:accuracy} present the averaged performance results for this approach in terms of accuracy, precision, and recall, as well as the results of baseline stratified random and majority classifiers following the same leave-one-group-out cross-validation procedure.
 Although the averaged precision and recall are comparable or better than those of the averaged individual results, this was at the cost of a large decrease in overall accuracy.

### 2020-10-08 IJHCSSubmission/bp_studyMethod.tex
Addressing the reviewers last minor comments

@@
 Keeping participants engaged in an eight-week study can be
-difficult. In the second month of the study, largely to incent participants to
+difficult. In the second month of the study, largely to \rev{incentivize} participants to
 continue, we offered participants two one-hour Tai Chi classes per
@@
 \label{sec:DataCol}
-We collected two datasets  from each study participant.
+We collected \rev{three} datasets  from each participant \rev{as described below:}
 
@@
 Each measurement is collected once per second, and each recorded observation has an associated timestamp and quality rating.
-Data collected by the Everion are uploaded to a server, from which we downloaded the data for use in our study.
-We chose these biometric measurements based on previous research that has indicated their potential (see references in Table~\ref{signals}) and the availability and feasibility of collecting these measurements with low invasiveness over a long duration.
+Data collected by the Everion \rev{is} uploaded to a server, from which we downloaded the data for use in our study.
+We chose these biometric measurements based on previous research \rev{which indicates} their potential (see references in Table~\ref{signals}) and the availability and feasibility of collecting these measurements with low invasiveness over a long duration.
 
@@
 text message a survey request to each participant two times per 
-work day. Pilot participants preferred the usage of text message, in part, due 
+workday. Pilot participants preferred the usage of text message, in part, due 
 to them being accessible and noticeable anywhere in the office.
@@
 %in time to self-report [2], or facilitate the self-log by using wearables.
-The same survey was sent each time:
+The same survey was sent each time \rev{consisting of the following three questions}:
 \begin{enumerate}
@@
 We used the phrase ``right now'' to capture each aspect in the moment (so as to permit later prediction of each aspect based on biometric data).
-The wordings of the questions are based on a previous survey of individuals in an organizational context~\cite{Gloor_etal:2010}.
+The \rev{wording of the questions is} based on a previous survey of individuals in an organizational context~\cite{Gloor_etal:2010}.
 The use of awakeness (rather than sleepiness) in Question 1 is inspired by previous work~\cite{Wilhelm_Schoebi:2007} and to some extent also captures the ``arousal'' aspect of the affective space~\cite{Russell:1980}.
@@
 %\todo{we are not using the last survey for anything in this paper, so I'd leave it out}
-One last survey was sent at the end of the day, at 4:25pm which asked the 
+One last survey was sent at the end of the day at 4:25pm\rev{,} which asked the 
 four different questions
@@
 \item How productive have you been today?
-\item How do you feel about your work day?
+\item How do you feel about your workday?
 \end{enumerate}
@@
 
-We did not ask our participants about their focus levels in the end of the day survey as focus
+We did not ask our participants about their focus levels in the end of the day survey\rev{,} as focus
 is more of an in-the-moment aspect than stress and awakeness.

### 2020-10-08 IJHCSSubmission/bp_threats.tex
Addressing the reviewers last minor comments

@@
 To mitigate this risk, we collected participants
-from a wide variety of different departments
+from a wide variety of departments
 with different age ranges, genders, work experience, and 
-working in different positions.
+working in different positions \rev{therefore providing evidence
+that our approach's performance is comprised of and can generalize to a wide range of 
+knowledge workers}.
 
-Another threat is that our results may not generalize
+\rev{Secondly,} our results may not generalize
 to a different office environment. We conducted

### 2020-10-08 IJHCSSubmission/elsarticle-template.tex
Addressing the reviewers last minor comments

@@
 %% for a journal layout:
-   \documentclass[final,1p,times,plainnat]{elsarticle}
+   \documentclass[final, 1p,times]{elsarticle}
 %% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
@@
 fragmented, disruptions are constant, tasks are complex, and work hours can be 
-long. These challenges can impact knowledge workers' stress, focus and awakeness, and in turn their interaction with the digital environment, the quality of work 
+long. These challenges can \rev{affect} knowledge workers' stress, focus and awakeness, and in turn their interaction with the digital environment, the quality of work 
 performed and their productivity in general. We report on a field study with 14

### 2020-10-12 IJHCSSubmission/bp_results_1.tex
minor last revisions

@@
     \end{tabularx}
-    \caption{\rev{Fixed effects estimates (F.E.E.), 95\% confidence intervals and associated p-values for the explanatory variables (day of week and proximity to month end) that we examined in our linear mixed model analysis.}}    \label{tab:mixedModel}}
+    \caption{\rev{Fixed effects estimates (F.E.E.), 95\% confidence intervals (C.I.) and associated p-values for the explanatory variables (day of week and proximity to month end) that we examined in our linear mixed model analysis.}}    \label{tab:mixedModel}}
 \end{table}

### 2020-10-20 IJHCSSubmission/bp_results_1.tex
fixed typo

@@
   \caption{The distribution of the responses of each participant to the three questions asked during the day are shown.
-Each bar in the histograms represent one of the \ref{five} values on the 5-point Likert scale we asked participants to respond with, where the far left side of the histograms are 1/Not at all, and the far right sides are 5/Extremely}
+Each bar in the histograms represent one of the five values on the 5-point Likert scale we asked participants to respond with, where the far left side of the histograms are 1/Not at all, and the far right sides are 5/Extremely}
    \label{responseDistribution}
